---
title: "Mock Communities ASV analyses"
author: "Gabriel A. Mendes", "Heron OH"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Mock Communities ASV versus MOTUs project - ASV analyses

## 1. Setup & Configuration

### Packages

-   **Load libraries.** Import all necessary packages for data manipulation, demultiplex, ASVs inference, taxonomy assigment and data visualization.

    ```{r}
    library(tidyverse)
    library(phyloseq)
    library(Biostrings)
    library(ShortRead)
    library(dada2)
    library(DECIPHER)
    library(future)
    library(viridis)
    library(ggh4x)
    library(furrr)
    library(writexl)
    library(tictoc)
    ```

### Directory Management

-   **Create directory structure.** Define project paths and create the directory structure required by the pipeline.

    ```{r}
    # Project folder path  
    prjct_path <- "/home/gabriel/projetos/mock_pipelines"

    # Derived paths
    {
      env_path <- file.path(prjct_path, "environments")
      data_path <- file.path(prjct_path, "data")
      pipe_libs <- file.path(data_path, "reads")
      raw_libs <- file.path(pipe_libs, "raw")
      paired_libs <- file.path(pipe_libs, "paired")
      cutadapt_libs <- file.path(pipe_libs, "cutadapt")
      qual_path <- file.path(prjct_path, "quality")
      results_path <- file.path(prjct_path, "results")
      blast_path <- file.path(results_path, "blast")
      swarm_path <- file.path(results_path, "swarm")
      dmx_path <- file.path(pipe_libs, "dmx")
      dmx_all_path <- file.path(dmx_path, "all_dmx")
      dmx_combined_path <- file.path(dmx_path, "all_combined")
    }

    # Paths to be created
    paths <- c(prjct_path, env_path, data_path, pipe_libs, raw_libs, 
             paired_libs, cutadapt_libs, qual_path, results_path,
             blast_path, swarm_path, dmx_path, dmx_all_path, 
             dmx_combined_path)

    for (dir in paths) {
      if (dir.exists(dir)) {
    print(paste("The folder", dir, "already exists!"))
      } else {
    print(paste("Creating", dir, "!" ))
    dir.create(dir) 
      }
    }

    # Files in the project folder
    list.files(prjct_path)
    ```

### Housekeeping

-   **Set Cutadapt path**

    ```{r}
    # Path to cutadapt executable
    cutadapt <- "/usr/local/bin/cutadapt"
    ```

-   **Define custom functions.**

    ```{r}
    # Convert IUPAC codes to regex. Enables searching degenerate primers directly on reads.
    iupac2seq <- function(dna_seq) {
      iupac_code <- c("A" = "A", "C" = "C", "G" = "G", "T" = "T",
                      "Y" = "[CT]", "R" = "[AG]", "H" = "[ACT]", "M" = "[AC]",
                      "K" = "[GT]", "S" = "[GC]", "W" = "[AT]", "B" = "[CGT]",
                      "D" = "[AGT]", "V" = "[ACG]", "N" = "[ACGT]")
      stringr::str_replace_all(dna_seq, iupac_code)
    }

    # Counts how many reads in a FASTQ file contain a given primer (allowing up to 1 mismatch)
    primerHits <- function(primer, file_name) {
      nhits <- Biostrings::vcountPattern(primer,
                                         ShortRead::sread(ShortRead::readFastq(file_name)),
                                         fixed        = FALSE,
                                         max.mismatch = 1)
      sum(nhits > 0)
      }

    # Applies primerHits to a set of primers for one FASTQ file and returns a table of counts per primer
    multi_primerHits <- function(Read_file, primers) {
      primer_counts <- purrr::map_df(primers,
                                     .f = primerHits,
                                     file_name = Read_file)
      primer_counts %>%
        dplyr::mutate(`Read file` = Read_file)
    }

    # Generates all 4 orientations (FWD/REV/COMP/REVCOMP) for a single primer
        allOrients <- function(primers_list) {
      require(Biostrings)
      
      dna <- Biostrings::DNAString(primers_list$`Primer seq`)
      
      orients <- 
        c(Forward= dna,
          Complement = Biostrings::complement(dna),
          Reverse    = Biostrings::reverse(dna),
          RevComp    = Biostrings::reverseComplement(dna))
      
      primer_tbl <- sapply(orients, toString)
      
      primer_tbl <- 
        tibble::tibble(Sequence = primer_tbl,
                       `Primer orientation` = names(primer_tbl)) %>%
        dplyr::mutate(Primer = primers_list$`Primer name`) %>%
        tidyr::unite(col = "Orientation name",
                     Primer,
                     `Primer orientation`,
                     remove = FALSE) %>%
        dplyr::mutate(`Primer pair` = stringr::str_remove_all(Primer, "__.*.$"),
                      `Parsed sequence` = iupac2seq(Sequence))
      
      base::names(primer_tbl$Sequence) <- primer_tbl$`Orientation name`
      
      primer_tbl
      }
    ```

-   **Load custom functions.** Custom functions for taxonomy retrievel.

    ```{r}
    # Demultiplex custom function
    source("/home/heron/prjcts/ecomol/R/dmx_metabar.R")
    ```

-   **Define Global Color Palettes.** Set consistent color schemes for plots.

    ```{r}
    # Color for each read situation 
    read_colors <- c("Raw reads (pairs)" = "#7A0403FF",
                     "Raw reads (R1)" = "#B11901FF",
                     "Raw reads (R2)" = "#B11901FF",
                     "Paired reads" = "#F25F14FF",
                     "Primer detected reads" = "#FE942AFF",
                     "Primer detected reads (R1)" = "#F7C13AFF",
                     "Primer detected reads (R2)" = "#F7C13AFF",
                     "Repaired reads" = "#DAE336FF",
                     "Denoised (R1)" = "#AFFA37FF",
                     "Denoised (R2)" = "#AFFA37FF",
                     "Merged" ="#76FE5BFF",
                     "Non-chimeric Merged" ="#34F395FF",
                     "Concatenated" = "#2DB6F1FF",
                     "Non-chimeric Concatenated" = "#458AFCFF",
                     "R1 + R2" = "#4145AAFF",
                     "Non-chimeric R1 + R2" = "#3A2C78FF")

    # Plot identified ASVs
    colors <- c("#330000", "darkred", "red", 
                "yellow", "green", "darkgreen")

    # BLAST Score Gradient (Low to High quality)
    blast_score_palette <- c("#330000", "darkred", "red", 
                             "yellow", "green", "darkgreen")

    # Metazoa Status (True/False)
    metazoa_palette <- c("TRUE" = "steelblue", "FALSE" = "firebrick") 

    # Shapes for Superkingdoms
    taxa_shapes <- c("Archaea" = 25, "Bacteria" = 24, 
                     "Eukaryota" = 21, "Viruses" = 23)
    ```

### Load Samples table

This is the most important input on the analysis, along with the raw data. The *primer_n_samples Table* holds the information of the samples file names, primers, controls, indexes, different projects. **All names under the Unique_file_name colunm must be unique.**

After filling in the table, save it as .csv an place it on the data folder, inside the project's main directory.

-   **Load primers, indexes and samples table.** Reads the configuration CSV containing sample metadata, primers, indexes, and project assignments.

    ```{r}
    primers_n_samples <- 
      readr::read_csv(file = file.path(prjct_path, "data", "mock_primer_n_samples.csv"))

    # Show if the primer_n_samples table was loaded successfully
    primers_n_samples %>% 
      View()
    ```

-   **Check if file names are really unique.** Validates that Unique_File_name and Sample columns contain no duplicates, stopping execution if found.

    ```{r}
    if (nrow(primers_n_samples) != length(unique(primers_n_samples$Unique_File_name))) {
      message("Your file names are not unique")
      message(paste0("The file names:"))
      message(paste0(primers_n_samples$Unique_File_name[duplicated(primers_n_samples$Unique_File_name)], collapse = "\n "))
      message(paste0("apear more than once"))
      } else if (nrow(primers_n_samples) != length(unique(primers_n_samples$Sample))) {
        message("Your Sample names are not unique")
        message(paste0("The sample names:"))
        cat(paste0(primers_n_samples$Sample[duplicated(primers_n_samples$Sample)], collapse = "\n "))
        message(paste0("apear more than once"))
        } else {
          message("All file names and sample names are unique!")
          }
    ```

-   **Defining all separate projects to analyze independently.** Extracts and lists all unique project names from the samples table.

    ```{r}
    all_projects <- BiocGenerics::unique(primers_n_samples$Project) %>%
      stringr::str_split(pattern = ";", simplify = F) %>%
      base::unlist() %>%
      BiocGenerics::unique()

    paste0("This analysis will by split into separate projects for:", 
           paste0(c("\t", all_projects), 
                  collapse = "\n\t")) %>% 
    cat() %>% 
    message()
    ```

-   **Identify files names radicals and levels.** Displays raw Unique_File_name values for manual inspection and establishes factor levels for consistent plotting order.

    ```{r}
    # if you need to print sample names and sort them yourself ----
    primers_n_samples$Unique_File_name %>% 
      paste0(collapse = '",\n"') %>% cat()

    # organize sample names in the order they will be printed on plots and results tables ----
    # sample_levels <- c(primers_n_samples$Unique_File_name) %>% unique()
    sample_levels <- c(primers_n_samples$Sample) %>% unique()

    # Show sample names
    sample_levels
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "1_setup_n_configuration-", Sys.Date()))
    ```

## 2. Raw Data Pre-processing

The second phase of the pipeline focuses on the pre-processing and diagnostics of the raw sequencing reads. Before proceeding to ASV inference, it is critical to assess read quality (via FastQC/MultiQC) and validate primer presence and orientation. These diagnostic analyses ensure data integrity and inform the parameters required for the subsequent filtering and cleaning steps.

### Raw data aquisition

-   **Get reads from Base Space.** This part is only required if your read files is on Basespace. If the read files are already downloaded and demultiplexed, place them on the *\$raw_libs* folder and proceed to the next section.

    ```{bash}
    # Navigate to raw files folder 
    cd  /home/gabriel/projetos/mock_pipelines/data/reads/raw;

    # Download run from BaseSpace ----
    # a - autenticate to basespace (must have a basespace account and shared projects/runs)
    bs auth;

    # 2b - list your projects data ----
    bs list datasets;

    # 2c - download multiplexed read files
    for id in $(cat ids.txt); \ do bs download dataset --id $id -o data/reads/raw --extension=fastq.gz; done
    ```

### Demultiplexing (Ignored. See Mock Pipelines Notion file)

-   **Run demultiplexing function.** Iterates over libraries in the configuration table (`primers_n_samples`) and applies the custom `dmx_metabar` function to split samples using the index information.

-   This stage its not needed because the samples are within different libraries.

    ```{r}
    # Identify all unique libraries
    all_Libs <- unique(primers_n_samples$Lib)

    # Loop over libraries and run demultiplexing
    for (LIB in all_Libs) {
      
     # Filter metadata for current library
    primers_n_samples_lib <- primers_n_samples %>%
     dplyr::filter(Lib %in% c(LIB))
      
    # Define raw R1/R2 file paths
    raw_files <- primers_n_samples_lib %>%
      dplyr::pull(`Undemultiplexed path`) %>%
      BiocGenerics::unique() %>%
      list.files(full.names = T) %>%
      Sys.glob() %>%
      normalizePath()
      
    # Set project parameters
    out_path <- primers_n_samples_lib %>%
      dplyr::pull(`Raw data path`) %>%
      BiocGenerics::unique()

    project  <-primers_n_samples_lib$Project %>%
      BiocGenerics::unique()

    # Run demultiplexing
    dmx_metabar(undmx_file_R1 = raw_files[1], #raw R1
                undmx_file_R2 = raw_files[2], #raw R2
                out_path = out_path, # pasta onde vai ser escrito todo output
                project = project, #nome do projeto, entre aspas tipo "eDNA_bromelias" sem acento nem espaço
                Unique_File_name = primers_n_samples_lib$Unique_File_name,  #coluna da tabela de configuração
                primers = primers_n_samples_lib$Primer,                     #coluna da tabela de configuração
                samples_names = primers_n_samples_lib$Sample,
                idxs_FWD_seqs = primers_n_samples_lib$`Index sequence FWD`, #coluna da tabela de configuração
                idxs_FWD_names = primers_n_samples_lib$`Index name FWD`,    #coluna da tabela de configuração
                idxs_REV_seqs = primers_n_samples_lib$`Index sequence REV`, #coluna da tabela de configuração
                idxs_REV_names = primers_n_samples_lib$`Index name REV`,    #coluna da tabela de configuração
                query_size = 12   #tamanho da sequencia a ser considerada do index + primer
    )
    }
    ```

### Quality checking

The raw sequencing reads are assessed for overall quality before any processing steps. This step runs FastQC and MultiQC on the demultiplexed files, generating per-project interactive reports. The resulting quality profiles — particularly per-position quality scores, GC content, and read length distributions — provide the baseline needed to interpret subsequent primer detection and filtering results.

-   **Check that demultiplexed files exist.** Confirms that the paths in `primers_n_samples` actually contain FASTQ files before running FastQC/MultiQC.

    ```{r}
    read_files <- primers_n_samples %>%
      dplyr::mutate(Read_file_pattern = file.path(`Raw data path`,
                                                  paste0(Unique_File_name, "*"))) %>%
      dplyr::pull(Read_file_pattern) %>%
      Sys.glob() %>%
      normalizePath() %>%
      unique()

    read_files

    if (length(read_files) > 0) {
      message(length(read_files), " read files found for QC.")
      } else {
        stop("No read files found for QC. Check `Raw data path` and file names `primers_n_samples`.")
        }
    ```

-   **Run FastQC and MultiQC by project.** Creates per‑project folders and runs quality control in batch using system calls.

    ```{r}
    for (PRJ in all_projects) {
      # Create output directories
      fastqc_path  <- file.path(qual_path, PRJ, "fastqc")
      multiqc_path <- file.path(qual_path, PRJ, "multiqc")
      
      dir.create(fastqc_path,  recursive = TRUE, showWarnings = FALSE)
      dir.create(multiqc_path, recursive = TRUE, showWarnings = FALSE)
      
      # Filter reads belonging to this project
      raw_path <- primers_n_samples$`Raw data path`[primers_n_samples$Project == PRJ] %>%
        unique()
      
      project_reads <- read_files[stringr::str_detect(read_files, 
                                                      paste0(raw_path, 
                                                             collapse = "|"))]
      
      # 1. Run FastQC (requires GNU parallel + FastQC installed)
            cmd_fastqc <- paste0("echo '", 
                                 paste(project_reads, collapse = "\n"), 
                                 "' | ",
                                 "parallel fastqc {1} -t 20 -o ", 
                                 fastqc_path)
            system(cmd_fastqc)
            
      # 2. Run MultiQC
            report_name <- paste0(PRJ, "--sequencing_quality-MultiQC_report.html")
             
            cmd_multiqc <- paste0("multiqc --interactive ",
                                  fastqc_path, 
                                  " --filename ",
                                  file.path(multiqc_path,
                                            report_name))
            system(cmd_multiqc)
            
      # 3. Show MultQC report
            browseURL(file.path(multiqc_path, report_name))
    }
    ```

### Primer detection Analysis

#### Identify paired-end read files

-   **List R1/R2 files.** Scans the raw data directory for FASTQ files matching common Illumina naming patterns for forward and reverse reads.

    ```{r}
    # Base directory of the read files (assumes all reads share the same root as the first file)
    data_dir <- dirname(read_files) %>% 
      unique()

    # List all forward (R1) FASTQ files using common naming patterns
    R1_pattern <- "_R1_001.fastq|_R1_001.fastq.gz|\\.1.fastq|\\.R1.fastq|R1.fq|R1.fastq|1.fq.gz|-R1.fastq.gz"

    all_fnFs <- list.files(data_dir, 
                           pattern = R1_pattern,
                           full.names = TRUE) %>%
      sort()

    # List all reverse (R2) FASTQ files using common naming patterns
    R2_pattern <- "_R2_001.fastq|_R2_001.fastq.gz|\\.2.fastq|\\.R2.fastq|R2.fq|R2.fastq|2.fq.gz|-R2.fastq.gz"

    all_fnRs <- list.files(data_dir, 
                           pattern = R2_pattern,
                           full.names = TRUE) %>%
      sort()

    # Final verification message
    n_R1 <- length(all_fnFs)
    n_R2 <- length(all_fnRs)

    if (n_R1 == n_R2 && n_R1 > 0) {
      message("✓ Paired: ", n_R1, " R1 files match exactly the ", n_R2, " R2 files.\n", 
              "\nVerify if all expected pairs are listed:\n", 
              "\nR1 files:\n", paste0("  ", basename(all_fnFs), collapse = "\n"),
              "\n",
              "\nR2 files:\n", paste0("  ", basename(all_fnRs), collapse = "\n"))
    } else {
      message("✗ MISMATCH: ", n_R1, " R1 files vs ", n_R2, " R2 files.",
              "\nR1 files:\n", paste0("  ", basename(all_fnFs), collapse = "\n"),
              "\n",
              "\nR2 files:\n", paste0("  ", basename(all_fnRs), collapse = "\n"),
              "\nCheck file names in `primers_n_samples` match actual FASTQ files on disk.")
    }
    ```

-   **Match files to samples.** Associates each sample in primers_n_samples with its corresponding R1/R2 files by matching directory path and Unique_File_name.

    ```{r}
    # Initialize sample table with metadata and empty columns for R1/R2 paths
    sample_idx_tbl_wide <- primers_n_samples %>%
      dplyr::select("Unique_File_name", "Lib", "Run", "Project",
                    "Researcher", "Sample", "Primer", "Type",
                    "Raw data path", dplyr::ends_with("ontrol")) %>%
      dplyr::mutate(FWD_R1 = NA_character_,
                    FWD_R2 = NA_character_)

    View(sample_idx_tbl_wide)

    # For each sample, find the corresponding R1 and R2 files by matching path + Unique_File_name
    for (i in seq_len(nrow(sample_idx_tbl_wide))) {
      
      pattern_i <- paste0(sample_idx_tbl_wide$`Raw data path`[i], "/",
                          sample_idx_tbl_wide$Unique_File_name[i], "_")
      
      # Match forward read (R1) file
      sample_idx_tbl_wide$FWD_R1[i] <- 
        all_fnFs[BiocGenerics::grep(pattern_i, all_fnFs)] %>%
        BiocGenerics::unique() %>% .[1]
      
      # Match reverse read (R2) file
      sample_idx_tbl_wide$FWD_R2[i] <- 
        all_fnRs[BiocGenerics::grep(pattern_i, all_fnRs)] %>%
        BiocGenerics::unique() %>% .[1]
      }

    # Show table
    View(sample_idx_tbl_wide)
    ```

-   **Build long-format file table.** Adds paths for future paired-output files and reshapes the table to long format, with one row per file per stage.

    ```{r}
        # Create paths for the future paired-output files and reshape to long format
    sample_idx_tbl <- 
      sample_idx_tbl_wide %>%
      dplyr::mutate(FWD_R1_paired = file.path(paired_libs,
                                              paste0(Unique_File_name, "--FWD_R1_paired.fastq.gz")),
                    FWD_R2_paired = file.path(paired_libs,
                                              paste0(Unique_File_name, "--FWD_R2_paired.fastq.gz"))) %>%
      tidyr::pivot_longer(
        cols = c(FWD_R1, FWD_R2, FWD_R1_paired, FWD_R2_paired),
        names_to  = "Stage",
        values_to = "Read file") %>%
      dplyr::mutate(Unique_File_name_Primer = Unique_File_name)

    # Show table
    View(sample_idx_tbl)
    ```

#### Primer database and orientations

-   **Primer database.** Primer names and sequences (with IUPAC codes where applicable). Comments are retained only where they document the biological identity of the primers.

    ```{r}
    # Define master vector of active primers (uncomment lines to activate specific primer pairs)
    all_primers <- 
      c(
        NeoFish_FWD = "CGCCGTCGCAAGCTTACCCT",
        NeoFish_REV = "AGTGACGGGCGGTGTGTGC",
        MiFish_FWD = "GTCGGTAAAACTCGTGCCAGC",
        MiFish_REV = "ACATAGTGGGGTATCTAATCCCAGTTTG"
        # MiBird__FWD = "GGGTTGGTAAATCTTGTGCCAGC",
        # MiBird__REV = "CATAGTGGGGTATCTAATCCCAGTTTG",
        # MiFish variants (12S, fishes)
        # MiFishEM__FWD  = "GTCGGTAAAACTCGTGCCAGC", 
        # MiFishEMs__FWD = "GTCGGTAAAACTCGTGCCAGC",
        # MiFish__FWD    = "AAACTCGTGCCAGC",       
        # MiFish2__REV   = "CATAGTGGGGTATCTAATCCCAGTTTG",
        # MiFishEMs__REV = "CATAGTGGGGTATCTAATCCCAGTT",
        # MiFishEM__REV  = "CATAGTGGGGTATCTAATCCCAGTTTG",
        # MiFish__REV    = "TCTAATCCCAGTTTG",
        # p12SV5__FWD = "TTAGATACCCCACTATGC",
        # p12SV5__REV = "TAGAACAGGCTCCTCTAG",
        # V12Su__FWD = "GTGCCAGCNRCCGCGGTYANAC",
        # V12Su__REV = "ATATRGGGTATCTAATCCYAGT",
        # V16Su__FWD = "ACGAGAAGACCCYRYGRARCTT",
        # V16Su__REV = "TCTHRRANAGGATTGCGCTGTTA",
        # VCOIu__FWD = "CAYGCHTTTGTGNATRATYYTYTT",
        # VCOIu__REV = "GGRGGRTADACDGTYCANCCNGT"
        # (Other primer pairs can be re‑added here following the same pattern)
        ) %>% 
      stats::na.omit()

    # Convert primer vector to list of single-row tibbles (for allOrients function)
    primers_list <- tibble::tibble(`Primer seq`  = all_primers,
                                   `Primer name` = names(all_primers)) %>%
      split(1:nrow(.))

    primers_list

    # Create reference table with IUPAC → Regex conversion for inspection
    all_primers_tbl <- tibble::tibble(
      Primer = names(all_primers),
      `IUPAC sequence` = all_primers) %>%
      dplyr::mutate(`Regex sequence` = iupac2seq(`IUPAC sequence`))

    # Show primers
    all_primers_tbl
    ```

-   **Generate all primer orientations** Produces all orientations (Forward/Reverse/Complement/RevComp.) and their regex representation for downstream searching.

    ```{r}
    # Apply all Orients to all primers and clean final table
    primers_all_orients <- purrr::map_dfr(primers_list, 
                                          allOrients) %>%
      dplyr::mutate(`Primer pair` = stringr::str_remove_all(Primer, 
                                                            "__.*.$"))
    # Inspect primers orientations
    View(primers_all_orients)
    ```

#### Per-file primer hits

-   **Define primer counting function.** The function `multi_primerHits()` applies `primerHits()` across all primer orientations for a single FASTQ file, returning a table of hit counts per orientation. Defining it here keeps the primer-counting logic close to where it is used.

    ```{r}
    multi_primerHits <- function(Read_file, primers) {
      primer_counts <- purrr::map_df(primers,
                                     .f = primerHits,
                                     file_name = Read_file)
      primer_counts %>%
        dplyr::mutate(`Read file` = Read_file)
    }
    ```

-   **Per‑file primer hits.** Counts the number of reads that contain a given primer pattern (with up to 1 mismatch). All four orientations generated by `allOrients()` — Forward, Complement, Reverse, and RevComp — are searched simultaneously against each R1 and R2 file, enabling detection of primers regardless of read orientation or library preparation strategy. The resulting table is then joined with sample metadata and annotated with total read counts per file to allow proportional comparisons across samples.

    ```{r}
    # Select only raw reads (R1/R2) for primer detection
    reads_seqs_tbl <- 
      sample_idx_tbl %>%
      dplyr::filter(Stage %in% c("FWD_R1", "FWD_R2")) %>%
      dplyr::select(`Read file`, Unique_File_name)

    # Create named vector of FASTQ paths (names = sample identifiers)
    reads_seqs <- reads_seqs_tbl$`Read file`
    names(reads_seqs) <- reads_seqs_tbl$Unique_File_name

    # Extract all primer orientations as named vector for counting
    primers_seqs <- primers_all_orients %>%
      dplyr::pull(Sequence, name = `Orientation name`)

    # Set up parallel processing (all cores minus 2)
    cores_to_use <- future::availableCores() - 2
    future::plan(future::multisession(workers = cores_to_use))

    message("Searching primers across ", length(reads_seqs), " files using ",
            cores_to_use, " cores. This may take a few minutes...")

    # Count primer occurrences across all files x all primer orientations (parallel)
    primers_in_Nreads <- furrr::future_map_dfr(reads_seqs, 
                                               .f = multi_primerHits,
                                               primers = primers_seqs,
                                               .options = furrr::furrr_options(seed = NULL))

    # Join with sample metadata and calculate total reads per file
    primers_in_Nreads <- 
      primers_in_Nreads %>%
      dplyr::left_join(sample_idx_tbl, by = "Read file") %>%
      dplyr::mutate(Read = dplyr::if_else(stringr::str_detect(Stage, "R1"), "R1", "R2")) %>%
      tidyr::unite(col = "Unique_File_name_Read",
                   Unique_File_name,
                   Read,
                   sep = " ",
                   remove = FALSE) %>%
      dplyr::mutate(`Total reads` = ShortRead::countFastq(dirPath = `Read file`)[, 1])

    message("✓ Primer search complete. Results in `primers_in_Nreads`:\n",
            "  ", length(unique(primers_in_Nreads$Unique_File_name)), " samples × ",
            length(primers_seqs), " primer orientations across ",
            length(reads_seqs), " files.")
    ```

#### Primers identified in each sample

-   **Prepare data.** Transform the primer hits table into a long format suitable for visualization with ggplot2. This step filters out unobserved primers and organizes metadata.

    ```{r}
    # Prepare primer counts for plots in ggplot

    # Check available orientation names (for debugging)
    # primers_all_orients$`Orientation name`

    # Print column names formatted for selection (helper for coding)
    # primers_in_Nreads %>% colnames() %>% paste0(collapse = '",\n') %>% cat()

    # Convert primer hits table to long format
    primers_in_Nreads_long <- primers_in_Nreads %>% 
      dplyr::select(!dplyr::ends_with(c("_Complement", "_Reverse"))) %>% 
      tidyr::gather(key = Sequences,
                    value = Count,
                    ends_with("FWD_Forward"),
                    ends_with("FWD_RevComp"),
                    ends_with("REV_Forward"),
                    ends_with("REV_RevComp")) %>% 
      dplyr::mutate(Sequences = as.factor(Sequences),
                    Unique_File_name = factor(Unique_File_name, 
                                              levels = sample_levels)) %>% 
      dplyr::select(-c("Read file", "Stage")) %>% 
      # Remove columns that are entirely NA
      dplyr::select(where(function(x) any(!is.na(x)))) %>% 
      # Filter out primers with zero counts
      dplyr::filter(Count != 0)

    #View table
    View(primers_in_Nreads_long)
    ```

-   **Configure plot settings.** Load necessary visualization libraries and define the factor levels for sample ordering in the plots.

    ```{r}
    # Create levels for Files display on plot based on the prepared data
    Unique_File_name_Read_levels <- primers_in_Nreads_long %>% 
      dplyr::pull(Unique_File_name_Read) %>% 
      base::unique()
    ```

-   **Generate diagnostic plot.** Iterate through each project to create specific primer abundance plots. The loop filters data per project, generates a tile/bar plot showing primer presence and proportion, and saves the output as a PDF.

    ```{r}
    # Loop through all projects defined in the workflow
    for (PRJ in all_projects) {
      
      options(scipen = 10000)

      # 1. Filter and prepare data ONCE for the current project
      # This avoids filtering the large dataset multiple times inside the loop
      project_data <- primers_in_Nreads_long %>%
        dplyr::filter(Project == PRJ) %>%
        dplyr::mutate(
          
          # Pre-format labels and calculate percentages to keep ggplot clean
          Sequences_lbl = stringr::str_replace_all(Sequences, "__|_", "\n"),
          Primer_lbl    = stringr::str_replace_all(Primer, ";", "\n"),
          Pct_Reads     = (Count / `Total reads` * 100))
      
      # 2. Calculate dynamic dimensions based on actual data
      n_samples <- length(unique(project_data$Unique_File_name_Read))
      n_primers <- length(unique(project_data$Primer))

      # 3. Create the plot object
      primers_tile <- project_data %>%
        ggplot2::ggplot(aes(y = Unique_File_name_Read)) +
        # Bar layer
        ggplot2::geom_bar(aes(x = `Total reads`,
                              fill = Pct_Reads,
                              color = Primer,
                              group = Sequences),
                          stat = "identity",
                          position = "dodge",
                          linewidth = 0.05,
                          linetype = 2,
                          alpha = 0.5) +
        # Text labels for counts
        ggplot2::geom_text(aes(x = `Total reads`, 
                               label = Count),
                           hjust = 0, 
                           size = 1) +
        # Color scales
        ggplot2::scale_fill_gradientn(name = "Proportion of reads\n     with primer (%)",
                                      colours = c("white", "red", "yellow", 
                                                  "green", "dark green"),
                                      values = c(0, 1),
                                      na.value = "white") +
        # Axes and Guides
        ggplot2::scale_y_discrete(limits = rev) +
        ggplot2::scale_x_log10(
          breaks = c(10, 100, 1000, 10000, 100000),
          expand = expansion(mult = c(0, 0.2))) +
        ggplot2::scale_alpha(guide = 'none') +
        ggplot2::guides(color = guide_legend(override.aes = list(fill = "white", 
                                                                 size = 10))) +
        
        # Theme adjustments
        ggplot2::theme_light(base_line_size = 0.025, base_size = 6) +
        ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1),
                       strip.text.y = element_text(size = 10),
                       plot.title = element_text(size = 10),
                       plot.subtitle = element_text(size = 6),
                       legend.text = element_text(size = 5),
                       legend.title = element_text(size = 7)) +
        
        # Dynamic horizontal lines based on actual sample count
        ggplot2::geom_hline(yintercept = seq(0.5, n_samples + 0.5, 2), 
                            color = "darkgrey") +
        
        # Titles and Facets
        ggplot2::labs(title = PRJ,
                      subtitle = "Presença de primers nas amostras:\nintensidade de cor relativa à contagem do respectivo primer no conjunto de reads\n(max. mismatch = 1)",
                      x = "Primers",
                      y = "Amostra") +
        ggplot2::facet_grid(rows = vars(Primer_lbl),
                            cols = vars(Sequences_lbl),
                            scales = "free",
                            space = "free")

      # 4. Save the plot
      ggplot2::ggsave(file = paste0(results_path, "/", 
                                    PRJ, 
                                    "-primers_found_in_reads_1e.pdf"),
                      plot = primers_tile,
                      device = "pdf",
                      width = 10 + (n_primers * 10),
                      height = (n_samples * 0.3) + 3,
                      units = "cm",
                      limitsize = FALSE,
                      dpi = 600)
       
      browseURL(paste0(results_path, "/", 
                          PRJ, 
                          "-primers_found_in_reads_1e.pdf"))
      }
    ```

-   **Embed PDF results.** Dynamically list and display the generated PDF plots within the HTML report.

    ```{r}
    # Extract the PDF path from the R object
    pdf_info <- list.files(path = results_path,
                           pattern = "primers_found_in_reads",
                           full.names = TRUE)

    # Generate the iframe HTML code
    if(length(pdf_info) > 0) {
      cat(paste0('<iframe src="', pdf_info, '" width="1200" height="800"></iframe>'))
      } else {
        cat("No primer plots found.")
        }
    ```

-   **Check orientations.** Analyze the generated plots to confirm that primers appear in the expected read files and orientations. The correct interpretation depends on two factors: **how the sequencing library was constructed** (standard vs. mixed orientation) and **whether read-through occurs** (i.e., whether the amplicon is shorter than the read length).

    **Understanding the four orientation columns.** The `allOrients()` function generates four variants of each primer sequence: `Forward` (original sequence), `Complement` (same direction, complementary bases), `Reverse` (original read 3'→5'), and `RevComp` (reverse complement). Only `Forward` and `RevComp` are biologically observable in Illumina sequencing — the polymerase always synthesizes 5'→3', so a primer can only appear at the start of a read in its `Forward` form, or at the end of a read in its `RevComp` form (due to read-through). Any counts in `Complement` or `Reverse` columns at `max.mismatch = 1` are near-chance artifacts of short sequence similarity and **must be treated as zero**.

    **Case 1 — Standard library, amplicon longer than the read length (no read-through).** In a standard Illumina paired-end library, R1 sequences the forward strand 5'→3' (starting from where the FWD primer bound), and R2 sequences the reverse strand 5'→3' (starting from where the REV primer bound). When the amplicon is long enough that neither read reaches the opposite primer, the expected result is:

    | Primer  | R1                 | R2                 |
    |:--------|:-------------------|:-------------------|
    | **FWD** | **Forward** (good) | \~0                |
    | **REV** | \~0                | **Forward** (good) |

    All other orientation/read combinations should be near zero. *Typical examples:* 16S V3–V4 (\~460 bp) with 2×250 bp reads; 12S V5 with 2×150 bp reads when the amplicon exceeds read length.

    **Case 2 — Standard library, amplicon shorter than the read length (read-through occurs).** When the amplicon is shorter than the read length, R1 reads through the entire insert and into the REV primer site, encountering it from the antisense direction — it therefore appears as `RevComp` at the 3' end of R1. Symmetrically, R2 reads through and encounters the FWD primer site as `RevComp` at its 3' end. The expected result is:\[\^1\]

    | Primer  | R1                                       | R2                                       |
    |:-----------------|:--------------------------|:--------------------------|
    | **FWD** | **Forward** (good, 5' start of R1)       | **RevComp** (read-through, 3' end of R2) |
    | **REV** | **RevComp** (read-through, 3' end of R1) | **Forward** (good, 5' start of R2)       |

    The `RevComp` read-through signal is not an error — it is expected and **must** be handled by Cutadapt, which trims both the 5' primer and the 3' reverse-complement of the opposite primer from each read. The proportion of reads with read-through signal will vary with the length distribution of the target region: short amplicons will show near-equal counts in `Forward` and `RevComp`, while longer ones will show a weaker `RevComp` signal. *Typical examples:* ITS1/ITS2 fungal barcodes; short COI barcode fragments; short 16S V4 amplicons sequenced with 2×300 bp reads.

    **Case 3 — Mixed orientation library.** Some library preparation protocols (e.g., HiSeq/NovaSeq workflows without strand-selection, certain ligation-based preps, or pooled multiplexing strategies) produce reads where approximately half the read pairs have FWD on R1 and the other half have REV on R1. In this scenario, the plot will show roughly equal counts of both primers in both read files in `Forward` orientation:

    | Primer  | R1                           | R2                           |
    |:--------|:-----------------------------|:-----------------------------|
    | **FWD** | **Forward** (\~50% of reads) | **Forward** (\~50% of reads) |
    | **REV** | **Forward** (\~50% of reads) | **Forward** (\~50% of reads) |

    This is not an error — it reflects the library design. However, DADA2 assumes that all R1 reads begin with the FWD primer and all R2 reads begin with the REV primer; mixed-orientation input will produce catastrophic failures at the merging step. Before proceeding, reads must be reoriented using one of the following strategies:

    -   **`vsearch --orient`**: Reorients each read relative to a reference database before any further processing, producing a consistently stranded output. Can be applied to R1 and R2 independently before they enter the pipeline.

    -   **`dada2::removePrimers(..., orient = TRUE)`**: When using DADA2's own primer removal function, the `orient` parameter detects read pairs where the FWD primer is found at the start of R2 instead of R1, swaps the pair, and unifies orientation — no external tool required.\[\^3\]

    -   **Dual Cutadapt strategy (recommended for maximum read retention)**: Run Cutadapt twice — once with FWD anchored on R1 and REV on R2 (`standard` pairs), and once with REV anchored on R1 and FWD on R2 (`flipped` pairs), using `--discard-untrimmed` on each run to keep only correctly matched pairs. Run the full DADA2 pipeline independently on each subset, then merge the two resulting ASV tables, reverse-complementing all sequences from the `flipped` run to unify orientation before merging. This is the approach recommended by B. Callahan for mixed-orientation data.

    **Troubleshooting: diagnosing orientation anomalies.**

    | Pattern observed in the plot                                                                 | Likely cause                                                                                                | Recommended fix                                                                                                                                               |
    |:------------------|:---------------------|:------------------------------|
    | FWD in R2/**Forward** (strong); REV in R1/**Forward** (strong)                               | R1/R2 file paths are swapped, or FWD/REV primer definitions are swapped in `primersnsamples`                | Swap R1↔R2 path assignments in `sampleidxtbl`, or swap FWD and REV primer entries                                                                             |
    | FWD in R1/**RevComp** (strong); REV in R2/**RevComp** (strong); near-zero **Forward** signal | Both primer sequences were entered in reverse-complement direction (e.g., copied from the antisense strand) | Replace both: `FWD <-     FWD.orient[["RevComp"]]` and `REV <- REV.orient[["RevComp"]]`                                                                       |
    | REV in R2/**RevComp** instead of R2/**Forward** (FWD signal looks correct)                   | REV primer sequence was entered as its own reverse complement                                               | Replace: `REV <-     REV.orient[["RevComp"]]` before proceeding \[\^1\]                                                                                       |
    | FWD and REV in both R1/**Forward** and R2/**Forward** in roughly equal proportions           | Mixed orientation library                                                                                   | Use `vsearch --orient`, `removePrimers(orient = TRUE)`, or the dual Cutadapt strategy described above \[\^2\]                                                 |
    | No primer detected in any orientation in any file                                            | Wrong primer pair selected; reads are already pre-trimmed; mismatches \> 1; incorrect file paths            | Verify primer sequences against the wet-lab protocol; confirm reads are not pre-trimmed (check FastQC/MultiQC report); increase `max.mismatch` to 2 as a test |

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "2_raw_data_pre_processing-", Sys.Date()))
    ```

## 3. DADA2 ASV Workflow

This section executes the DADA2 core workflow to infer Amplicon Sequence Variants (ASVs) from the pre-processed reads. The workflow is organized into two phases:

1.  **Read preparation**: ambiguous bases are removed, primers are physically excised with Cutadapt, their removal is verified, and a final quality filter is applied.

2.  **DADA2 inference steps**: a run-specific error model is learned, reads are denoised and merged, and chimeric sequences are removed, resulting in a high-resolution ASV table analogous to a traditional OTU table.

### Read Preparation and Primer Removal

Before DADA2 can learn error rates and infer ASVs, reads must be free of ambiguous bases, primer sequences, and low-quality bases. This preparation is carried out in four sequential steps: a pre-filter to remove reads containing Ns, primer removal with Cutadapt, a sanity check confirming successful trimming, and a final quality filter applied to primer-free reads.

#### Pre-filtering reads

DADA2 requires sequences to be completely free of ambiguous nucleotides (Ns) to accurately perform its core sample inference. Furthermore, the presence of Ns can interfere with accurate primer recognition and mapping during the Cutadapt step. Following the official DADA2 ITS workflow, this pre-filtering step is executed strictly to discard read pairs containing any Ns (**`maxN = 0`**). Comprehensive quality filtering, error tracking (such as **`maxEE`**), and length truncation are purposefully deferred until after primer removal.

-   **Prepare file paths.** Identify paired-end read files that contain primers and organize their paths for the filtering step.

    ```{r}
    # Select files with identified primers and reshape to wide format for processing
    primer_paired_files <- sample_idx_tbl %>%
      dplyr::filter(stringr::str_detect(string = Stage,
                                        pattern =
                                          "FWD_R1|FWD_R2|FWD_R1_paired|FWD_R2_paired"
                                        )) %>% 
      tidyr::pivot_wider(id_cols = c("Unique_File_name", "Primer", "Lib"),
                         names_from = "Stage",
                         values_from = "Read file")

    # Verify basenames (optional check)
    # primer_paired_files$FWD_R1 %>% BiocGenerics::basename()
    # primer_paired_files$FWD_R1_paired %>% basename()
    ```

-   **Define pre-filtering function.** Create a wrapper function for filterAndTrim to handle forward and reverse reads and removing ambiguous bases (Ns).

    ```{r}
    # Custom function to repair and filter reads using DADA2
    pre_filter_Ns <- function(Unique_File_name,
                              FWD_R1_in,
                              FWD_R2_in,
                              FWD_R1_out,
                              FWD_R2_out) {
      # Check if input files exist
      if (file.exists(FWD_R1_in) && file.exists(FWD_R2_in)) {
        
        # Run filterAndTrim with ONLY maxN = 0 to prepare for Cutadapt
        sample_filtered_out <- dada2::filterAndTrim(fwd = FWD_R1_in,
                                                    filt = FWD_R1_out,
                                                    rev = FWD_R2_in,
                                                    filt.rev = FWD_R2_out,
                                                    maxN = 0,
                                                    multithread = TRUE,
                                                    compress = TRUE,
                                                    verbose = TRUE)
        
        # Format output
        tibble::as_tibble(sample_filtered_out) %>%
          dplyr::mutate(Unique_File_name = Unique_File_name)
        
        } else {
          warning(paste("Missing files for:", Unique_File_name))
          return(tibble::tibble())
        }
      }
    ```

-   **Execute filtering.** Run the pre-filtering function in parallel across all samples and generate a summary table of read retention.

    ```{r}
    # Setup parallel processing
    cores_to_use <- future::availableCores() - 2 
    future::plan(future::multisession(workers = cores_to_be_used))

    # Run pre-filtering in parallel
    {
      tictoc::tic()
      
      all_filtered_out_fun <-
        furrr::future_pmap_dfr(list(primer_paired_files$Unique_File_name,
                                    primer_paired_files$FWD_R1,
                                    primer_paired_files$FWD_R2,
                                    primer_paired_files$FWD_R1_paired,
                                    primer_paired_files$FWD_R2_paired),
                               .f = pre_filter_Ns,
                               .options = furrr::furrr_options(seed = TRUE))
      tictoc::toc()
    }

    # Generate summary table
    pre_filtered_summary <- all_filtered_out_fun %>%
      dplyr::mutate(Proportion = round((reads.out / reads.in * 100), 
                                       digits = 2)) %>% 
      dplyr::rename("Raw reads (pairs)" = "reads.in",
                    "Filtered reads"    = "reads.out")

    # View result
    pre_filtered_summary
    ```

#### Remove Primers with Cutadapt

The Cutadapt software (DOI:10.14806/ej.17.1.200) is used to remove primer sequences from reads, also capable of handling read-through scenarios.

-   **Configure output paths.** Map sample names to new Cutadapt output filenames, ensuring separation by primer set if necessary.

    ```{r}
    # Map sample names to Cutadapt output files (identify paired samples for Cutadapt input)
    paired_files_per_primer <- sample_idx_tbl %>% 
      dplyr::filter(Stage %in% c("FWD_R1_paired", 
                                 "FWD_R2_paired")) %>% 
      dplyr::filter(file.exists(`Read file`))

    # Generate Cutadapt output paths by replacing 'paired' with 'cutadapt'
    cut_files_per_primer <- paired_files_per_primer %>%  
      dplyr::mutate(Stage = gsub(Stage, pattern = "_paired", 
                                 replacement = "_cutadapt"),
                    `Read file` = stringr::str_replace_all(string = `Read file`,
                                                           pattern = "paired",
                                                           replacement = "cutadapt"))

    # Combine input (paired) and output (cutadapt) metadata
    cut_files_per_primer <- dplyr::bind_rows(cut_files_per_primer,     
                                             paired_files_per_primer)

    # OPTIONAL Separate paths by primer if multiple primers exist per sample 
    # cut_files_per_primer <- cut_files_per_primer %>% 
    #   tidyr::separate_longer_delim(cols = "Primer", 
    #                                delim = ";") %>% 
    #   dplyr::mutate(`Read file` = dplyr::case_when(
    #     Stage %in% c("FWD_R1_cutadapt", 
    #                  "FWD_R2_cutadapt") ~ 
    #       stringr::str_replace_all(string = `Read file`,
    #                                pattern = paste0(cutadapt_libs, "/"),
    #                                replacement = paste0(cutadapt_libs, "/", 
    #                                                     Primer, "--")),
    #     TRUE ~ `Read file`))
    ```

-   **Extract primer sequences.** For each primer set, retrieve the forward and reverse primer sequences — and their reverse complements — from the orientations table. These are needed to construct the correct adapter flags for both R1 and R2 reads, accounting for read-through at the 3′ end.

    ```{r}
    # Identify unique primer sets to process
    primer_sets <- primers_all_orients$`Primer pair` %>% 
      BiocGenerics::unique()

    # Initialize list to store flags per primer set
    cutadapt_flags <- list()

    # Store flags using grep to find specific orientation keys constructed in 'allOrients' function
    for (PRIMER_SET in primer_sets) {
      
      # Forward primer (for R1 5')
      fwd_seq <- primers_all_orients$Sequence[
        grep(paste0(str_extract(PRIMER_SET, "^[^_]+"), "_FWD_Forward"), 
         primers_all_orients$`Orientation name`)]
      
      # Reverse complement of Forward primer (for R2 3' read-through)
      fwd_rc_seq <- primers_all_orients$Sequence[
    grep(paste0(str_extract(PRIMER_SET, "^[^_]+"), "_FWD_RevComp"), 
         primers_all_orients$`Orientation name`)]
      
      # Reverse primer (for R2 5')
      rev_seq <- primers_all_orients$Sequence[
    grep(paste0(str_extract(PRIMER_SET, "^[^_]+"), "_REV_Forward"), 
         primers_all_orients$`Orientation name`)]
      
      # Reverse complement of Reverse primer (for R1 3' read-through)
      rev_rc_seq <- primers_all_orients$Sequence[
    grep(paste0(str_extract(PRIMER_SET, "^[^_]+"), "_REV_RevComp"), 
         primers_all_orients$`Orientation name`)]
      
      # Store sequences for use in flag construction
      cutadapt_flags[[PRIMER_SET]] <- list(fwd_seq    = fwd_seq,
                                           fwd_rc_seq = fwd_rc_seq,
                                           rev_seq    = rev_seq,
                                           rev_rc_seq = rev_rc_seq)
      }
    ```

-   **Build adapter flags.** Define the -g, -a, -G, and -A Cutadapt flags for each primer set.

    -   **`-g`**: Removes the forward primer from the 5′ end of R1. Use this when the forward primer is expected at the start of the forward read.

    -   **`-a`**: Removes the reverse complement of the reverse primer from the 3′ end of R1. Only relevant when read-through occurs and the sequencing read extends beyond the amplicon insert into the opposite primer.

    -   **`-G`**: Removes the reverse primer from the 5′ end of R2. Mirrors the -g logic for the reverse read.

    -   **`-A`**: Removes the reverse complement of the forward primer from the 3′ end of R2. Again, only active when read-through is present.

    ```{r}
        for (PRIMER_SET in primer_sets) {
          
          seqs <- cutadapt_flags[[PRIMER_SET]]
          
          # -g: Anchor FWD to start of R1; -a: Remove REV_RC from end of R1
          R1_flags <- paste0("-g", seqs$fwd_seq, 
                             "-a", seqs$rev_rc_seq)
          
          # -G: Anchor REV to start of R2; -A: Remove FWD_RC from end of R2
          R2_flags <- paste0("-G", seqs$rev_seq, 
                             "-A", seqs$fwd_rc_seq)
          
          # Store back for use in the execution step
          cutadapt_flags[[PRIMER_SET]]$R1_flags <- R1_flags
          cutadapt_flags[[PRIMER_SET]]$R2_flags <- R2_flags
          }
    ```

-   **Prepare file lists per primer set.** Filter the combined metadata table to the current primer set and pivot to wide format, separating input (paired) and output (cutadapt) file paths. Only samples with both R1 and R2 input files present on disk are retained.

    ```{r}
    for (PRIMER_SET in primer_sets) {
      
      # Filter samples belonging to this primer set and pivot to wide format
      PRIMER_SET_files <- cut_files_per_primer %>% 
        dplyr::filter(stringr::str_detect(Primer, str_extract(PRIMER_SET, "^[^_]+"))) %>%
        tidyr::pivot_wider(id_cols = c("Unique_File_name", "Unique_File_name_Primer"),
                           names_from  = "Stage",
                           values_from = "Read file")

      # Validate input file existence
      valid_indices <- file.exists(PRIMER_SET_files$FWD_R1_paired) & 
                   file.exists(PRIMER_SET_files$FWD_R2_paired)
      
      # Define input/output vectors restricted to valid pairs
      cutadapt_flags[[PRIMER_SET]]$fnFs_in  <- PRIMER_SET_files$FWD_R1_paired[valid_indices]
      
      cutadapt_flags[[PRIMER_SET]]$fnRs_in  <- PRIMER_SET_files$FWD_R2_paired[valid_indices]
      
      cutadapt_flags[[PRIMER_SET]]$fnFs_out <- PRIMER_SET_files$FWD_R1_cutadapt[valid_indices]
      
      cutadapt_flags[[PRIMER_SET]]$fnRs_out <- PRIMER_SET_files$FWD_R2_cutadapt[valid_indices]
      }
    ```

-   **Execute Cutadapt.** Iterate over each valid input file pair and invoke the Cutadapt **`system2`** command. **Key parameters to configure for your dataset**:

    -   **`-e`**: Maximum error rate for adapter matching. Use lower values (e.g., 0.05) for primers without degenerate bases; values around 0.1 (10%) are typical when IUPAC ambiguity codes are present.

    -   **`-j`**: Number of CPU threads. Set to `0` to let Cutadapt auto-detect all available cores.

    -   **`-n`**: Number of adapter trimming rounds per read. Set to `1` for standard libraries; increase if read-through could place a primer and its reverse complement on the same read.

    -   **`--minimum-length`**: Discard reads shorter than this value after trimming. Prevents short artefactual fragments from reaching DADA.

    -   **`--discard-untrimmed`**: Removes any pair in which no adapter was detected. Appropriate when all reads are expected to originate from the target amplicon; omit if you want to retain undetected reads.

    -   **`--match-read-wildcards`**: Enables IUPAC degenerate base matching in adapter sequences. Required for primers containing ambiguous positions.

    -   **`--pair-filter`**: Determines when a pair is discarded. **`any`** (default) removes the pair if either read fails; both removes it only if both fail. When combined with **`--discard-untrimmed`**, always prefer any to avoid retaining pairs where only one read carries the expected primer.

    ```{r}
    for (PRIMER_SET in primer_sets) {
      
      flags <- cutadapt_flags[[PRIMER_SET]]
      fnFs_in <- flags$fnFs_in
      fnRs_in <- flags$fnRs_in
      fnFs_out <- flags$fnFs_out
      fnRs_out <- flags$fnRs_out
      
      if (length(fnFs_in) > 0) {
        for (i in seq_along(fnFs_in)) {
          system2(command = cutadapt, 
                  args = c(flags$R1_flags, 
                           flags$R2_flags,
                           "-e", "0.1",
                           "-j", "70",
                           "-n", "5",
                           "--minimum-length", "30",
                           "--discard-untrimmed",
                           "--match-read-wildcards",
                           # "--pair-filter = both",
                           "-o", fnFs_out[i],
                           "-p", fnRs_out[i], fnFs_in[i], fnRs_in[i]
                  ))
          }
        } else {
          message(paste("No valid input files found for primer set:", PRIMER_SET))
        }
      }
    ```

#### Primer Removal Verification

After running Cutadapt, it is essential to confirm that primer sequences were effectively removed from all reads before proceeding to quality filtering and ASV inference. Any residual primer signal in the trimmed files indicates misconfigured adapter flags, unexpected primer orientations, or issues in the library preparation. This section reuses the primer‑detection logic to count residual primer hits in the Cutadapt outputs and summarizes how clean each sample is.

-   **Collect Cutadapt output files.** Assemble a table of trimmed FASTQ files (R1/R2) per sample and primer set.

    ```{r}
    # Restrict to Cutadapt outputs (R1/R2 trimmed files that actually exist)
    cutadapt_check_files <- cut_files_per_primer %>%
      dplyr::filter(stringr::str_detect(Stage, "_cutadapt")) %>%
      dplyr::filter(file.exists(`Read file`)) %>%
      tidyr::pivot_wider(id_cols = c("Unique_File_name", 
                                     "Unique_File_name_Primer",
                                     "Primer", 
                                     "Run"),
                     names_from  = "Stage",
                     values_from = "Read file")

    # View result
    cutadapt_check_files
    ```

-   **Prepare primer pairs table.** Build a compact reference table with one row per primer pair, containing the forward and reverse sequences in their biologically relevant orientations.

    ```{r}
    primer_pairs_tbl <- primers_all_orients %>% 
      dplyr::filter(`Primer orientation` == "Forward") %>%
      dplyr::select(`Primer pair`, Primer, Sequence) %>%
      tidyr::pivot_wider(id_cols = `Primer pair`,
                         names_from = Primer,
                         values_from = Sequence,
                         names_sep = "_")
    # View result
    primer_pairs_tbl
    ```

-   **Define helper function to check one sample.** For a given sample and its R1/R2 Cutadapt files, count residual hits of each primer pair.

    ```{r}
    check_one_sample_cutadapt <- function(sample_name,
                                          R1_file,
                                          R2_file,
                                          primer_pairs_tbl) {
      
      purrr::map_dfr(primer_pairs_tbl$`Primer pair`, 
                     function(PRIMER_SET) {
                       
                       fwd_seq <- primer_pairs_tbl %>%
                         dplyr::filter(`Primer pair` == PRIMER_SET) %>%
                         dplyr::pull(dplyr::ends_with("FWD")) %>%
                         .[[1]]
                       
                       rev_seq <- primer_pairs_tbl %>%
                         dplyr::filter(`Primer pair` == PRIMER_SET) %>%
                         dplyr::pull(dplyr::ends_with("REV")) %>%
                         .[[1]]
                       
                       rbind(primerHits(primer = fwd_seq, file_name = R1_file),
                             primerHits(primer = rev_seq, file_name = R2_file)) %>%
                         dplyr::mutate(Sample = sample_name,
                                       `Primer pair` = PRIMER_SET,
                                       File = c("R1", "R2"))
                       }
      )
    }
    ```

-   **Count residual primer hits across all samples** Apply the helper function to every sample with Cutadapt outputs.

    ```{r}
    primers_in_cutadapt_reads <- 
      purrr::pmap_dfr(list(cutadapt_check_files$Unique_File_name_Primer,
                           cutadapt_check_files$FWD_R1_cutadapt,
                           cutadapt_check_files$FWD_R2_cutadapt),
                      check_one_sample_cutadapt,
                      primer_pairs_tbl = primer_pairs_tbl)

    # View result
    primers_in_cutadapt_reads
    ```

-   **Summarize and reshape for inspection.** Join metadata, compute total reads per file, and convert to long format to inspect which primer pairs still leave residual signal.

    ```{r}
    # Add metadata and total reads per file
    primers_in_cutadapt_reads <- primers_in_cutadapt_reads %>%
      dplyr::left_join(cutadapt_check_files, 
                       by = c("Sample" = "Unique_File_name_Primer")) %>%
      dplyr::mutate(`Total reads` = ShortRead::countFastq(dirPath = FWD_R1_cutadapt)[, 1] +
                      ShortRead::countFastq(dirPath = FWD_R2_cutadapt)[, 1])

    # Reshape to long format and filter non‑zero hits
    primers_in_cutadapt_reads_long <- primers_in_cutadapt_reads %>%
      dplyr::filter(File %in% c("R1", "R2")) %>%
      dplyr::mutate(Sample = factor(Sample, levels = Unique_File_name_Primer),
                    Pct_reads = (..density.. * 100))

    # View result
    primers_in_cutadapt_reads_long
    ```

-   **Generate diagnostic plot.** If residual primers are detected, visualize their distribution across samples.

    ```{r}
    # Only generate plots if residual primers found
    residual_hits <- primers_in_cutadapt_reads_long %>%
      dplyr::filter(Count > 0)

    if (nrow(residual_hits) > 0) {
      
      for (PRJ in all_projects) {
        
        options(scipen = 10000)
        
        # 1. Filter and prepare data for the current project
        # This avoids filtering the large dataset multiple times inside the loop
        project_data <- residual_hits  %>%
          dplyr::filter(Project == PRJ) %>%
          dplyr::mutate( 
            
            # Pre-format labels and calculate percentages to keep ggplot clean
            Sequences_lbl = stringr::str_replace_all(Sequences, "__|_", "\\n"),
            Primer_lbl = stringr::str_replace_all(Primer, ";", "\\n"),
            Pct_Reads = (Count / `Total reads` * 100))
        
        # 2. Calculate dynamic dimensions based on actual data
        n_samples <- length(unique(project_data$Sample))
        n_primers <- length(unique(project_data$`Primer pair`))
        
        # 3. Create the plot 
        primers_tile_residual <- project_data %>%
          ggplot2::ggplot(aes(y = Sample)) +
          # Bar layer
          ggplot2::geom_bar(aes(x = `Total reads`,
                                fill = Pct_Reads,
                                color = `Primer pair`,
                                group = Sequences),
                            stat = "identity",
                            position = "dodge",
                            linewidth = 0.05,
                            linetype = 2,
                            alpha = 0.5) +
          # Text labels for counts
          ggplot2::geom_text(aes(x = `Total reads`,
                                 label = Count),
                             hjust = 0,
                             size = 1) +
          # Color scales
          ggplot2::scale_fill_gradientn(
            name    = "Residual primer\nproportion (%)",
            colours = c("white", "red", "yellow", "green", "dark green"),
            values  = c(0, 1),
            na.value = "white") +
          # Axes and Guides
          ggplot2::scale_y_discrete(limits = rev) +
          ggplot2::scale_x_log10(
            breaks = c(10, 100, 1000, 10000, 100000),
            expand = expansion(mult = c(0, 0.2))) +
          ggplot2::guides(color = guide_legend(
            override.aes = list(fill = "white", size = 10))) +
          
          # Theme adjustments
          ggplot2::theme_light(base_line_size = 0.025, base_size = 6) +
          ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1),
                         strip.text.y = element_text(size = 10),
                         plot.title = element_text(size = 10),
                         plot.subtitle = element_text(size = 6),
                         legend.text = element_text(size = 5),
                         legend.title = element_text(size = 7)) +
          
          # Dynamic horizontal lines based on actual sample count
          ggplot2::geom_hline(yintercept = seq(0.5, n_samples + 0.5, 2),
                              color = "darkgrey") +
          
          # Titles and Facets
          ggplot2::labs(title = paste0(PRJ, " — Post-Cutadapt Residual Primers"),
                        subtitle = "Non-zero counts indicate incomplete primer removal.\nCheck Cutadapt flags and primer orientations.",
                        x = "Primers",
                        y = "Sample") +
          ggplot2::facet_grid(rows = vars(Primer_lbl),
                              cols = vars(Sequences_lbl),
                              scales = "free",
                              space = "free")
        
        # 4. Save the plot
        ggplot2::ggsave(
          file = paste0(results_path, "/", PRJ, "-residual_primers_post_cutadapt.pdf"),
          plot = primers_tile_residual,
          device= "pdf",
          width = 10 + (n_primers * 10),
          height = (n_samples * 0.3) + 3,
          units = "cm",
          limitsize = FALSE,
          dpi = 600
          )
        }
      } else {
        message("Primer removal successful: no residual primer hits detected across all samples.")
        }
    ```

#### Quality checking post Cutadapt

With primers removed by Cutadapt, it is possible to assess the intrinsic quality of the reads before defining the filtering parameters. This step runs FastQC and MultiQC on the trimmed output files, generating interactive reports per project. The resulting quality profiles — particularly per-position quality scores and read length distributions — should directly inform the truncLen and maxEE values used in filterAndTrim in the next step.

-   **Check that trimmed files exist.** Confirms that Cutadapt output files exist before running FastQC/MultiQC.

    ```{r}
    read_files <- cut_files_per_primer %>%
      dplyr::filter(stringr::str_detect(Stage, "_cutadapt")) %>%
      dplyr::pull(`Read file`) %>%
      normalizePath() %>%
      unique()

    read_files

    if (length(read_files) > 0) {
      
      message(length(read_files), " read files found for QC.")
      
      } else {

        stop("No read files found for QC. Check if Cutadapt ran successfully and output files were written.")
      }
    ```

-   **Run FastQC and MultiQC by project.** Creates per‑project folders and runs quality control in batch using system calls.

    ```{r}
    for (PRJ in all_projects) {
      # Create output directories
      fastqc_path  <- file.path(qual_path, PRJ, "fastqc")
      multiqc_path <- file.path(qual_path, PRJ, "multiqc")
      
      dir.create(fastqc_path,  recursive = TRUE, showWarnings = FALSE)
      dir.create(multiqc_path, recursive = TRUE, showWarnings = FALSE)
      
      # Filter reads belonging to this project
      cutadapted_path <- 
    cut_files_per_primer$`Read file`[cut_files_per_primer$Project == PRJ] %>%
    unique()
      
      project_reads <- 
    read_files[stringr::str_detect(read_files, paste0(cutadapted_path, collapse = "|"))]
      
      # 1. Run FastQC (requires GNU parallel + FastQC installed)
        cmd_fastqc <- paste0(
          "echo '", paste(project_reads, collapse = "\n"), "' | ",
          "parallel fastqc {1} -t 20 -o ", fastqc_path)
        system(cmd_fastqc)
        
      # 2. Run MultiQC
            report_name <- paste0(PRJ, "--sequencing_quality-MultiQC_report.html")
             
            cmd_multiqc <- paste0(
              "multiqc --interactive ",
              fastqc_path, " --filename ",
              file.path(multiqc_path, report_name))
            system(cmd_multiqc)
            
      # 3. Show MultQC report
            browseURL(file.path(multiqc_path, report_name))
    }
    ```

#### Filter and Trim

With primers removed by Cutadapt, reads now reflect only the biological sequence of the target amplicon and are ready for quality-based filtering. Unlike the pre-filtering step — which targeted only ambiguous bases — `filterAndTrim` is now applied with parameters that directly evaluate read quality, length, and integrity before ASV inference.

-   **Prepare file paths.** Map Cutadapt output files to new filtered output paths.

    ```{r}
    # Map Cutadapt output to filtered file paths
    filter_files_per_primer <- cut_files_per_primer %>%
      dplyr::filter(stringr::str_detect(string = Stage, pattern = "_cutadapt")) %>%
      dplyr::filter(file.exists(`Read file`)) %>%
      tidyr::pivot_wider(id_cols = c("Unique_File_name", "Unique_File_name_Primer",
                                 "Primer", "Run"),
                     names_from  = "Stage",
                     values_from = "Read file") %>%
      dplyr::mutate(
    FWD_R1_filtered = stringr::str_replace_all(FWD_R1_cutadapt, "cutadapt", "filtered"),
    FWD_R2_filtered = stringr::str_replace_all(FWD_R2_cutadapt, "cutadapt", "filtered")
      ) %>%
      tidyr::unite(col    = "Unique_File_name_Primer",
               sep    = "--",
               remove = FALSE,
               Unique_File_name, Primer)
    ```

-   **Define filtering function.** Create a wrapper for `filterAndTrim` applying quality thresholds.

    The key parameters to adjust for your dataset are:

    -   `maxEE`: Maximum number of expected errors tolerated per read. Lower values (e.g., c(2, 2)) are more stringent and recommended as a starting point. If read retention drops below \~70%, consider relaxing to c(2, 5) for the reverse read, which typically shows steeper quality decay.

    -   `truncQ`: Truncates the read at the first base with a Phred quality score ≤ this value. The default of 2 effectively removes reads with untrustworthy terminal bases.

    -   `minLen`: Discards reads shorter than this threshold after trimming. Set according to the minimum expected amplicon length for your marker (e.g., 100 for MiFish/NeoFish 12S amplicons).

    -   `truncLen`: Truncates all reads to a fixed length. Useful for markers with predictable amplicon sizes when quality degrades before the end of the read. Leave unset (or set to 0) for variable-length amplicons. Inspect `dada2::plotQualityProfile()` on your Cutadapt output files to decide.

    -   `rm.phix`: Removes reads matching the PhiX bacteriophage genome, a standard Illumina sequencing control that can contaminate libraries.

    ```{r}
    quality_filter <- function(Unique_File_name_Primer,
                               FWD_R1_in,
                               FWD_R2_in,
                               FWD_R1_out,
                               FWD_R2_out) {
      
      if (file.exists(FWD_R1_in) && file.exists(FWD_R2_in)) {
        
        sample_filtered_out <- dada2::filterAndTrim(fwd = FWD_R1_in,
                                                    filt = FWD_R1_out,
                                                    rev = FWD_R2_in,
                                                    filt.rev = FWD_R2_out,
                                                    maxN = 0,
      # Vector of length 2: first value applies to R1, second to R2.
                                                    maxEE = c(2, 2),
                                                    truncQ = 2,
                                                    minLen = 100,
                                                    rm.phix = TRUE,
                                                    # truncLen = c(0, 0),
                                                    multithread = TRUE,
                                                    compress    = TRUE,
                                                    verbose     = FALSE)
        
        tibble::as_tibble(sample_filtered_out) %>%
          
          dplyr::mutate(Unique_File_name_Primer = Unique_File_name_Primer)
        } else {
          
          warning(paste("Missing files for:", Unique_File_name_Primer))
          
          return(tibble::tibble())
        }
      }
    ```

-   **Execute filtering.** Run the filtering function in parallel across all samples and generate a summary table of read retention.

    ```{r}
    # Setup parallel processing
    cores_to_be_used <- future::availableCores() - 2
    future::plan(future::multisession(workers = cores_to_be_used))

    # Run filtering in parallel
    {
      tictoc::tic()
      
      all_filtered_out_fun <- 
        furrr::future_pmap_dfr(list(filter_files_per_primer$Unique_File_name_Primer,
                                    filter_files_per_primer$FWD_R1_cutadapt,
                                    filter_files_per_primer$FWD_R2_cutadapt,
                                    filter_files_per_primer$FWD_R1_filtered,
                                    filter_files_per_primer$FWD_R2_filtered),
                               .f = quality_filter,
                               .options = furrr::furrr_options(seed = TRUE))
      tictoc::toc()
    }

    # Generate summary table
    all_filtered_summary <- all_filtered_out_fun %>%
      dplyr::mutate(Proportion = round((reads.out / reads.in * 100), digits = 2)) %>%
      dplyr::rename("Primer-trimmed reads (pairs)" = "reads.in",
                "Filtered reads (pairs)"       = "reads.out")

    # View result
    all_filtered_summary
    ```

### Learn Error Rates and Infer ASVs

-   **Process by sequencing run.** DADA2 error models must be learned separately for each sequencing run. This step iterates through runs to learn errors, dereplicate reads, and infer ASVs.

    ```{r}
    # Initialize containers for DADA2 objects
    all_dadaFs <- list()
    all_dadaRs <- list()
    all_derep_forward <- list()
    all_derep_reverse <- list()

    # Loop through each sequencing run ----
    for (seq in unique(primers_n_samples$Run)) {
      
      message(paste0("Processing Run: ", seq))
      
      # 1. Get file lists for this run ----
      seq_files <- repair_files_per_primer %>%
    dplyr::filter(Run == seq) %>%
    dplyr::filter(file.exists(FWD_R1_repaired) & file.exists(FWD_R2_repaired))

      seq_fnFs <- seq_files$FWD_R1_repaired
      seq_fnRs <- seq_files$FWD_R2_repaired
      names(seq_fnFs) <- seq_files$Unique_File_name_Primer
      names(seq_fnRs) <- seq_files$Unique_File_name_Primer
      
      # 2. Learn Error Rates ----
      # Using nbases=1e9 (1 billion bases) is usually sufficient and faster than 1e16
      seq_errF <- dada2::learnErrors(
    fls = seq_fnFs, multithread = TRUE, randomize = TRUE, verbose = TRUE
      )
      seq_errR <- dada2::learnErrors(
    fls = seq_fnRs, multithread = TRUE, randomize = TRUE, verbose = TRUE
      )
      
      # Visualize errors (optional, saved to plot list if needed, here just printed)
      print(dada2::plotErrors(seq_errF, nominalQ = TRUE) + ggplot2::ggtitle(paste("Forward Errors -", seq)))
      print(dada2::plotErrors(seq_errR, nominalQ = TRUE) + ggplot2::ggtitle(paste("Reverse Errors -", seq)))

      # 3. Dereplication ----
      seq_derep_forward <- dada2::derepFastq(seq_fnFs, verbose = TRUE)
      seq_derep_reverse <- dada2::derepFastq(seq_fnRs, verbose = TRUE)
      
      # 4. Sample Inference (DADA2 Algorithm) ----
      seq_dadaFs <- dada2::dada(seq_derep_forward, err = seq_errF, multithread = TRUE)
      seq_dadaRs <- dada2::dada(seq_derep_reverse, err = seq_errR, multithread = TRUE)
      
      # Append to global lists
      # Note: lists are appended directly. If 'seq_dadaFs' is a list of objects, c() combines them flatly.
      all_dadaFs <- c(all_dadaFs, seq_dadaFs)
      all_dadaRs <- c(all_dadaRs, seq_dadaRs)
      all_derep_forward <- c(all_derep_forward, seq_derep_forward)
      all_derep_reverse <- c(all_derep_reverse, seq_derep_reverse)
    }
    ```

### Construct Sequence Table and Remove Chimeras

-   **Merge paired reads.** Combine forward and reverse reads to reconstruct the full amplicon.

    ```{r}
    # Merge read pairs
    all_mergers <- dada2::mergePairs(
      dadaF = all_dadaFs,
      derepF = all_derep_forward,
      dadaR = all_dadaRs,
      derepR = all_derep_reverse,
      minOverlap = 10,
      maxMismatch = 0,
      verbose = TRUE
    )

    # Construct Sequence Table (ASV table
    mergers_seqtab <- dada2::makeSequenceTable(all_mergers)

    # Check ASV length distribution
    table(nchar(dada2::getSequences(mergers_seqtab)))
    ```

-   **Remove chimeras.** Identify and remove artifactual chimeric sequences.

    ```{r}
    # Remove Chimeras
    mergers_seqtab_nochim <- dada2::removeBimeraDenovo(
      mergers_seqtab, 
      method = "consensus", 
      multithread = TRUE, 
      verbose = TRUE
    )

    # Calculate chimera removal stats
    chimera_prop <- sum(mergers_seqtab_nochim) / sum(mergers_seqtab)
    message(paste0("Retained ", round(chimera_prop * 100, 2), "% of reads after chimera removal."))

    # Inspect final length distribution
    table(nchar(dada2::getSequences(mergers_seqtab_nochim)))
    ```

### Track Read Retention

-   **Compile statistics.** Gather read counts from every processing step (Raw -\> Paired -\> Repaired -\> Denoised -\> Merged -\> Non-chimeric) to assess pipeline quality.

    ```{r}
    # Helper function to extract counts from DADA objects
    getN <- function(x) sum(dada2::getUniques(x))

    # Extract counts for DADA2 steps
    tbl_Denoised_R1 <- tibble::enframe(sapply(all_dadaFs, getN), name = "Unique_File_name_Primer", value = "Denoised (R1)")
    tbl_Denoised_R2 <- tibble::enframe(sapply(all_dadaRs, getN), name = "Unique_File_name_Primer", value = "Denoised (R2)")
    tbl_Merged <- tibble::enframe(rowSums(mergers_seqtab), name = "Unique_File_name_Primer", value = "Merged")
    tbl_NonChim <- tibble::enframe(rowSums(mergers_seqtab_nochim), name = "Unique_File_name_Primer", value = "Non-chimeric Merged")

    # Combine with previous stats
    # (Assumes 'all_filtered_out' and 'all_repaired' exist from previous chunks)

    # Prepare base table from Repaired stats
    all_track <- all_repaired %>%
      tidyr::separate(Unique_File_name_Primer, into = c("Unique_File_name", "Primer"), sep = "--", remove = FALSE) %>%
      dplyr::left_join(all_filtered_out, by = "Unique_File_name") %>%
      dplyr::left_join(tbl_Denoised_R1, by = "Unique_File_name_Primer") %>%
      dplyr::left_join(tbl_Denoised_R2, by = "Unique_File_name_Primer") %>%
      dplyr::left_join(tbl_Merged, by = "Unique_File_name_Primer") %>%
      dplyr::left_join(tbl_NonChim, by = "Unique_File_name_Primer") %>%
      dplyr::left_join(primers_n_samples, by = "Unique_File_name") # Add metadata

    # Save stats to Excel
    for (PRJ in all_projects) {
      prj_track <- all_track %>% dplyr::filter(Project == PRJ)
      writexl::write_xlsx(
    prj_track,
    path = file.path(results_path, paste0(PRJ, "-reads_and_seqs_counts-", Sys.Date(), ".xlsx"))
      )
    }

    # (Optional: Visualization code would follow here, using 'track_tbl' logic from original script)
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "3_DADA2_worflow-", Sys.Date()))
    ```

## 4. Phyloseq Integration (ou Downstream Analysis?)

### Generate Sample Metadata

-   **Format metadata table.** Prepare the sample metadata by organizing identifiers and ensuring row names match the sequence table, a requirement for Phyloseq.

    ```{r}
    # Create sample metadata dataframe for Phyloseq
    samdf <- primers_n_samples %>% 
      dplyr::select(
    Project, Sample, Researcher, Unique_File_name, Primer, Lib, Type,
    dplyr::starts_with("Metadata"),
    obs, `Extraction control`, `PCR control`, `Filtration control`
      ) %>% 
      BiocGenerics::unique() %>%
      # Expand rows if multiple primers exist per sample (semi-colon separated)
      tidyr::separate_longer_delim(cols = Primer, delim = ";") %>% 
      # Create unique identifier combining file name and primer
      tidyr::unite(
    col = "Unique_File_name_Primer", 
    Unique_File_name, Primer, 
    sep = "--", 
    remove = FALSE
      ) %>%
      as.data.frame()

    # Assign row names to match ASV table dimensions
    rownames(samdf) <- samdf$Unique_File_name_Primer

    # Optional: Preview metadata structure
    # str(samdf)

    # Check consistency between sequence table and metadata
    seq_names <- rownames(mergers_seqtab_nochim)
    meta_names <- rownames(samdf)

    if (!all(seq_names %in% meta_names)) {
      warning("Caution: Some samples in the sequence table are missing from metadata!")
      print(setdiff(seq_names, meta_names))
    } else {
      message("All samples matched correctly between Sequence Table and Metadata.")
    }
    ```

### Create Phyloseq Objects

-   **Import DADA2 results.** Combine the ASV table (no-chimera) and the formatted sample data into a Phyloseq object.

    ```{r}
    # Create Phyloseq object from DADA2 results
    # Note: 'mergers_seqtab_nochim' comes from the DADA2 workflow section
    mergers_ps <- phyloseq::phyloseq(
      phyloseq::otu_table(mergers_seqtab_nochim, taxa_are_rows = FALSE),
      phyloseq::sample_data(samdf)
      # phyloseq::tax_table(taxa_matrix) # Add taxonomy here if available later
    )

    # If concatenation was used (optional workflow branch), creating its object:
    # concat_ps <- phyloseq::phyloseq(
    #   phyloseq::otu_table(concat_seqtab_nochim, taxa_are_rows = FALSE),
    #   phyloseq::sample_data(samdf)
    # )
    ```

### Melt and Combine for Visualization

-   **Transform to tidy format.** Convert the Phyloseq object into a long-format tibble (psmelt), filtering out zero-abundance ASVs to optimize file size and plotting performance.

    ```{r}
    # Convert Phyloseq object to long tibble
    mergers_ps_tbl <- phyloseq::psmelt(mergers_ps) %>% 
      tibble::as_tibble() %>% 
      dplyr::mutate(`Read origin` = "merged") %>% 
      dplyr::filter(Abundance > 0) %>% # Filter strictly positive abundance
      dplyr::rename(ASV = OTU, ASV_tip = OTU) # Keep ASV sequence as identifier

    # (Optional) Process concatenated data if it exists
    # concat_ps_tbl <- ...

    # Combine all processed tables (if multiple sources exist) ----
    all_ps_tbl <- mergers_ps_tbl %>%
      # dplyr::bind_rows(concat_ps_tbl) %>% # Uncomment to join multiple sources
      dplyr::select(-Sample) %>%  # Remove duplicate/confusing columns if necessary
      dplyr::rename(Sample = sample_Sample) %>% # Standardize Sample column name
      dplyr::mutate(`ASV Size (bp)` = nchar(ASV))

    # View final tidy table
    # head(all_ps_tbl)
    ```

### Calculate Relative Abundances

-   **Compute relative abundances.** Calculate the proportion of each ASV relative to the total library size and the total dataset. This normalization is crucial for comparing samples with unequal sequencing depths.

    ```{r, echo=TRUE, eval=FALSE}
    # Calculate relative abundances per sample and project
    all_ps_tbl <- all_ps_tbl %>%
      # Calculate Total Abundance of the entire dataset once
      dplyr::mutate(Dataset_Total_Abundance = sum(Abundance)) %>%
      # Group by sample context to calculate sample-specific totals
      dplyr::group_by(Project, Unique_File_name, Primer, `Read origin`) %>%
      dplyr::mutate(
    Sample_Total_Abundance = sum(Abundance),
    Relative_Abundance_Dataset = round((Abundance / Dataset_Total_Abundance), digits = 6),
    Relative_Abundance_Sample  = round((Abundance / Sample_Total_Abundance), digits = 6)
      ) %>%
      dplyr::ungroup() %>%
      # Clean up temporary total column if desired, or keep for checks
      dplyr::select(-Dataset_Total_Abundance) %>%
      dplyr::relocate(Sample_Total_Abundance, Relative_Abundance_Dataset, Relative_Abundance_Sample, .after = Abundance)
    ```

### Check ASV Length Distribution

-   **Visualize ASV lengths.** Inspect the distribution of ASV lengths before taxonomic assignment. This helps identify potential artifacts or non-target amplicons.

    ```{r, echo=TRUE, eval=FALSE}
    # Plot ASV length distribution by sample
    # Define color palette if not already defined globally
    # viridis_pal <- viridis::turbo(n = 12)

    asv_length_plot <- all_ps_tbl %>%
      ggplot2::ggplot(aes(
    y = interaction(`Read origin`, Unique_File_name),
    x = `ASV Size (bp)`,
    fill = `Read origin`,
    col = `Read origin`,
    size = Relative_Abundance_Sample,
    alpha = 0.25
      )) +
      ggplot2::geom_jitter(height = 0.3, width = 0.3) +
      ggplot2::scale_x_continuous(
    breaks = seq(20, max(all_ps_tbl$`ASV Size (bp)`, na.rm = TRUE), 10),
    expand = c(0.02, 0.02)
      ) +
      ggplot2::scale_fill_manual(values = viridis::turbo(n = 12)[c(3, 6, 9, 12)]) +
      ggplot2::scale_colour_manual(values = viridis::turbo(n = 12)[c(3, 6, 9, 12)]) +
      ggplot2::labs(
    x = "ASV Size (bp)",
    y = "Sample",
    title = "ASV Length Distribution by Sample",
    subtitle = "Size and Read Origin per Sample (all identified ASVs)"
      ) +
      ggplot2::theme_bw(base_size = 8) +
      ggplot2::theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(size = 10, angle = 0)
      ) +
      ggplot2::geom_vline(xintercept = c(10, max(all_ps_tbl$`ASV Size (bp)`, na.rm = TRUE))) +
      ggplot2::facet_grid(
    rows = vars(Project, Primer, Type),
    scales = 'free_y', 
    space = 'free_y'
      )

    # Save plot
    # ggsave(...) # Add your ggsave logic here as needed
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "4_phyloseq-", Sys.Date()))
    ```

## 5. Taxonomic Assignment (BLASTn)

This section performs taxonomic identification of ASVs using local BLAST databases.

### Prepare and Execute BLASTn

-   **Select unique ASVs.** Extract unique sequences from the dataset for identification to avoid redundant BLAST searches.

    ```{r}
    # Prepare unique ASVs list
    asvs_blast_all <- all_ps_tbl %>%
      dplyr::pull(ASV_tip) %>% 
      unique() %>% 
      stats::na.omit() %>% 
      as.character()

    # Optional: Check for artifacts (poly-N runs)
    # which(stringr::str_detect(asvs_blast_all, "NNNNNNN"))
    ```

-   **Run Parallel BLAST.** Execute BLASTn in batches using the `BLASTr` package to handle large datasets efficiently without overloading memory.

    ```{r, echo=TRUE, eval=FALSE}
    # Define BLAST Parameters
    # Load BLASTr (ensure it is installed)
    # devtools::install_github("heronoh/BLASTr", force = TRUE)
    library(BLASTr)

    # Define chunk size for batch processing (e.g., 20,000 sequences per batch)
    chunk_size <- 20000
    asv_chunks <- split(asvs_blast_all, ceiling(seq_along(asvs_blast_all) / chunk_size))

    # Define databases path (adjust as needed)
    db_paths <- "/data/databases/core_nt/core_nt /data/databases/LGC12Sdb/mai25/LGC12Sdb_complete_noGaps-2025-05-21.fasta"

    # Execute BLASTn in Batches
    blast_results_list <- list()

    tictoc::tic("Total BLASTn taxonomy assignment")

    for (i in seq_along(asv_chunks)) {
      
      message(paste("Processing BLAST chunk", i, "of", length(asv_chunks)))
      
      chunk_res <- BLASTr::parallel_blast(db_path = db_paths,
                                          query_seqs = asv_chunks[[i]],
                                          out_file = file.path(blast_path, paste0("blast_out_res_", i, ".csv")),
                                          out_RDS  = file.path(blast_path, paste0("blast_out_res_", i, ".RDS")),
                                          total_cores = 60,       # Adjust based on server capacity
                                          num_threads = 2,
                                          blast_type = "blastn",
                                          perc_id = 80,
                                          perc_qcov_hsp = 80,
                                          num_alignments = 3
                                          )
      
      blast_results_list[[i]] <- chunk_res
      
      # Optional: Save intermediate environment
      # base::save.image(paste0(env_path, "/", "5_BLASTr-", Sys.Date())) 
    }

    tictoc::toc()

    # Combine Results
    blast_res_full <- dplyr::bind_rows(blast_results_list) %>% 
      dplyr::filter(!is.na(`1_subject header`)) %>% 
      dplyr::rename(ASV_tip = Sequence)

    # View summary
    # head(blast_res_full)
    ```

### Correct Database TaxIDs (LGC12Sdb)

-   **Map missing TaxIDs.** For custom databases like LGC12Sdb, fill in missing TaxIDs using an external mapping file.

    ```{r, echo=TRUE, eval=FALSE}
    # Load TaxID mapping file
    lgc_taxid_file <- "/data/databases/LGC12Sdb/oct24/LGC12Sdb_complete_noGaps-2024-09-20.txt"

    if (file.exists(lgc_taxid_file)) {
      
      LGC12db_tax_IDs <- read.csv(
    file = lgc_taxid_file,
    sep = " ",
    col.names = c("1_subject", "1_staxid"),
    header = FALSE,
    check.names = FALSE
      ) %>%
    tibble::as_tibble() %>%
    dplyr::mutate(`1_staxid` = as.character(`1_staxid`)) %>%
    unique()

      # Apply correction
      blast_res_full <- blast_res_full %>%
    # Convert N/A text to real NA
    dplyr::mutate(`1_staxid` = dplyr::if_else(`1_staxid` == "N/A", NA_character_, `1_staxid`)) %>%
    # Join with mapping file to find missing IDs
    dplyr::left_join(LGC12db_tax_IDs, by = "1_subject", suffix = c("", "_mapped")) %>%
    # Coalesce: keep original if exists, otherwise take mapped
    dplyr::mutate(`1_staxid` = dplyr::coalesce(`1_staxid`, `1_staxid_mapped`)) %>%
    dplyr::select(-`1_staxid_mapped`)

    } else {
      warning("LGC TaxID mapping file not found. Skipping TaxID correction.")
    }

    # Check final TaxID distribution
    # table(blast_res_full$`1_staxid`, useNA = "always")
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "5_Taxonomic_assignment-", Sys.Date()))
    ```

## 6. Taxonomy Curation & Assignment

This section refines the raw BLAST results, retrieves complete lineage information from NCBI, and standardizes taxonomic ranks.

### Refine BLAST Hits

-   **Filter low-quality hits.** Remove uninformative descriptions (e.g., "Uncultured") and select the best reliable hit among the top 3 alignments.

    ```{r, echo=TRUE, eval=FALSE}
    # Define patterns for non-informative hits
    bad_res_ids <- paste0(c(
      "Uncultured", "Uncultured organism", "Uncultured prokaryote",
      "Eukaryotic synthetic construct", "16S rRNA amplicon fragment",
      "Invertebrate environmental", "Uncultured Candidatus",
      "Uncultured bacterium", "Uncultured archaeon",
      "Complete Metagenome-Assembled", "PREDICTED: Nomascus"
    ), collapse = "|")

    # Vectorized selection of best hit (1st, 2nd, or 3rd)
    blast_res_processed <- blast_res_full %>%
      dplyr::mutate(
    # Check reliability of each hit
    is_bad_1 = stringr::str_detect(`1_subject header`, bad_res_ids),
    is_bad_2 = stringr::str_detect(`2_subject header`, bad_res_ids),
    is_bad_3 = stringr::str_detect(`3_subject header`, bad_res_ids),

    # Select best valid hit
    Selected_Hit_Origin = dplyr::case_when(
      !is_bad_1 ~ "1",
      is_bad_1 & !is.na(`2_subject header`) & !is_bad_2 ~ "2",
      is_bad_1 & is_bad_2 & !is.na(`3_subject header`) & !is_bad_3 ~ "3",
      TRUE ~ "None"
    ),

    # Assign Final ID and TaxID based on selection
    `blast ID` = dplyr::case_when(
      Selected_Hit_Origin == "1" ~ substr(`1_subject header`, 1, 40),
      Selected_Hit_Origin == "2" ~ substr(`2_subject header`, 1, 40),
      Selected_Hit_Origin == "3" ~ substr(`3_subject header`, 1, 40),
      TRUE ~ "Match_not_reliable"
    ),

    query_taxID = dplyr::case_when(
      Selected_Hit_Origin == "1" ~ `1_staxid`,
      Selected_Hit_Origin == "2" ~ `2_staxid`,
      Selected_Hit_Origin == "3" ~ `3_staxid`,
      TRUE ~ NA_character_
    )
      ) %>%
      # Second pass clean-up for any lingering "Uncultured" that slipped through
      dplyr::mutate(
    `blast ID` = dplyr::if_else(
      stringr::str_detect(`blast ID`, "ncultur|nvironmental"),
      "Match_not_reliable",
      `blast ID`
    )
      )

    # Check results
    # table(blast_res_processed$Selected_Hit_Origin)
    ```

### Clean Species Names

-   **Standardize nomenclature.** Clean up species names by removing artifacts, tags (e.g., "cf.", "aff."), and reducing names to Binomial format (Genus species).

    ```{r, echo=TRUE, eval=FALSE}
    # Clean strings
    blast_res_processed <- blast_res_processed %>%
      dplyr::mutate(
    `blast ID` = `blast ID` %>%
      stringr::str_replace(": ", ":") %>%
      stringr::str_remove("\\n$") %>%
      stringr::str_remove("^(  | |Uncultured |uncultured |Candidatus |MAG:|TPA_asm:|^Cf. |candidate division )") %>%
      stringr::str_remove_all("\\[|\\]") %>%
      stringr::str_replace("cf\\. ", "") %>%
      stringr::str_replace("nr\\. ", "") %>%
      stringr::str_replace(",", "") %>%
      # Extract only the first two words (Genus species)
      stringr::word(1, 2) %>%
      # Handle specific manual corrections
      dplyr::case_match(
        "Human DNA" ~ "Homo sapiens",
        "Human chromosome" ~ "Homo sapiens",
        .default = .
      )
      )

    # Verify cleaned names
    # head(unique(blast_res_processed$`blast ID`))
    ```

### Retrieve Lineage from NCBI

-   **Fetch complete taxonomy.** Query NCBI using the valid TaxIDs to retrieve full lineage information (Kingdom to Species).

    ```{r, echo=TRUE, eval=FALSE}
    # Identify unique TaxIDs to query
    taxids_to_search <- unique(na.omit(blast_res_processed$query_taxID))

    # Execute parallel taxonomy retrieval
    # Note: Ensure BLASTr is loaded. Using 'parallel_get_tax' from the package.
    # devtools::load_all("~/prjcts/BLASTr") # Only if local dev version needed

    taxonomy_tbl <- BLASTr::parallel_get_tax(
      organisms_taxIDs = taxids_to_search,
      retry_times = 5,
      verbose = "full"
    ) %>% 
      BiocGenerics::unique()

    # Check for missing IDs (failed queries)
    missing_taxids <- setdiff(taxids_to_search, taxonomy_tbl$query_taxID)
    if (length(missing_taxids) > 0) {
      message(paste(length(missing_taxids), "TaxIDs could not be retrieved."))
    }
    ```

### Fill Missing Ranks

-   **Impute missing levels.** Fill gaps in the taxonomic tree (e.g., missing Family) by propagating information from higher ranks with a prefix (e.g., "family of Order X").

    ```{r, echo=TRUE, eval=FALSE}
    # Function to fill missing ranks iteratively
    fill_rank <- function(df, target_rank, source_rank, prefix) {
      df[[target_rank]] <- ifelse(
    is.na(df[[target_rank]]) | df[[target_rank]] == "",
    paste0(prefix, df[[source_rank]]),
    df[[target_rank]]
      )
      return(df)
    }

    # Apply filling logic
    taxonomy_tbl_filled <- taxonomy_tbl %>%
      # Manual Superkingdom corrections based on Kingdom/Phylum
      dplyr::mutate(
    `Superkingdom (NCBI)` = dplyr::case_when(
      is.na(`Superkingdom (NCBI)`) & `Kingdom (NCBI)` %in% c("Metazoa", "Fungi", "Viridiplantae") ~ "Eukaryota",
      is.na(`Superkingdom (NCBI)`) & stringr::str_detect(`Phylum (NCBI)`, "bacteria$") ~ "Bacteria",
      TRUE ~ `Superkingdom (NCBI)`
    )
      ) %>%
      # Fill NAs sequentially (top-down)
      fill_rank("Superkingdom (NCBI)", "Kingdom (NCBI)", "superkingdom of ") %>%
      fill_rank("Kingdom (NCBI)", "Superkingdom (NCBI)", "kingdom of ") %>%
      fill_rank("Phylum (NCBI)", "Kingdom (NCBI)", "phylum of ") %>%
      fill_rank("Class (NCBI)", "Phylum (NCBI)", "class of ") %>%
      fill_rank("Order (NCBI)", "Class (NCBI)", "order of ") %>%
      fill_rank("Family (NCBI)", "Order (NCBI)", "family of ") %>%
      fill_rank("Genus (NCBI)", "Family (NCBI)", "genus of ")

    # Merge taxonomy back to BLAST results
    blast_res_tax <- dplyr::left_join(
      blast_res_processed,
      taxonomy_tbl_filled,
      by = "query_taxID"
    )

    # Join back to Main ASV Table
    all_ps_tbl_final <- dplyr::left_join(
      all_ps_tbl,
      blast_res_tax %>% dplyr::select(ASV_tip, `blast ID`, `Superkingdom (NCBI)`:`Species (NCBI)`),
      by = "ASV_tip"
    ) %>%
      dplyr::mutate(`Final ID` = `blast ID`) %>%
      dplyr::filter(Abundance > 0)
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "6_Taxonomic_curation-", Sys.Date()))
    ```

## 7. Contamination Analysis

This section identifies potential contaminant ASVs by comparing their abundance in samples against their abundance in the negative controls (PCR, extraction, filtration).

### Identify Contaminant ASVs

-   **Create a contaminant reference.** Isolate all ASVs present in control samples to build a reference table of potential contaminants and their maximum relative abundance within those controls.

    ```{r, echo=TRUE, eval=FALSE}
    # Define keywords for control sample types
    control_types <- c(
      "Extraction control", "Extraction Control", "Ext. control", "Ext. Control",
      "Filt. Control", "PCR control", "PCR Control", "Filt. control",
      "Filtration control", "Filtration Control", "Negative control",
      "Negative Control", "Positive control", "Positive Control", "Control"
    )

    # Create a lookup table of ASVs found in any control sample
    all_contam_asvs <- all_ps_tbl_final %>%
      dplyr::filter(Type %in% control_types) %>%
      dplyr::group_by(`ASV (Sequence)`, Unique_File_name) %>%
      # Find the max abundance of each ASV in each specific control well
      dplyrsummarise(`Max. ASV abd. in control` = max(`Relative abundance on sample`, na.rm = TRUE), .groups = "drop") %>%
      # Join back to get the control Type (PCR, Extraction, etc.)
      dplyr::left_join(
    all_ps_tbl_final %>% dplyr::select(Unique_File_name, Type, `Final ID`),
    by = "Unique_File_name"
      ) %>%
      dplyr::distinct()

    # Check the results
    # head(all_contam_asvs)
    # unique(all_contam_asvs$Unique_File_name)
    ```

### Flag Potential Contaminants in Samples

-   **Compare sample vs. control abundance.** For each ASV in a true sample, compare its abundance to its maximum observed abundance in the corresponding negative controls. ASVs that are not significantly more abundant in the sample than in the control are flagged as potential contaminants.

    ```{r, echo=TRUE, eval=FALSE}
    # This vectorized approach replaces the slow row-by-row 'for' loop.
    # It works by reshaping the data to link each sample-ASV pair to its controls.

    # 1. Prepare a long-format map of samples to their specific controls
    sample_to_control_map <- all_ps_tbl_final %>%
      dplyr::filter(!Type %in% control_types) %>%
      dplyr::select(
    Unique_File_name, `ASV (Sequence)`, `Relative abundance on sample`,
    `PCR control`, `Extraction control`, `Filtration control`
      ) %>%
      tidyr::pivot_longer(
    cols = c(`PCR control`, `Extraction control`, `Filtration control`),
    names_to = "Control_Type_Source",
    values_to = "Control_Name_Raw"
      ) %>%
      # Expand rows for semicolon-separated control names
      tidyr::separate_longer_delim(Control_Name_Raw, delim = ";")

    # 2. Create a clean lookup table for control abundances
    control_abundance_lookup <- all_contam_asvs %>%
      dplyr::select(
    Control_Name_Raw = Unique_File_name,
    `ASV (Sequence)`,
    `Max. ASV abd. in control`
      )

    # 3. Join the sample map with the control abundances
    contamination_check_tbl <- sample_to_control_map %>%
      dplyr::left_join(control_abundance_lookup, by = c("Control_Name_Raw", "ASV (Sequence)")) %>%
      # For each sample-ASV, find the highest contamination signal from any of its controls
      dplyr::group_by(Unique_File_name, `ASV (Sequence)`) %>%
      dplyr::summarise(
    max_control_abd = max(`Max. ASV abd. in control`, na.rm = TRUE),
    sample_abd = dplyr::first(`Relative abundance on sample`),
    .groups = "drop"
      ) %>%
      # Calculate fold change (handle cases where control abundance is 0 or NA)
      dplyr::mutate(
    fold_change = gtools::foldchange(sample_abd, max_control_abd),
    `Contamination status` = dplyr::if_else(
      fold_change > 1, # Rule: Must be more abundant in sample than control
      "True Detection",
      "Possible Contamination"
    )
      )

    # 4. Join contamination status back to the main table
    all_ps_tbl_final <- all_ps_tbl_final %>%
      dplyr::left_join(
    contamination_check_tbl %>% dplyr::select(`ASV (Sequence)`, Unique_File_name, `Contamination status`),
    by = c("ASV (Sequence)", "Unique_File_name")
      ) %>%
      # For controls themselves, status is irrelevant; for samples not in the check_tbl, they are True Detections
      dplyr::mutate(
    `Contamination status` = dplyr::coalesce(`Contamination status`, "True Detection")
      )

    # Check results
    # table(all_ps_tbl_final$`Contamination status`)
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "7_Contamination_analysis-", Sys.Date()))
    ```

## 8. Final Results Visualization

This section generates plots to visualize the distribution of all identified ASVs, their taxonomic assignment, size, and relative abundance across samples.

### Plot ASV Distributions by Project

-   **Generate overview plots.** For each project, create a comprehensive plot showing every ASV, colored by BLAST score and shaped by superkingdom. This allows for a high-level assessment of the taxonomic landscape and data quality.

    ```{r, echo=TRUE, eval=FALSE}
    # Define shapes for taxonomic groups
    SHAPES <- c("Archaea" = 25, "Bacteria" = 24, "Eukaryota" = 21, "Viruses" = 23)

    # Loop through each project to create and save plots
    for (PRJ in all_projects) {

      message(paste0("Plotting project: ", PRJ))

      # Prepare data for plotting (calculate pseudo-score)
      plot_data <- all_ps_tbl_final %>%
    dplyr::filter(Project == PRJ) %>%
    dplyr::mutate(
      `BLASTn pseudo-score` = ((`1_indentity` * (`1_qcovhsp` / 10)) / 10),
      # Nullify score for uncultured hits
      `BLASTn pseudo-score` = dplyr::case_when(
        stringr::str_detect(`1_subject header`, "ncult") ~ NA_real_,
        TRUE ~ `BLASTn pseudo-score`
      )
    ) %>%
    dplyr::arrange(dplyr::desc(`BLASTn pseudo-score`))

      # Build the plot
      asv_dist_plot <- ggplot2::ggplot(
    plot_data,
    aes(
      y = interaction(Sample, Unique_File_name, sep = " - "),
      x = `ASV Size (pb)`,
      fill = `BLASTn pseudo-score`,
      col = `BLASTn pseudo-score`,
      shape = `Superkingdom (NCBI)`,
      size = `Relative abundance on sample`,
      alpha = 0.2,
      text = paste0(
        "Final ID: ", `Final ID`, "<br>",
        "Sample: ", Sample, "<br>",
        "ASV Size (bp): ", `ASV Size (bp)`, "<br>",
        "Pseudo-score: ", round(`BLASTn pseudo-score`, 2)
      )
    )
      ) +
      ggplot2::geom_jitter(height = 0.4, width = 0.25) +
      ggplot2::scale_fill_gradientn(
    name = "BLASTn pseudo-score (%)", na.value = "grey95",
    colours = blast_score_palette,
    limits = c(60, 100), breaks = seq(60, 100, 5)
      ) +
      ggplot2::scale_color_gradientn(guide = "none", na.value = "grey80", colours = c("#330000", "darkred", "red", "yellow", "green", "darkgreen"), limits = c(60, 100)) +
      ggplot2::scale_size_continuous(name = "Relative Abundance (%)") +
      ggplot2::scale_shape_manual(name = "Superkingdom (NCBI)", values = SHAPES, na.value = 22) +
      ggplot2::labs(x = "ASV Size (bp)", y = "Sample", title = paste("EcoMol -", PRJ)) +
      ggplot2::theme_bw(base_size = 14) +
      ggplot2::theme(legend.position = "bottom", legend.key.width = unit(2, 'cm')) +
      ggplot2::facet_grid(rows = vars(`Read origin`, Primer), scales = "free_y", space = "free_y")

      # Save static plot
      n_samples_in_project <- length(unique(plot_data$Unique_File_name))
      ggsave(
    filename = file.path(results_path, paste0(PRJ, "-ASV_length_by_sample--All_ASVs.pdf")),
    plot = asv_dist_plot,
    device = "pdf",
    width = 40,
    height = round((n_samples_in_project * 0.5) + 5),
    units = "cm",
    limitsize = FALSE,
    dpi = 300
      )

      # Save interactive plot
      asv_dist_plotly <- plotly::ggplotly(asv_dist_plot, tooltip = "text")
      htmlwidgets::saveWidget(
    widget = asv_dist_plotly,
    selfcontained = TRUE,
    file = file.path(results_path, paste0(PRJ, "-ASV_length_by_sample--All_ASVs.html"))
      )
    }
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "8_Final_visualization-", Sys.Date()))
    ```

## 9. Final Results Curation

This section refines the identification confidence levels and categorizes organisms into functional groups (e.g., Metazoa) for downstream ecological analysis.

### Assign Identification Confidence

-   **Calculate confidence scores.** Compute a pseudo-score based on BLAST identity and coverage, and assign taxonomic ranks (Species, Genus, Family, etc.) based on standard percentage thresholds.

    ```{r, echo=TRUE, eval=FALSE}
    # Calculate scores and assign taxonomic resolution
    all_ps_tbl_final <- all_ps_tbl_final %>%
      dplyr::mutate(
    `Curated ID` = `Final ID`,
    # Weighted score: heavily weights identity (10x) over coverage
    `BLASTn pseudo-score` = ((10 * `1_indentity`) + `1_qcovhsp`) / 11,

    # Assign resolution level based on score thresholds
    Identification = dplyr::case_when(
      `BLASTn pseudo-score` >= 98 ~ `Final ID`, 
      `BLASTn pseudo-score` >= 95 ~ `Genus (NCBI)`, 
      `BLASTn pseudo-score` >= 90 ~ `Family (NCBI)`, 
      `BLASTn pseudo-score` >= 80 ~ `Order (NCBI)`, 
      `BLASTn pseudo-score` >= 60 ~ `Class (NCBI)`,
      TRUE ~ "Unidentified"
    ),

    `Identification Max. taxonomy` = dplyr::case_when(
      `BLASTn pseudo-score` >= 98 ~ "Species",
      `BLASTn pseudo-score` >= 95 ~ "Genus",
      `BLASTn pseudo-score` >= 90 ~ "Family",
      `BLASTn pseudo-score` >= 80 ~ "Order",
      `BLASTn pseudo-score` >= 60 ~ "Class",
      TRUE ~ "Unidentified"
    )
      )
    ```

### Mark Metazoan Taxa

-   **Categorize Metazoans.** Flag ASVs belonging to Metazoa or other target groups (Zooplankton, Phytoplankton, Benthos) based on their taxonomy. This facilitates filtering for specific ecological guilds.

    ```{r, echo=TRUE, eval=FALSE}
    # 1. Define Biological Groups Lists (Expert Curation)
    # Benthos
    bentos_phy <- c("Annelida", "Arthropoda", "Mollusca")
    bentos_cls <- c("Bivalvia", "Clitellata", "Gastropoda", "Insecta")

    # Zooplankton
    zplnk_phy <- c("Amoebozoa", "Arthropoda", "Cercozoa", "Ciliophora", "Rotifera")
    zplnk_cls <- c("Branchiopoda", "Eurotatoria", "Filosia", "Hexanauplia", "Lobosa", "Oligohymenophorea")

    # Periphyton
    prftn_phy <- c("Bacillariophyta", "Charophyta", "Chlorophyta", "Cryptophyta", "Cyanobacteria", "Euglenophyta", "Rhodophyta")
    prftn_cls <- c("Bacillariophyceae", "Chlorophyceae", "Chrysophyceae", "Coscinodiscophyceae", "Cryptophyceae", "Cyanophyceae", "Euglenophyceae", "Florideophyceae", "Mediophyceae", "Ulvophyceae", "Zygnematophyceae")

    # Phytoplankton (Comprehensive list)
    ftplnk_phy <- c(
      "Bacillariophyta", "Charophyta", "Chlorophyta", "Cryptophyta", "Cyanobacteria", "Dinophyta", "Euglenophyta", "Ochrophyta",
      "Apicomplexa", "Cercozoa", "Ciliophora", "Discosea", "Endomyxa", "Euglenozoa", "Evosea", "Foraminifera", "Fornicata",
      "Haptophyta", "Hemimastigophora", "Heterolobosea", "Nebulidia", "Nibbleridia", "Oomycota", "Parabasalia", "Perkinsozoa",
      "Preaxostyla", "Rhodophyta", "Tubulinea"
    )
    ftplnk_cls <- c(
      "Bacillariophyceae", "Chlorophyceae", "Chrysophyceae", "Coscinodiscophyceae", "Cryptophyceae", "Cyanophyceae",
      "Dinophyceae", "Euglenophyceae", "Mediophyceae", "Trebouxiophyceae", "Ulvophyceae", "Zygnematophyceae",
      "Actinophryidae", "Aphelidea", "Bigyra", "Bolidophyceae", "Breviatea", "Centroplasthelida", "Choanoflagellata",
      "Developea", "Dictyochophyceae", "Eustigmatophyceae", "Filasterea", "Glaucocystophyceae", "Hyphochytriomycetes",
      "Ichthyosporea", "Olisthodiscophyceae", "Pelagophyceae", "Phaeophyceae", "Phaeothamniophyceae", "Raphidophyceae",
      "Synurophyceae", "Xanthophyceae"
    )

    # Combine all target classes and phyla for quick lookup
    all_target_phy <- unique(c(bentos_phy, zplnk_phy, prftn_phy, ftplnk_phy))
    all_target_cls <- unique(c(bentos_cls, zplnk_cls, prftn_cls, ftplnk_cls))

    # 2. Apply Logic to Mark Metazoa/Targets
    all_ps_tbl_final <- all_ps_tbl_final %>%
      dplyr::mutate(
    `Possible Metazoa` = dplyr::case_when(
      # Exclusions (Hard filters)
      stringr::str_detect(`blast ID`, "Chordate environmental") ~ FALSE,
      stringr::str_detect(`blast ID`, "Match_not_reliable") ~ FALSE,
      stringr::str_detect(`Superkingdom (NCBI)`, "Bacteria|Archaea|Viruses") ~ FALSE, # Note: Viruses added
      is.na(`Kingdom (NCBI)`) ~ FALSE,
      
      # Inclusions (Target Groups)
      stringr::str_detect(`Superkingdom (NCBI)`, "Eukaryota") & (`Phylum (NCBI)` %in% all_target_phy) ~ TRUE,
      stringr::str_detect(`Superkingdom (NCBI)`, "Eukaryota") & (`Class (NCBI)` %in% all_target_cls) ~ TRUE,
      
      # General Metazoa/Viridiplantae catch-all (if not excluded above)
      stringr::str_detect(`Kingdom (NCBI)`, "Metazoa|Viridiplantae") ~ TRUE,
      
      # Default
      TRUE ~ FALSE
    )
      )

    # Check distribution
    # table(all_ps_tbl_final$`Possible Metazoa`, useNA = "always")
    ```

### Final Table Formatting

-   **Reorganize columns.** Select and rename columns to produce a clean, user-friendly output table for reporting.

    ```{r, echo=TRUE, eval=FALSE}
    # Create final clean table
    results_table_clean <- all_ps_tbl_final %>%
      # Rename control columns for clarity
      dplyr::rename(
    `PCR Control` = `PCR control`,
    `Ext. Control` = `Extraction control`,
    `Filt. Control` = `Filtration control`
      ) %>%
      # Select and order essential columns
      dplyr::select(
    Researcher, Project, Primer, Sample, Unique_File_name,
    `Read origin`, `Relative abundance on sample`, Abundance,
    `BLASTn pseudo-score`, Identification, `Curated ID`, `Possible Metazoa`,
    `Superkingdom (NCBI)`, `Phylum (NCBI)`, `Class (NCBI)`, `Order (NCBI)`, 
    `Family (NCBI)`, `Genus (NCBI)`, `Species (NCBI)`,
    `ASV (Sequence)`, `ASV Size (pb)`,
    `Contamination status`,
    starts_with("Metadata")
      )

    # Save final results
    writexl::write_xlsx(
      results_table_clean,
      path = file.path(results_path, paste0("Final_ASV_Table_", Sys.Date(), ".xlsx"))
    )
    ```

### Save the Environment!

-   **Save the environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "9_Final_results-", Sys.Date()))
    ```
