---
title: "Mock Communities ASV analyses"
author: "Gabriel A. Mendes", "Heron OH"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Mock Communities ASV versus MOTUs project - ASV analyses

## 1. Setup & Configuration

### Packages

-   **Load libraries.** Import all necessary packages for data manipulation, demultiplex, ASVs inference, taxonomy assigment and data visualization.

    ```{r}
    library(tidyverse)
    library(phyloseq)
    library(Biostrings)
    library(ShortRead)
    library(dada2)
    library(DECIPHER)
    library(future)
    library(viridis)
    library(ggh4x)
    library(furrr)
    library(writexl)
    library(tictoc)
    library(BLASTr)
    ```

### Directory Management

-   **Create directory structure.** Define project paths and create the directory structure required by the pipeline.

    ```{r}
    # Project folder path  
    prjct_path <- "/home/gabriel/projetos/mock_pipelines"

    # Derived paths
    {
      env_path <- file.path(prjct_path, "environments")
      data_path <- file.path(prjct_path, "data")
      pipe_libs <- file.path(data_path, "reads")
      raw_libs <- file.path(pipe_libs, "raw")
      paired_libs <- file.path(pipe_libs, "paired")
      cutadapt_libs <- file.path(pipe_libs, "cutadapt")
      qual_path <- file.path(prjct_path, "quality")
      results_path <- file.path(prjct_path, "results")
      figs_path <- file.path(prjct_path, "figures")
      blast_path <- file.path(results_path, "blast")
      swarm_path <- file.path(results_path, "swarm")
      dmx_path <- file.path(pipe_libs, "dmx")
      dmx_all_path <- file.path(dmx_path, "all_dmx")
      dmx_combined_path <- file.path(dmx_path, "all_combined")
      nochim_path <- file.path(results_path, "non_chimera")
      tables_path <- file.path(results_path, "tables")
      track_path <- file.path(results_path, "track_plots")
      error_path <- file.path(results_path, "error_plot")
    }

    # Paths to be created
    paths <- c(prjct_path, env_path, data_path, pipe_libs, raw_libs,
               paired_libs, cutadapt_libs, qual_path, results_path,
               figs_path, blast_path, swarm_path, dmx_path, dmx_all_path,
               dmx_combined_path, nochim_path, tables_path, track_path,
               error_path)

    for (dir in paths) {
      if (dir.exists(dir)) {
    print(paste("The folder", dir, "already exists!"))
      } else {
    print(paste("Creating", dir, "!" ))
    dir.create(dir) 
      }
    }

    # Files in the project folder
    list.files(prjct_path)
    ```

### Housekeeping

-   **Set Cutadapt path**

    ```{r}
    # Path to cutadapt executable
    cutadapt <- "/usr/local/bin/cutadapt"
    ```

-   **Define custom functions.**

    ```{r}
    # Convert IUPAC codes to regex. Enables searching degenerate primers directly on reads.
    iupac2seq <- function(dna_seq) {
      iupac_code <- c("A" = "A", "C" = "C", "G" = "G", "T" = "T",
                      "Y" = "[CT]", "R" = "[AG]", "H" = "[ACT]", "M" = "[AC]",
                      "K" = "[GT]", "S" = "[GC]", "W" = "[AT]", "B" = "[CGT]",
                      "D" = "[AGT]", "V" = "[ACG]", "N" = "[ACGT]")
      stringr::str_replace_all(dna_seq, iupac_code)
    }

    # Counts how many reads in a FASTQ file contain a given primer (allowing up to 1 mismatch)
    primerHits <- function(primer, file_name) {
      nhits <- Biostrings::vcountPattern(primer,
                                         ShortRead::sread(ShortRead::readFastq(file_name)),
                                         fixed        = FALSE,
                                         max.mismatch = 1)
      sum(nhits > 0)
      }

    # Generates all 4 orientations (FWD/REV/COMP/REVCOMP) for a single primer
        allOrients <- function(primers_list) {
      require(Biostrings)
      
      dna <- Biostrings::DNAString(primers_list$`Primer seq`)
      
      orients <- 
        c(Forward= dna,
          Complement = Biostrings::complement(dna),
          Reverse    = Biostrings::reverse(dna),
          RevComp    = Biostrings::reverseComplement(dna))
      
      primer_tbl <- sapply(orients, toString)
      
      primer_tbl <- 
        tibble::tibble(Sequence = primer_tbl,
                       `Primer orientation` = names(primer_tbl)) %>%
        dplyr::mutate(Primer = primers_list$`Primer name`) %>%
        tidyr::unite(col = "Orientation name",
                     Primer,
                     `Primer orientation`,
                     remove = FALSE) %>%
        dplyr::mutate(`Primer pair` = stringr::str_remove_all(Primer, "_FWD$|_REV$"),
                      `Parsed sequence` = iupac2seq(Sequence))
      
      base::names(primer_tbl$Sequence) <- primer_tbl$`Orientation name`
      
      primer_tbl
      }
    ```

-   **Load custom functions.** Custom functions for taxonomy retrievel.

    ```{r}
    # Demultiplex custom function
    source("/home/heron/prjcts/ecomol/R/dmx_metabar.R")
    ```

-   **Define Global Color Palettes.** Set consistent color schemes for plots.

    ```{r}
    # Color for each read situation 
    read_colors <- c("Raw reads (pairs)" = "#7A0403FF",
                     "Raw reads (R1)" = "#B11901FF",
                     "Raw reads (R2)" = "#B11901FF",
                     "Paired reads" = "#F25F14FF",
                     "Primer detected reads (pairs)" = "#FE942AFF",
                     "Primer detected reads (R1)" = "#F7C13AFF",
                     "Primer detected reads (R2)" = "#F7C13AFF",
                     "Repaired reads" = "#DAE336FF",
                     "Denoised (R1)" = "#AFFA37FF",
                     "Denoised (R2)" = "#AFFA37FF",
                     "Merged" ="#76FE5BFF",
                     "Non-chimeric Merged" ="#34F395FF",
                     "Concatenated" = "#2DB6F1FF",
                     "Non-chimeric Concatenated" = "#458AFCFF",
                     "R1 + R2" = "#4145AAFF",
                     "Non-chimeric R1 + R2" = "#3A2C78FF")

    # Plot identified primers
    primer_colours <- c("white", "red", "yellow",
                        "green", "dark green")

    # Plot identified ASVs
    colors <- c("#330000", "darkred", "red", 
                "yellow", "green", "darkgreen")

    # BLAST Score Gradient (Low to High quality)
    blast_score_palette <- c("#330000", "darkred", "red", 
                             "yellow", "green", "darkgreen")

    # Metazoa Status (True/False)
    metazoa_palette <- c("TRUE" = "steelblue", "FALSE" = "firebrick") 

    # Shapes for Superkingdoms
    taxa_shapes <- c("Archaea" = 25, "Bacteria" = 24, 
                     "Eukaryota" = 21, "Viruses" = 23)
    ```

### Load Samples table

This is the most important input on the analysis, along with the raw data. The *primer_n_samples Table* holds the information of the samples file names, primers, controls, indexes, different projects. **All names under the Unique_file_name colunm must be unique.**

After filling in the table, save it as .csv an place it on the data folder, inside the project's main directory.

-   **Load primers, indexes and samples table.** Reads the configuration CSV containing sample metadata, primers, indexes, and project assignments.

    ```{r}
    primers_n_samples <- 
      readr::read_csv(file = file.path(prjct_path, "data", "mock_primer_n_samples.csv"))

    # Show if the primer_n_samples table was loaded successfully
    primers_n_samples %>% 
      View()
    ```

-   **Check if file names are really unique.** Validates that Unique_File_name and Sample columns contain no duplicates, stopping execution if found.

    ```{r}
    if (nrow(primers_n_samples) != length(unique(primers_n_samples$Unique_File_name))) {
      message("Your file names are not unique")
      message(paste0("The file names:"))
      message(paste0(primers_n_samples$Unique_File_name[duplicated(primers_n_samples$Unique_File_name)], collapse = "\n "))
      message(paste0("apear more than once"))
      } else if (nrow(primers_n_samples) != length(unique(primers_n_samples$Sample))) {
        message("Your Sample names are not unique")
        message(paste0("The sample names:"))
        cat(paste0(primers_n_samples$Sample[duplicated(primers_n_samples$Sample)], collapse = "\n "))
        message(paste0("apear more than once"))
        } else {
          message("All file names and sample names are unique!")
          }
    ```

-   **Defining all separate projects to analyze independently.** Extracts and lists all unique project names from the samples table.

    ```{r}
    all_projects <- BiocGenerics::unique(primers_n_samples$Project) %>%
      stringr::str_split(pattern = ";", simplify = F) %>%
      base::unlist() %>%
      BiocGenerics::unique()

    message(paste0("This analysis will by split into separate projects for:", 
           paste0(c("\t", all_projects), 
                  collapse = "\n\t"))) 
    ```

-   **Identify files names radicals and levels.** Displays raw Unique_File_name values for manual inspection and establishes factor levels for consistent plotting order.

    ```{r}
    # if you need to print sample names and sort them yourself ----
    primers_n_samples$Unique_File_name %>% 
      paste0(collapse = '",\n"') %>% cat()

    # organize sample names in the order they will be printed on plots and results tables ----
    # sample_levels <- c(primers_n_samples$Unique_File_name) %>% unique()
    sample_levels <- c(primers_n_samples$Sample) %>% unique()

    # Show sample names
    sample_levels
    ```

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "1_setup_n_configuration-", Sys.Date()))
    ```

## 2. Raw Data Pre-processing

The second phase of the pipeline focuses on the pre-processing and diagnostics of the raw sequencing reads. Before proceeding to ASV inference, it is critical to assess read quality (via FastQC/MultiQC) and validate primer presence and orientation. These diagnostic analyses ensure data integrity and inform the parameters required for the subsequent filtering and cleaning steps.

### Raw data aquisition

-   **Get reads from Base Space.** This part is only required if your read files is on Basespace. If the read files are already downloaded and demultiplexed, place them on the *\$raw_libs* folder and proceed to the next section.

    ```{bash}
    # Navigate to raw files folder 
    cd  /home/gabriel/projetos/mock_pipelines/data/reads/raw;

    # Download run from BaseSpace ----
    # a - autenticate to basespace (must have a basespace account and shared projects/runs)
    bs auth;

    # 2b - list your projects data ----
    bs list datasets;

    # 2c - download multiplexed read files
    for id in $(cat ids.txt); \ do bs download dataset --id $id -o data/reads/raw --extension=fastq.gz; done
    ```

### Demultiplexing (Ignored. See Mock Pipelines Notion file)

-   **Run demultiplexing function.** Iterates over libraries in the configuration table (`primers_n_samples`) and applies the custom `dmx_metabar` function to split samples using the index information.

-   This stage its not needed because the samples are within different libraries.

    ```{r}
    # Identify all unique libraries
    all_Libs <- unique(primers_n_samples$Lib)

    # Loop over libraries and run demultiplexing
    for (LIB in all_Libs) {
      
     # Filter metadata for current library
    primers_n_samples_lib <- primers_n_samples %>%
     dplyr::filter(Lib %in% c(LIB))
      
    # Define raw R1/R2 file paths
    raw_files <- primers_n_samples_lib %>%
      dplyr::pull(`Undemultiplexed path`) %>%
      BiocGenerics::unique() %>%
      list.files(full.names = T) %>%
      Sys.glob() %>%
      normalizePath()
      
    # Set project parameters
    out_path <- primers_n_samples_lib %>%
      dplyr::pull(`Raw data path`) %>%
      BiocGenerics::unique()

    project  <-primers_n_samples_lib$Project %>%
      BiocGenerics::unique()

    # Run demultiplexing
    dmx_metabar(undmx_file_R1 = raw_files[1], #raw R1
                undmx_file_R2 = raw_files[2], #raw R2
                out_path = out_path, # pasta onde vai ser escrito todo output
                project = project, #nome do projeto, entre aspas tipo "eDNA_bromelias" sem acento nem espaço
                Unique_File_name = primers_n_samples_lib$Unique_File_name,  #coluna da tabela de configuração
                primers = primers_n_samples_lib$Primer,                     #coluna da tabela de configuração
                samples_names = primers_n_samples_lib$Sample,
                idxs_FWD_seqs = primers_n_samples_lib$`Index sequence FWD`, #coluna da tabela de configuração
                idxs_FWD_names = primers_n_samples_lib$`Index name FWD`,    #coluna da tabela de configuração
                idxs_REV_seqs = primers_n_samples_lib$`Index sequence REV`, #coluna da tabela de configuração
                idxs_REV_names = primers_n_samples_lib$`Index name REV`,    #coluna da tabela de configuração
                query_size = 12   #tamanho da sequencia a ser considerada do index + primer
    )
    }
    ```

### Quality checking

The raw sequencing reads are assessed for overall quality before any processing steps. This step runs FastQC and MultiQC on the demultiplexed files, generating per-project interactive reports. The resulting quality profiles — particularly per-position quality scores, GC content, and read length distributions — provide the baseline needed to interpret subsequent primer detection and filtering results.

-   **Check that demultiplexed files exist.** Confirms that the paths in `primers_n_samples` actually contain FASTQ files before running FastQC/MultiQC.

    ```{r}
    read_files <- primers_n_samples %>%
      dplyr::mutate(Read_file_pattern = file.path(`Raw data path`,
                                                  paste0(Unique_File_name, "*"))) %>%
      dplyr::pull(Read_file_pattern) %>%
      Sys.glob() %>%
      normalizePath() %>%
      unique()

    read_files

    if (length(read_files) > 0) {
      message(length(read_files), " read files found for QC.")
      } else {
        stop("No read files found for QC. Check `Raw data path` and file names `primers_n_samples`.")
        }
    ```

-   **Run FastQC and MultiQC by project.** Creates per‑project folders and runs quality control in batch using system calls.

    ```{r}
    for (PRJ in all_projects) {
      # Create output directories
      fastqc_path  <- file.path(qual_path, PRJ, "fastqc")
      multiqc_path <- file.path(qual_path, PRJ, "multiqc")
      
      dir.create(fastqc_path,  recursive = TRUE, showWarnings = FALSE)
      dir.create(multiqc_path, recursive = TRUE, showWarnings = FALSE)
      
      # Filter reads belonging to this project
      raw_path <- primers_n_samples$`Raw data path`[primers_n_samples$Project == PRJ] %>%
        unique()
      
      project_reads <- read_files[stringr::str_detect(read_files, 
                                                      paste0(raw_path, 
                                                             collapse = "|"))]
      
      # 1. Run FastQC (requires GNU parallel + FastQC installed)
            cmd_fastqc <- paste0("echo '", 
                                 paste(project_reads, collapse = "\n"), 
                                 "' | ",
                                 "parallel fastqc {1} -t 20 -o ", 
                                 fastqc_path)
            system(cmd_fastqc)
            
      # 2. Run MultiQC
            report_name <- paste0(PRJ, "--sequencing_quality-MultiQC_report.html")
             
            cmd_multiqc <- paste0("multiqc --interactive ",
                                  fastqc_path, 
                                  " --filename ",
                                  file.path(multiqc_path,
                                            report_name))
            system(cmd_multiqc)
            
      # 3. Show MultQC report
            browseURL(file.path(multiqc_path, report_name))
    }
    ```

### Primer detection Analysis

The following workflow performs a critical diagnostic check on the raw sequencing reads before any actual sequence trimming or filtering takes place. In this section, the script intentionally searches for all defined primer sequences across all samples, evaluating four different sequence orientations (Forward, Complement, Reverse, and Reverse-Complement). This comprehensive "all-versus-all" screening serves three primary quality control purposes:

1.  It validates that the expected primers are successfully detected within their corresponding samples;

2.  It checks for potential laboratory cross-contamination or sequencing index-hopping by ensuring samples do not harbor unexpected primer sets;

3.  It diagnoses the library's structural geometry (e.g., standard vs. mixed orientation) and highlights the presence of read-through events in short amplicons. Ultimately, this step generates a detailed visual matrix, allowing users to verify the integrity, purity, and orientation of the sequencing run before deploying Cutadapt for the actual primer excision.

#### Identify paired-end read files

-   **List R1/R2 files.** Scans the raw data directory for FASTQ files matching common Illumina naming patterns for forward and reverse reads.

    ```{r}
    # Base directory of the read files (assumes all reads share the same root as the first file)
    data_dir <- dirname(read_files) %>% 
      unique()

    # List all forward (R1) FASTQ files using common naming patterns
    R1_pattern <- "_R1_001.fastq|_R1_001.fastq.gz|\\.1.fastq|\\.R1.fastq|R1.fq|R1.fastq|1.fq.gz|-R1.fastq.gz"

    all_fnFs <- list.files(data_dir, 
                           pattern = R1_pattern,
                           full.names = TRUE) %>%
      sort()

    # List all reverse (R2) FASTQ files using common naming patterns
    R2_pattern <- "_R2_001.fastq|_R2_001.fastq.gz|\\.2.fastq|\\.R2.fastq|R2.fq|R2.fastq|2.fq.gz|-R2.fastq.gz"

    all_fnRs <- list.files(data_dir, 
                           pattern = R2_pattern,
                           full.names = TRUE) %>%
      sort()

    # Final verification message
    n_R1 <- length(all_fnFs)
    n_R2 <- length(all_fnRs)

    if (n_R1 == n_R2 && n_R1 > 0) {
      message("✓ Paired: ", n_R1, " R1 files match exactly the ", n_R2, " R2 files.\n", 
              "\nVerify if all expected pairs are listed:\n", 
              "\nR1 files:\n", paste0("  ", basename(all_fnFs), collapse = "\n"),
              "\n",
              "\nR2 files:\n", paste0("  ", basename(all_fnRs), collapse = "\n"))
    } else {
      message("✗ MISMATCH: ", n_R1, " R1 files vs ", n_R2, " R2 files.",
              "\nR1 files:\n", paste0("  ", basename(all_fnFs), collapse = "\n"),
              "\n",
              "\nR2 files:\n", paste0("  ", basename(all_fnRs), collapse = "\n"),
              "\nCheck file names in `primers_n_samples` match actual FASTQ files on disk.")
    }
    ```

-   **Match files to samples.** Associates each sample in primers_n_samples with its corresponding R1/R2 files by matching directory path and Unique_File_name.

    ```{r}
    # Initialize sample table with metadata and empty columns for R1/R2 paths
    sample_idx_tbl_wide <- primers_n_samples %>%
      dplyr::select("Unique_File_name", "Lib", "Run", "Project",
                    "Researcher", "Sample", "Primer", "Type",
                    "Raw data path", dplyr::ends_with("ontrol")) %>%
      dplyr::mutate(FWD_R1 = NA_character_,
                    FWD_R2 = NA_character_)

    View(sample_idx_tbl_wide)

    # For each sample, find the corresponding R1 and R2 files by matching path + Unique_File_name
    for (i in seq_len(nrow(sample_idx_tbl_wide))) {
      
      pattern_i <- paste0(sample_idx_tbl_wide$`Raw data path`[i], "/",
                          sample_idx_tbl_wide$Unique_File_name[i], "_")
      
      # Match forward read (R1) file
      sample_idx_tbl_wide$FWD_R1[i] <- 
        all_fnFs[BiocGenerics::grep(pattern_i, all_fnFs)] %>%
        BiocGenerics::unique() %>% .[1]
      
      # Match reverse read (R2) file
      sample_idx_tbl_wide$FWD_R2[i] <- 
        all_fnRs[BiocGenerics::grep(pattern_i, all_fnRs)] %>%
        BiocGenerics::unique() %>% .[1]
      }

    # Show table
    View(sample_idx_tbl_wide)
    ```

-   **Build long-format file table.** Adds paths for future paired-output files and reshapes the table to long format, with one row per file per stage.

    ```{r}
        # Create paths for the future paired-output files and reshape to long format
    sample_idx_tbl <- 
      sample_idx_tbl_wide %>%
      dplyr::mutate(FWD_R1_paired = file.path(paired_libs,
                                              paste0(Unique_File_name, "--FWD_R1_paired.fastq.gz")),
                    FWD_R2_paired = file.path(paired_libs,
                                              paste0(Unique_File_name, "--FWD_R2_paired.fastq.gz"))) %>%
      tidyr::pivot_longer(
        cols = c(FWD_R1, FWD_R2, FWD_R1_paired, FWD_R2_paired),
        names_to  = "Stage",
        values_to = "Read file") %>%
      dplyr::mutate(Unique_File_name_Primer = Unique_File_name)

    # Show table
    View(sample_idx_tbl)
    ```

#### Primer database and orientations

-   **Primer database.** Primer names and sequences (with IUPAC codes where applicable). Comments are retained only where they document the biological identity of the primers.

    ```{r}
    # Define master vector of active primers (uncomment lines to activate specific primer pairs)
    all_primers <- 
      c(
        NeoFish_FWD = "CGCCGTCGCAAGCTTACCCT",
        NeoFish_REV = "AGTGACGGGCGGTGTGTGC",
        MiFish_FWD = "GTCGGTAAAACTCGTGCCAGC",
        MiFish_REV = "ACATAGTGGGGTATCTAATCCCAGTTTG"
        # MiBird__FWD = "GGGTTGGTAAATCTTGTGCCAGC",
        # MiBird__REV = "CATAGTGGGGTATCTAATCCCAGTTTG",
        # MiFish variants (12S, fishes)
        # MiFishEM__FWD  = "GTCGGTAAAACTCGTGCCAGC", 
        # MiFishEMs__FWD = "GTCGGTAAAACTCGTGCCAGC",
        # MiFish__FWD    = "AAACTCGTGCCAGC",       
        # MiFish2__REV   = "CATAGTGGGGTATCTAATCCCAGTTTG",
        # MiFishEMs__REV = "CATAGTGGGGTATCTAATCCCAGTT",
        # MiFishEM__REV  = "CATAGTGGGGTATCTAATCCCAGTTTG",
        # MiFish__REV    = "TCTAATCCCAGTTTG",
        # p12SV5__FWD = "TTAGATACCCCACTATGC",
        # p12SV5__REV = "TAGAACAGGCTCCTCTAG",
        # V12Su__FWD = "GTGCCAGCNRCCGCGGTYANAC",
        # V12Su__REV = "ATATRGGGTATCTAATCCYAGT",
        # V16Su__FWD = "ACGAGAAGACCCYRYGRARCTT",
        # V16Su__REV = "TCTHRRANAGGATTGCGCTGTTA",
        # VCOIu__FWD = "CAYGCHTTTGTGNATRATYYTYTT",
        # VCOIu__REV = "GGRGGRTADACDGTYCANCCNGT"
        # (Other primer pairs can be re‑added here following the same pattern)
        ) %>% 
      stats::na.omit()

    # Convert primer vector to list of single-row tibbles (for allOrients function)
    primers_list <- tibble::tibble(`Primer seq`  = all_primers,
                                   `Primer name` = names(all_primers)) %>%
      split(1:nrow(.))

    primers_list

    # Create reference table with IUPAC → Regex conversion for inspection
    all_primers_tbl <- tibble::tibble(
      Primer = names(all_primers),
      `IUPAC sequence` = all_primers) %>%
      dplyr::mutate(`Regex sequence` = iupac2seq(`IUPAC sequence`))

    # Show primers
    all_primers_tbl
    ```

-   **Generate all primer orientations** Produces all orientations (Forward/Reverse/Complement/RevComp.) and their regex representation for downstream searching.

    ```{r}
    # Apply all Orients to all primers and clean final table
    primers_all_orients <- purrr::map_dfr(primers_list, 
                                          allOrients) %>%
      dplyr::mutate(`Primer pair` = stringr::str_remove_all(Primer, "_FWD$|_REV$"))
    # Inspect primers orientations
    View(primers_all_orients)
    ```

#### Per-file primer hits

-   **Define primer counting function.** The function `multi_primerHits()` applies B. Callahan's `primerHits()` function across all primer orientations for a single FASTQ file, returning a table of hit counts per orientation. Defining it here keeps the primer-counting logic close to where it is used.

    ```{r}
    multi_primerHits <- function(Read_file, primers) {
      primer_counts <- purrr::map_df(primers,
                                     .f = primerHits,
                                     file_name = Read_file)
      primer_counts %>%
        dplyr::mutate(`Read file` = Read_file)
    }
    ```

-   **Per‑file primer hits.** Counts the number of reads that contain a given primer pattern (with up to 1 mismatch). All four orientations generated by `allOrients()` — Forward, Complement, Reverse, and RevComp — are searched simultaneously against each R1 and R2 file, enabling detection of primers regardless of read orientation or library preparation strategy. The resulting table is then joined with sample metadata and annotated with total read counts per file to allow proportional comparisons across samples.

    ```{r}
    # Select only raw reads (R1/R2) for primer detection
    reads_seqs_tbl <- sample_idx_tbl %>%
      dplyr::filter(Stage %in% c("FWD_R1", "FWD_R2")) %>%
      dplyr::select(`Read file`, Unique_File_name, Stage,
                    Project, Run, Primer)

    # Create named vector of FASTQ paths (names = sample identifiers)
    reads_seqs <- reads_seqs_tbl$`Read file`
    names(reads_seqs) <- reads_seqs_tbl$Unique_File_name

    # Extract all primer orientations as named vector for counting
    primers_seqs <- primers_all_orients %>%
      dplyr::pull(Sequence, name = `Orientation name`)

    # Set up parallel processing (all cores minus 2)
    cores_to_use <- future::availableCores() - 2
    future::plan(future::multisession(workers = cores_to_use))

    # Count primer occurrences across all files x all primer orientations (parallel)
    {
      tictoc::tic("Counting primers")
      
      message("Searching primers across ", length(reads_seqs), " files using ",
              cores_to_use, " cores. This may take a few minutes...")
      
      primers_in_Nreads <- furrr::future_map_dfr(reads_seqs,
                                                 .f = multi_primerHits,
                                                 primers = primers_seqs,
                                                 .options = 
                                                   furrr::furrr_options(seed = NULL))
      
      tictoc::toc()
    }

    # Join with sample metadata and calculate total reads per file
    primers_in_Nreads <- 
      primers_in_Nreads %>%
      dplyr::left_join(sample_idx_tbl, by = "Read file") %>%
      dplyr::mutate(Read = dplyr::if_else(stringr::str_detect(Stage, "R1"), "R1", "R2")) %>%
      tidyr::unite(col = "Unique_File_name_Read",
                   Unique_File_name,
                   Read,
                   sep = " ",
                   remove = FALSE) %>%
      dplyr::mutate(`Total reads` = ShortRead::countFastq(dirPath = `Read file`)[, 1])

    message("✓ Primer search complete. Results in `primers_in_Nreads`:\n",
            "  ", length(unique(primers_in_Nreads$Unique_File_name)), " samples × ",
            length(primers_seqs), " primer orientations across ",
            length(reads_seqs), " files.")
    ```

#### Primers identified in each sample

-   **Prepare data.** Transform the primer hits table into a long format suitable for visualization with `ggplot2`. This step filters out unobserved primers and organizes metadata.

    ```{r}
    # Prepare primer counts for plots in ggplot

    # Check available orientation names (for debugging)
    # primers_all_orients$`Orientation name`

    # Print column names formatted for selection (helper for coding)
    # primers_in_Nreads %>% colnames() %>% paste0(collapse = '",\n') %>% cat()

    # Convert primer hits table to long format
    primers_in_Nreads_long <- primers_in_Nreads %>% 
      dplyr::select(!dplyr::ends_with(c("_Complement", "_Reverse"))) %>% 
      tidyr::gather(key = Sequences,
                    value = Count,
                    ends_with("FWD_Forward"),
                    ends_with("FWD_RevComp"),
                    ends_with("REV_Forward"),
                    ends_with("REV_RevComp")) %>% 
      dplyr::mutate(Sequences = as.factor(Sequences),
                    Unique_File_name = factor(Unique_File_name, 
                                              levels = sample_levels)) %>% 
      dplyr::select(-c("Read file", "Stage")) %>% 
      # Remove columns that are entirely NA
      dplyr::select(where(function(x) any(!is.na(x)))) %>% 
      # Filter out primers with zero counts
      dplyr::filter(Count != 0)

    #View table
    View(primers_in_Nreads_long)
    ```

-   **Configure plot settings.** Load necessary visualization libraries and define the factor levels for sample ordering in the plots.

    ```{r}
    # Create levels for Files display on plot based on the prepared data
    Unique_File_name_Read_levels <- primers_in_Nreads_long %>% 
      dplyr::pull(Unique_File_name_Read) %>% 
      base::unique()
    ```

-   **Generate diagnostic plot.** Iterate through each project to create specific primer abundance plots. The loop filters data per project, generates a tile/bar plot showing primer presence and proportion, and saves the output as a PDF.

    ```{r}
    # Loop through all projects defined in the workflow
    for (PRJ in all_projects) {
      
      options(scipen = 10000)

      # 1. Filter and prepare data ONCE for the current project
      # This avoids filtering the large dataset multiple times inside the loop
      project_data <- primers_in_Nreads_long %>%
        dplyr::filter(Project == PRJ) %>%
        dplyr::mutate(
          
          # Pre-format labels and calculate percentages to keep ggplot clean
          Sequences_lbl = stringr::str_replace_all(Sequences, "__|_", "\n"),
          Primer_lbl    = stringr::str_replace_all(Primer, ";", "\n"),
          Pct_Reads     = (Count / `Total reads` * 100))
      
      # 2. Calculate dynamic dimensions based on actual data
      n_samples <- length(unique(project_data$Unique_File_name_Read))
      n_primers <- length(unique(project_data$Primer))

      # 3. Create the plot object
      primers_tile <- project_data %>%
        ggplot2::ggplot(aes(y = Unique_File_name_Read)) +
        # Bar layer
        ggplot2::geom_bar(aes(x = `Total reads`,
                              fill = Pct_Reads,
                              color = Primer,
                              group = Sequences),
                          stat = "identity",
                          position = "dodge",
                          linewidth = 0.05,
                          linetype = 2,
                          alpha = 0.5) +
        # Text labels for counts
        ggplot2::geom_text(aes(x = `Total reads`, 
                               label = Count),
                           hjust = 0, 
                           size = 1) +
        # Color scales
        ggplot2::scale_fill_gradientn(name = "Proportion of reads\n with primer (%)",
                                      colours = primer_colours,
                                      values = c(0, 1),
                                      na.value = "white") +
        # Axes and Guides
        ggplot2::scale_y_discrete(limits = rev) +
        ggplot2::scale_x_log10(
          breaks = c(10, 100, 1000, 10000, 100000),
          expand = expansion(mult = c(0, 0.2))) +
        ggplot2::scale_alpha(guide = 'none') +
        ggplot2::guides(color = guide_legend(override.aes = list(fill = "white", 
                                                                 size = 10))) +
        
        # Theme adjustments
        ggplot2::theme_light(base_line_size = 0.025, base_size = 6) +
        ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1),
                       strip.text.y = element_text(size = 10),
                       plot.title = element_text(size = 10),
                       plot.subtitle = element_text(size = 6),
                       legend.text = element_text(size = 5),
                       legend.title = element_text(size = 7)) +
        
        # Dynamic horizontal lines based on actual sample count
        ggplot2::geom_hline(yintercept = seq(0.5, n_samples + 0.5, 2), 
                            color = "darkgrey") +
        
        # Titles and Facets
        ggplot2::labs(title = PRJ,
                      subtitle = "Presença de primers nas amostras:\nintensidade de cor relativa à contagem do respectivo primer no conjunto de reads\n(max. mismatch = 1)",
                      x = "Primers",
                      y = "Amostra") +
        ggplot2::facet_grid(rows = vars(Primer_lbl),
                            cols = vars(Sequences_lbl),
                            scales = "free",
                            space = "free")

      # 4. Save the plot
      ggplot2::ggsave(file = paste0(track_path, "/", 
                                    PRJ, 
                                    "-primers_found_in_reads.pdf"),
                      plot = primers_tile,
                      device = "pdf",
                      width = 10 + (n_primers * 10),
                      height = (n_samples * 0.3) + 3,
                      units = "cm",
                      limitsize = FALSE,
                      dpi = 600)
      
      # 5. Open the plot 
      utils::browseURL(paste0(track_path, "/", PRJ, "-primers_found_in_reads.pdf"))
      }
    ```

-   **Embed PDF results.** Dynamically list and display the generated PDF plots within the HTML report.

    ```{r}
    # Extract the PDF path from the R object
    pdf_info <- list.files(path = results_path,
                           pattern = "primers_found_in_reads",
                           full.names = TRUE)

    # Generate the iframe HTML code
    if(length(pdf_info) > 0) {
      cat(paste0('<iframe src="', pdf_info, '" width="1200" height="800"></iframe>'))
      } else {
        cat("No primer plots found.")
        }
    ```

-   **Check orientations.** Analyze the generated plots to confirm that primers appear in the expected read files and orientations. The correct interpretation depends on two factors: **how the sequencing library was constructed** (standard vs. mixed orientation) and **whether read-through occurs** (i.e., whether the amplicon is shorter than the read length).

    **Understanding the four orientation columns.** The `allOrients()` function generates four variants of each primer sequence: `Forward` (original sequence), `Complement` (same direction, complementary bases), `Reverse` (original read 3'→5'), and `RevComp` (reverse complement). Only `Forward` and `RevComp` are biologically observable in Illumina sequencing — the polymerase always synthesizes 5'→3', so a primer can only appear at the start of a read in its `Forward` form, or at the end of a read in its `RevComp` form (due to read-through). Any counts in `Complement` or `Reverse` columns at `max.mismatch = 1` are near-chance artifacts of short sequence similarity and **must be treated as zero**.

    **Case 1 — Standard library, amplicon longer than the read length (no read-through).** In a standard Illumina paired-end library, R1 sequences the forward strand 5'→3' (starting from where the FWD primer bound), and R2 sequences the reverse strand 5'→3' (starting from where the REV primer bound). When the amplicon is long enough that neither read reaches the opposite primer, the expected result is:

    | Primer  | R1                 | R2                 |
    |:--------|:-------------------|:-------------------|
    | **FWD** | **Forward** (good) | \~0                |
    | **REV** | \~0                | **Forward** (good) |

    All other orientation/read combinations should be near zero. *Typical examples:* 16S V3–V4 (\~460 bp) with 2×250 bp reads; 12S V5 with 2×150 bp reads when the amplicon exceeds read length.

    **Case 2 — Standard library, amplicon shorter than the read length (read-through occurs).** When the amplicon is shorter than the read length, R1 reads through the entire insert and into the REV primer site, encountering it from the antisense direction — it therefore appears as `RevComp` at the 3' end of R1. Symmetrically, R2 reads through and encounters the FWD primer site as `RevComp` at its 3' end. The expected result is:

    | Primer  | R1                                       | R2                                       |
    |:-----------------|:--------------------------|:--------------------------|
    | **FWD** | **Forward** (good, 5' start of R1)       | **RevComp** (read-through, 3' end of R2) |
    | **REV** | **RevComp** (read-through, 3' end of R1) | **Forward** (good, 5' start of R2)       |

    The `RevComp` read-through signal is not an error — it is expected and **must** be handled by Cutadapt, which trims both the 5' primer and the 3' reverse-complement of the opposite primer from each read. The proportion of reads with read-through signal will vary with the length distribution of the target region: short amplicons will show near-equal counts in `Forward` and `RevComp`, while longer ones will show a weaker `RevComp` signal. *Typical examples:* ITS1/ITS2 fungal barcodes; short COI barcode fragments; short 16S V4 amplicons sequenced with 2×300 bp reads.

    **Case 3 — Mixed orientation library.** Some library preparation protocols (e.g., HiSeq/NovaSeq workflows without strand-selection, certain ligation-based preps, or pooled multiplexing strategies) produce reads where approximately half the read pairs have FWD on R1 and the other half have REV on R1. In this scenario, the plot will show roughly equal counts of both primers in both read files in `Forward` orientation:

    | Primer  | R1                           | R2                           |
    |:--------|:-----------------------------|:-----------------------------|
    | **FWD** | **Forward** (\~50% of reads) | **Forward** (\~50% of reads) |
    | **REV** | **Forward** (\~50% of reads) | **Forward** (\~50% of reads) |

    This is not an error — it reflects the library design. However, DADA2 assumes that all R1 reads begin with the FWD primer and all R2 reads begin with the REV primer; mixed-orientation input will produce catastrophic failures at the merging step. Before proceeding, reads must be reoriented using one of the following strategies:

    -   **`vsearch --orient`**: Reorients each read relative to a reference database before any further processing, producing a consistently stranded output. Can be applied to R1 and R2 independently before they enter the pipeline.

    -   **`dada2::removePrimers(..., orient = TRUE)`**: When using DADA2's own primer removal function, the `orient` parameter detects read pairs where the FWD primer is found at the start of R2 instead of R1, swaps the pair, and unifies orientation — no external tool required.

    -   **Dual Cutadapt strategy (recommended for maximum read retention)**: Run Cutadapt twice — once with FWD anchored on R1 and REV on R2 (`standard` pairs), and once with REV anchored on R1 and FWD on R2 (`flipped` pairs), using `--discard-untrimmed` on each run to keep only correctly matched pairs. Run the full DADA2 pipeline independently on each subset, then merge the two resulting ASV tables, reverse-complementing all sequences from the `flipped` run to unify orientation before merging. This is the approach recommended by B. Callahan for mixed-orientation data.

    **Troubleshooting: diagnosing orientation anomalies.**

    | Pattern observed in the plot                                                                 | Likely cause                                                                                                | Recommended fix                                                                                                                                               |
    |:------------------|:---------------------|:------------------------------|
    | FWD in R2/**Forward** (strong); REV in R1/**Forward** (strong)                               | R1/R2 file paths are swapped, or FWD/REV primer definitions are swapped in `primersnsamples`                | Swap R1↔R2 path assignments in `sampleidxtbl`, or swap FWD and REV primer entries                                                                             |
    | FWD in R1/**RevComp** (strong); REV in R2/**RevComp** (strong); near-zero **Forward** signal | Both primer sequences were entered in reverse-complement direction (e.g., copied from the antisense strand) | Replace both: `FWD <-     FWD.orient[["RevComp"]]` and `REV <- REV.orient[["RevComp"]]`                                                                       |
    | REV in R2/**RevComp** instead of R2/**Forward** (FWD signal looks correct)                   | REV primer sequence was entered as its own reverse complement                                               | Replace: `REV <-     REV.orient[["RevComp"]]` before proceeding \[\^1\]                                                                                       |
    | FWD and REV in both R1/**Forward** and R2/**Forward** in roughly equal proportions           | Mixed orientation library                                                                                   | Use `vsearch --orient`, `removePrimers(orient = TRUE)`, or the dual Cutadapt strategy described above \[\^2\]                                                 |
    | No primer detected in any orientation in any file                                            | Wrong primer pair selected; reads are already pre-trimmed; mismatches \> 1; incorrect file paths            | Verify primer sequences against the wet-lab protocol; confirm reads are not pre-trimmed (check FastQC/MultiQC report); increase `max.mismatch` to 2 as a test |

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "2_raw_data_pre_processing-", Sys.Date()))
    ```

## 3. DADA2 ASV Workflow

This section executes the DADA2 core workflow to infer Amplicon Sequence Variants (ASVs) from the pre-processed reads. The workflow is organized into two phases:

1.  **Read preparation**: ambiguous bases are removed, primers are excised with Cutadapt, their removal is verified, and a final quality filter is applied.

2.  **DADA2 inference steps**: a run-specific error model is learned, reads are denoised and merged, and chimeric sequences are removed, resulting in a high-resolution ASV table analogous to a traditional OTU table.

### Read Preparation and Primer Removal

Before DADA2 can learn error rates and infer ASVs, reads must be free of ambiguous bases, primer sequences, and low-quality bases. This preparation is carried out in four sequential steps: a pre-filter to remove reads containing Ns, primer removal with Cutadapt, a sanity check confirming successful trimming, and a final quality filter applied to primer-free reads.

#### Pre-filtering reads

DADA2 requires sequences to be completely free of ambiguous nucleotides (Ns) to accurately perform its core sample inference. Furthermore, the presence of Ns can interfere with accurate primer recognition and mapping during the Cutadapt step. Following the official DADA2 ITS workflow, this pre-filtering step is executed strictly to discard read pairs containing any Ns (**`maxN = 0`**). Comprehensive quality filtering, error tracking (such as **`maxEE`**), and length truncation are purposefully deferred until after primer removal.

-   **Prepare file paths.** Identify paired-end read files that contain primers and organize their paths for the filtering step.

    ```{r}
    # Select files with identified primers and reshape to wide format for processing
    primer_paired_files <- sample_idx_tbl %>%
      dplyr::filter(stringr::str_detect(string = Stage,
                                        pattern =
                                          "FWD_R1|FWD_R2|FWD_R1_paired|FWD_R2_paired"
                                        )) %>% 
      tidyr::pivot_wider(id_cols = c("Unique_File_name", "Primer", "Lib"),
                         names_from = "Stage",
                         values_from = "Read file")

    # Show plot
    View(primer_paired_files)

    # Verify basenames (optional check)
    # primer_paired_files$FWD_R1 %>% BiocGenerics::basename()
    # primer_paired_files$FWD_R1_paired %>% basename()
    ```

-   **Define pre-filtering function.** Create a wrapper function for filterAndTrim to handle forward and reverse reads and removing ambiguous bases (Ns).

    ```{r}
    # Custom function to repair and filter reads using DADA2
    pre_filter_Ns <- function(Unique_File_name,
                              FWD_R1_in,
                              FWD_R2_in,
                              FWD_R1_out,
                              FWD_R2_out) {
      # Check if input files exist
      if (file.exists(FWD_R1_in) && file.exists(FWD_R2_in)) {
        
        total_files = length(FWD_R1_in) + length(FWD_R2_in)
        
        # Run filterAndTrim with ONLY maxN = 0 to prepare for Cutadapt
        sample_filtered_out <- dada2::filterAndTrim(fwd = FWD_R1_in,
                                                    filt = FWD_R1_out,
                                                    rev = FWD_R2_in,
                                                    filt.rev = FWD_R2_out,
                                                    maxN = 0,
                                                    multithread = TRUE,
                                                    compress = TRUE,
                                                    verbose = TRUE)
        
        # Format output
        tibble::as_tibble(sample_filtered_out) %>%
          dplyr::mutate(Unique_File_name = Unique_File_name)
        
        } else {
          warning(paste("Missing files for:", Unique_File_name))
          return(tibble::tibble())
        }
      }
    ```

-   **Execute filtering.** Run the pre-filtering function in parallel across all samples and generate a summary table of read retention.

    ```{r}
    # Setup parallel processing
    cores_to_use <- future::availableCores() - 2 
    future::plan(future::multisession(workers = cores_to_use))

    # How much files 
    files_to_filt <- c(primer_paired_files$FWD_R1,
                       primer_paired_files$FWD_R2)

    # Run pre-filtering in parallel
    {
      tictoc::tic("Pre-filtering samples")
      
      message("Removing ambigous bases across ", length(files_to_filt), " files using ", 
              cores_to_use, " cores. This may take a few minutes...")
      
      pre_filtered_out <-
        furrr::future_pmap_dfr(list(primer_paired_files$Unique_File_name,
                                    primer_paired_files$FWD_R1,
                                    primer_paired_files$FWD_R2,
                                    primer_paired_files$FWD_R1_paired,
                                    primer_paired_files$FWD_R2_paired),
                               .f = pre_filter_Ns,
                               .options = furrr::furrr_options(seed = TRUE))
      
      tictoc::toc()
    }

    # Generate summary table
    pre_filtered_summary <- pre_filtered_out %>%
      dplyr::mutate(Proportion = round((reads.out / reads.in * 100), 
                                       digits = 2)) %>% 
      dplyr::rename("Raw reads (pairs)" = "reads.in",
                    "Pre-Filtered reads (pairs)" = "reads.out")

    # View result
    pre_filtered_summary
    ```

#### Remove Primers with Cutadapt

The Cutadapt software (DOI:10.14806/ej.17.1.200) is used to remove primer sequences from reads, also capable of handling read-through scenarios.

-   **Configure output paths.** Map sample names to new Cutadapt output filenames, ensuring separation by primer set if necessary.

    ```{r}
    # Map sample names to Cutadapt output files (identify paired samples for Cutadapt input)
    paired_files_per_primer <- sample_idx_tbl %>% 
      dplyr::filter(Stage %in% c("FWD_R1_paired", 
                                 "FWD_R2_paired")) %>% 
      dplyr::filter(file.exists(`Read file`))

    # Generate Cutadapt output paths by replacing 'paired' with 'cutadapt'
    cut_files_per_primer <- paired_files_per_primer %>%  
      dplyr::mutate(Stage = gsub(Stage, pattern = "_paired", 
                                 replacement = "_cutadapt"),
                    `Read file` = stringr::str_replace_all(string = `Read file`,
                                                           pattern = "paired",
                                                           replacement = "cutadapt"))

    # Combine input (paired) and output (cutadapt) metadata
    cut_files_per_primer <- dplyr::bind_rows(cut_files_per_primer,     
                                             paired_files_per_primer)

    # OPTIONAL Separate paths by primer if multiple primers exist per sample 
    # cut_files_per_primer <- cut_files_per_primer %>% 
    #   tidyr::separate_longer_delim(cols = "Primer", 
    #                                delim = ";") %>% 
    #   dplyr::mutate(`Read file` = dplyr::case_when(
    #     Stage %in% c("FWD_R1_cutadapt", 
    #                  "FWD_R2_cutadapt") ~ 
    #       stringr::str_replace_all(string = `Read file`,
    #                                pattern = paste0(cutadapt_libs, "/"),
    #                                replacement = paste0(cutadapt_libs, "/", 
    #                                                     Primer, "--")),
    #     TRUE ~ `Read file`))
    ```

-   **Extract primer sequences.** For each primer set, retrieve the forward and reverse primer sequences — and their reverse complements — from the orientations table. These are needed to construct the correct adapter flags for both R1 and R2 reads, accounting for read-through at the 3′ end.

    ```{r}
    # Identify unique primer sets to process
    primer_sets <- primers_all_orients$`Primer pair` %>%
      stringr::str_remove_all("_FWD$|_REV$") %>%
      BiocGenerics::unique()

    # Initialize list to store flags per primer set
    cutadapt_flags <- list()

    # Store flags using grep to find specific orientation keys constructed in 'allOrients' function
    for (PRIMER_SET in primer_sets) {
      
      # Forward primer (for R1 5')
      fwd_seq <- primers_all_orients$Sequence[grep(paste0(PRIMER_SET, "_FWD_Forward"),
                                                   primers_all_orients$`Orientation name`)]
      
      # Reverse complement of Forward primer (for R2 3' read-through)
      fwd_rc_seq <- primers_all_orients$Sequence[grep(paste0(PRIMER_SET, "_FWD_RevComp"), 
                                                      primers_all_orients$`Orientation name`)]
      
      # Reverse primer (for R2 5')
      rev_seq <- primers_all_orients$Sequence[grep(paste0(PRIMER_SET, "_REV_Forward"),
                                                   primers_all_orients$`Orientation name`)]
      
      # Reverse complement of Reverse primer (for R1 3' read-through)
      rev_rc_seq <- primers_all_orients$Sequence[grep(paste0(PRIMER_SET,  "_REV_RevComp"),
                                                      primers_all_orients$`Orientation name`)]
      
      # Store sequences for use in flag construction
      cutadapt_flags[[PRIMER_SET]] <- list(fwd_seq = fwd_seq,
                                           fwd_rc_seq = fwd_rc_seq,
                                           rev_seq    = rev_seq,
                                           rev_rc_seq = rev_rc_seq)
      }
    ```

-   **Build adapter flags.** Define the -g, -a, -G, and -A Cutadapt flags for each primer set.

    -   **`-g`**: Removes the forward primer from the 5′ end of R1. Use this when the forward primer is expected at the start of the forward read.

    -   **`-a`**: Removes the reverse complement of the reverse primer from the 3′ end of R1. Only relevant when read-through occurs and the sequencing read extends beyond the amplicon insert into the opposite primer.

    -   **`-G`**: Removes the reverse primer from the 5′ end of R2. Mirrors the -g logic for the reverse read.

    -   **`-A`**: Removes the reverse complement of the forward primer from the 3′ end of R2. Again, only active when read-through is present.

    ```{r}
        for (PRIMER_SET in primer_sets) {
          
          seqs <- cutadapt_flags[[PRIMER_SET]]
          
          # -g: Anchor FWD to start of R1; -a: Remove REV_RC from end of R1
          R1_flags <- c("-g", seqs$fwd_seq,
                        "-a", seqs$rev_rc_seq)
          
          # -G: Anchor REV to start of R2; -A: Remove FWD_RC from end of R2
          R2_flags <- c("-G", seqs$rev_seq,
                        "-A", seqs$fwd_rc_seq)
          
          # Store back for use in the execution step
          cutadapt_flags[[PRIMER_SET]]$R1_flags <- R1_flags
          cutadapt_flags[[PRIMER_SET]]$R2_flags <- R2_flags
          }
    ```

-   **Prepare file lists per primer set.** Filter the combined metadata table to the current primer set and pivot to wide format, separating input (paired) and output (cutadapted) file paths. Only samples with both R1 and R2 input files present on disk are retained.

    ```{r}
    for (PRIMER_SET in primer_sets) {
      
      # Filter samples belonging to this primer set and pivot to wide format
      PRIMER_SET_files <- cut_files_per_primer %>%
        dplyr::filter(stringr::str_detect(Primer, PRIMER_SET)) %>%
        tidyr::pivot_wider(id_cols = c("Unique_File_name", 
                                       "Unique_File_name_Primer"),
                           names_from  = "Stage",
                           values_from = "Read file")

      # Validate input file existence
      valid_indices <- file.exists(PRIMER_SET_files$FWD_R1_paired) &
        file.exists(PRIMER_SET_files$FWD_R2_paired)
      
      # Define input/output vectors restricted to valid pairs
      cutadapt_flags[[PRIMER_SET]]$fnFs_in <- PRIMER_SET_files$FWD_R1_paired[valid_indices]
      
      cutadapt_flags[[PRIMER_SET]]$fnRs_in <- PRIMER_SET_files$FWD_R2_paired[valid_indices]
      
      cutadapt_flags[[PRIMER_SET]]$fnFs_out <- PRIMER_SET_files$FWD_R1_cutadapt[valid_indices]
      
      cutadapt_flags[[PRIMER_SET]]$fnRs_out <- PRIMER_SET_files$FWD_R2_cutadapt[valid_indices]
      }
    ```

-   **Execute Cutadapt.** Iterate over each valid input file pair and invoke the Cutadapt **`system2`** command. **Key parameters to configure for your dataset**:

    -   **`-j`**: Number of CPU threads. Set to `0` to let Cutadapt auto-detect all available cores.

    -   **`-n`**: Number of adapter trimming rounds per read. Set to `1` for standard libraries; increase if read-through could place a primer and its reverse complement on the same read.

    -   **`--minimum-length`**: Discard reads shorter than this value after trimming. Prevents short artefactual fragments from reaching DADA.

    -   **`--discard-untrimmed`**: Removes any pair in which no adapter was detected. Appropriate when all reads are expected to originate from the target amplicon; omit if you want to retain undetected reads.

    -   **`--match-read-wildcards`**: Enables IUPAC degenerate base matching in adapter sequences. Required for primers containing ambiguous positions.

    -   **`--pair-filter`**: Determines when a pair is discarded. **`any`** (default) removes the pair if either read fails; both removes it only if both fail. When combined with **`--discard-untrimmed`**, always prefer any to avoid retaining pairs where only one read carries the expected primer.

    ```{r}
    for (PRIMER_SET in primer_sets) {
      
      flags <- cutadapt_flags[[PRIMER_SET]]
      fnFs_in <- flags$fnFs_in
      fnRs_in <- flags$fnRs_in
      fnFs_out <- flags$fnFs_out
      fnRs_out <- flags$fnRs_out
      
      if (length(fnFs_in) > 0) {
        for (i in seq_along(fnFs_in)) {
          system2(command = cutadapt, 
                  args = c(flags$R1_flags, 
                           flags$R2_flags,
                           "-j", "70", # teoricamente se deixar 0 
                                      #escolhe o numero de cores vazios, mas estava usando todos
                           "-n", "2",
                           "--minimum-length", "30",
                           "--discard-untrimmed",
                           "--match-read-wildcards",
                           # "--pair-filter", "both",
                           "-o", fnFs_out[i],
                           "-p", fnRs_out[i], fnFs_in[i], fnRs_in[i]
                  ))
          }
        } else {
          message(paste("No valid input files found for primer set:", PRIMER_SET))
        }
      }
    ```

#### Primer Removal Verification

This section confirms that Cutadapt successfully excised all primer sequences before reads enter quality filtering and ASV inference. Using the same multi_primerHits() logic and primers_all_orients table from the raw-read detection step, residual primer hits are counted across all four orientations in every trimmed R1 and R2 file. Counts are then normalized by total reads per file to allow proportional comparison across samples.

A well-trimmed library should show zero counts in all orientations for all primer sequences. Any residual Forward signal at the 5′ position of a read indicates that the corresponding adapter flag (-g or -G) failed to anchor and trim the primer — typically caused by a mismatch between the primer sequence entered and the one actually used in the library preparation, or by an incorrect orientation assignment in primers_n_samples. Residual RevComp signal at the 3′ end, on the other hand, indicates incomplete read-through trimming — check that the -a / -A flags carry the correct reverse-complement sequences and that --pair-filter any was active during the Cutadapt run. If residual hits exceed \~1% of reads in any sample, revisit the adapter flags before proceeding

-   **Collect Cutadapt output files.** Builds a long-format table of trimmed FASTQ paths (R1 and R2) per sample and primer set.

    ```{r}
    # Restrict to Cutadapt outputs (R1/R2 trimmed files that actually exist)
    cutadapt_check_seqs_tbl <- cut_files_per_primer %>%
      dplyr::filter(stringr::str_detect(Stage, "cutadapt")) %>%
      dplyr::filter(file.exists(`Read file`)) %>%
      dplyr::select(`Read file`, Unique_File_name = Unique_File_name_Primer,
                    Stage, Project, Run, Primer)
    #Show table
    View(cutadapt_check_seqs_tbl)
    ```

-   **Per-file primer hits.** Counts residual primer occurrences in the Cutadapt-trimmed files using the same `multi_primerHits()` function and `primers_all_orients` table from the raw-read detection step. All four orientations — Forward, Complement, Reverse, and RevComp — are searched simultaneously across every R1 and R2 file in parallel. The resulting table is joined with sample metadata and annotated with total read counts per file to allow proportional comparison across samples.

    ```{r}
    # Build named vector of Cutadapt FASTQ paths (names = sample identifiers)
    cutadapt_reads_seqs <- cutadapt_check_seqs_tbl$`Read file`
    names(cutadapt_reads_seqs) <- cutadapt_check_seqs_tbl$Unique_File_name

    # Extract all primer orientations as named vector (same object used pre-Cutadapt)
    primers_seqs <- primers_all_orients %>%
      dplyr::pull(Sequence, name = `Orientation name`)

    # Set up parallel processing (all cores minus 2)
    cores_to_use <- future::availableCores() - 2
    future::plan(future::multisession(workers = cores_to_use))

    # Count residual primer occurrences across all files × all primer orientations (parallel)
    {
      tictoc::tic("Counting primers")
      
      message("Searching residual primers across ", length(cutadapt_reads_seqs),
              " post-Cutadapt files using ", cores_to_use, " cores. ",
              "This may take a few minutes...")

      primers_in_cutadapt_reads <- 
        furrr::future_map_dfr(cutadapt_reads_seqs,
                              .f = multi_primerHits,
                              primers = primers_seqs,
                              .options = furrr::furrr_options(seed = NULL))
      
      tictoc::toc()
    }

    # Join with sample metadata and calculate total reads per file
    primers_in_cutadapt_reads <- primers_in_cutadapt_reads %>%
      dplyr::left_join(cutadapt_check_seqs_tbl, 
                       by = "Read file") %>%
      dplyr::mutate(Read = dplyr::if_else(stringr::str_detect(Stage, "R1"), "R1", "R2")) %>%
      tidyr::unite(col = "Unique_File_name_Read",
                   Unique_File_name,
                   Read,
                   sep = " ",
                   remove = FALSE) %>%
      dplyr::mutate(`Total reads` = ShortRead::countFastq(dirPath = `Read file`)[, 1])

    message("✓ Post-Cutadapt primer search complete. ",
            "Results in `primers_in_cutadapt_reads`:\n",
            "  ", length(unique(primers_in_cutadapt_reads$Unique_File_name)),
            " samples × ", length(primers_seqs), " primer orientations across ",
            length(cutadapt_reads_seqs), " files.")
    ```

-   **Prepare data.** Transforms the post-Cutadapt primer hits table into a long format suitable for visualization with `ggplot2`. Columns corresponding to Complement and Reverse orientations are dropped before reshaping. Rows with zero counts are removed, so that the resulting table contains only genuine residual primer signals.

    ```{r}
    # Convert to long format, retaining only Forward and RevComp orientations
    primers_in_cutadapt_reads_long <- primers_in_cutadapt_reads %>%
      dplyr::select(!dplyr::ends_with(c("Complement", "Reverse"))) %>%
      tidyr::gather(key = "Sequences", value = "Count",
                    dplyr::ends_with("Forward"),
                    dplyr::ends_with("RevComp")) %>%
      dplyr::mutate(Sequences = as.factor(Sequences),
                    Unique_File_name = factor(Unique_File_name, 
                                              levels = sample_levels)) %>%
      dplyr::select(-c(`Read file`, Stage)) %>%
      # Remove columns that are entirely NA
      dplyr::select(where(function(x) any(!is.na(x)))) %>%
      # Keep only rows with residual primer signal
      dplyr::filter(Count != 0)

    # Create ordered levels for the y-axis (File × Read direction)
    Unique_File_name_Read_levels <- primers_in_cutadapt_reads_long %>%
      dplyr::pull(Unique_File_name_Read) %>%
      base::unique()
    ```

-   **Generate diagnostic plot.** If residual primers are detected, visualize their distribution across samples.

    ```{r}
    # Isolate non-zero residual hits
    residual_hits <- primers_in_cutadapt_reads_long %>%
      dplyr::filter(Count != 0)

    if (nrow(residual_hits) > 0) {
      for (PRJ in all_projects) {
        options(scipen = 10000)

        # 1. Filter and prepare data ONCE for the current project
        # This avoids filtering the large dataset multiple times inside the loop
        project_data <- residual_hits %>%
          dplyr::filter(Project == PRJ) %>%
          dplyr::mutate(
            # Pre-format labels and calculate percentages to keep ggplot clean
            Sequences_lbl = stringr::str_replace_all(Sequences, "_", "\n"),
            Primer_lbl = stringr::str_replace_all(Primer,    "_", "\n"),
            Pct_Reads = Count / `Total reads` * 100)

        if (nrow(project_data) == 0) next

        # 2. Calculate dynamic dimensions based on actual data
        n_samples <- length(unique(project_data$Unique_File_name_Read))
        n_primers <- length(unique(project_data$Primer))

        # 3. Create the plot object
        primers_tile_residual <-
          project_data %>%
          ggplot2::ggplot(aes(y = Unique_File_name_Read)) +

          # Bar layer
          ggplot2::geom_bar(aes(x = `Total reads`,
                                fill = Pct_Reads,
                                color = Primer,
                                group = Sequences),
                            stat = "identity",
                            position = "dodge",
                            linewidth = 0.05,
                            linetype = 2,
                            alpha = 0.5) +

          # Text labels for counts
          ggplot2::geom_text(ggplot2::aes(x = `Total reads`, label = Count),
                             hjust = 0,
                             size  = 1) +

          # Color scales
          ggplot2::scale_fill_gradientn(name = "% reads with\nresidual primer",
                                        colours = primer_colours,
                                        values = c(0, 1),
                                        na.value = "white") +

          # Axes and Guides
          ggplot2::scale_y_discrete(limits = rev) +
          ggplot2::scale_x_log10(breaks = c(10, 100, 1000, 10000, 100000),
                                 expand = ggplot2::expansion(mult = c(0, 0.2))) +
          ggplot2::scale_alpha(guide = "none") +
          ggplot2::guides(color = guide_legend(override.aes = list(fill = "white",
                                                                   size = 10))) +

          # Theme adjustments
          ggplot2::theme_light(base_line_size = 0.025, base_size = 6) +
          ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1),
                         strip.text.y = element_text(size = 10),
                         plot.title = element_text(size = 10),
                         plot.subtitle = element_text(size = 6),
                         legend.text = element_text(size = 5),
                         legend.title = element_text(size = 7)) +

          # Dynamic horizontal lines based on actual sample count
          ggplot2::geom_hline(yintercept = seq(0.5, n_samples + 0.5, 2),
                              color = "darkgrey") +

          # Titles and Facets
          ggplot2::labs(title = paste0(PRJ, " — Post-Cutadapt Residual Primers"),
                        subtitle = paste0("Non-zero counts indicate incomplete primer removal. ","Forward and RevComp orientations only (max.mismatch = 1)."),
                        x = "Primer orientation",
                        y = "Sample (file × read direction)") +
          ggplot2::facet_grid(rows = ggplot2::vars(Primer_lbl),
                              cols = ggplot2::vars(Sequences_lbl),
                              scales = "free",
                              space = "free")
        # 4. Save the plot
        ggplot2::ggsave(file = paste0(track_path, "/", PRJ,
                             "-residual_primers_post_cutadapt.pdf"),
                        plot = primers_tile_residual,
                        device = "pdf",
                        width = 10 + n_primers * 10,
                        height = n_samples * 0.3 + 3,
                        units = "cm",
                        limitsize = FALSE,
                        dpi = 600)
        
        # 5. Open the plot
        utils::browseURL(paste0(track_path, "/", PRJ, "-residual_primers_post_cutadapt.pdf"))
      }

    } else {
      message("✓ Primer removal successful — no residual primer hits detected ",
              "across all ", length(cutadapt_reads_seqs),
              " post-Cutadapt files.")
    }
    ```

#### Quality checking post-Cutadapt

With primer sequences removed, the Cutadapt output files now reflect only the biological sequence of the target amplicon. Running FastQC and MultiQC at this stage — before any quality-based filtering — reveals the true per-position quality landscape of the amplicon reads, free from the confounding effect of primer bases at the 5′ end. The resulting quality profiles should be used to set the `truncLen`, `maxEE`, and `minLen` parameters in the `filterAndTrim` step that follows. Pay particular attention to the per-position mean quality score plot: identify the position at which quality drops below Q30 (or Q25 for R2) and use that as the truncLen cutoff. The read length distribution panel will also reveal whether Cutadapt discarded a large fraction of reads due to the `--minimum-length` threshold — if so, consider relaxing it before re-running.

-   **Check that trimmed files exist.** Confirms that Cutadapt output files exist before running FastQC/MultiQC.

    ```{r}
    cutadapted_read_files <- cut_files_per_primer %>%
      dplyr::filter(stringr::str_detect(Stage, "_cutadapt")) %>%
      dplyr::pull(`Read file`) %>%
      normalizePath() %>%
      unique()

    cutadapted_read_files

    if (length(cutadapted_read_files) > 0) {
      
      message(length(cutadapted_read_files), " read files found for QC.")
      
      } else {

        stop("No read files found for QC. Check if Cutadapt ran successfully and output files were written.")
      }
    ```

-   **Run FastQC and MultiQC by project.** Creates per‑project folders and runs quality control in batch using system calls.

    ```{r}
    for (PRJ in all_projects) {
      # Create output directories
      fastqc_path  <- file.path(qual_path, PRJ, "fastqc")
      multiqc_path <- file.path(qual_path, PRJ, "multiqc")
      
      dir.create(fastqc_path, recursive = TRUE, showWarnings = FALSE)
      dir.create(multiqc_path, recursive = TRUE, showWarnings = FALSE)
      
      # Filter reads belonging to this project
      cutadapted_path <- cut_files_per_primer$`Read file`[cut_files_per_primer$Project == PRJ] %>%
        unique()
      
      project_cutadapted_reads <- cutadapted_read_files[stringr::str_detect(cutadapted_read_files,
                                                                            paste0(cutadapted_path,
                                                                                   collapse = "|"))]
      
      # 1. Run FastQC
      cmd_fastqc <- paste0("echo '", paste(project_cutadapted_reads, collapse = "\n"), "' | ",
                           "parallel fastqc {1} -t 20 -o ", 
                           fastqc_path)
      
      system(cmd_fastqc)
        
      # 2. Run MultiQC
      cutadapt_report_name <- paste0(PRJ, "--post_cutadapt_quality-MultiQC_report.html")
      
      cmd_multiqc <- paste0("multiqc --interactive ",
                            fastqc_path, " --filename ",
                            file.path(multiqc_path, cutadapt_report_name))
      
      system(cmd_multiqc)
            
      # 3. Show MultQC report
      browseURL(file.path(multiqc_path, cutadapt_report_name))
    }
    ```

#### Filter and Trim

With primers removed by Cutadapt, reads now reflect only the biological sequence of the target amplicon and are ready for quality-based filtering. Unlike the pre-filtering step — which targeted only ambiguous bases — `filterAndTrim` is now applied with parameters that directly evaluate read quality, length, and integrity before ASV inference.

-   **Prepare file paths.** Map Cutadapt output files to their respective future filtered paths.

    ```{r}
    # Map Cutadapt output to filtered file paths
     filter_files_per_primer <- cut_files_per_primer %>%
      dplyr::filter(stringr::str_detect(Stage, "_cutadapt")) %>%
      dplyr::filter(file.exists(`Read file`)) %>%
      tidyr::pivot_wider(id_cols = c("Project",
                                     "Unique_File_name", 
                                     "Unique_File_name_Primer",
                                     "Primer", 
                                     "Run"),
                         names_from  = "Stage",
                         values_from = "Read file") %>%
      dplyr::mutate(FWD_R1_filtered = stringr::str_replace_all(FWD_R1_cutadapt, "cutadapt", "filtered"),
                    FWD_R2_filtered = stringr::str_replace_all(FWD_R2_cutadapt, "cutadapt", "filtered")) %>%
      tidyr::unite(col    = "Unique_File_name_Primer",
                   Unique_File_name, Primer,
                   sep    = "--",
                   remove = FALSE)

    # Show prepared file paths
    View(filter_files_per_primer)
    ```

-   **Define filtering function.** Create a wrapper for `filterAndTrim` applying quality thresholds.

-   **Execute filtering.** Run DADA2's native `filterAndTrim` on all samples simultaneously.

The key parameters to adjust for your dataset are:

-   `maxEE`: Maximum number of expected errors tolerated per read. Lower values (e.g., c(2, 2)) are more stringent and recommended as a starting point. If read retention drops below \~70%, consider relaxing to c(2, 5) for the reverse read, which typically shows steeper quality decay.

-   `truncQ`: Truncates the read at the first base with a Phred quality score ≤ this value. The default of 2 effectively removes reads with untrustworthy terminal bases.

-   `minLen`: Discards reads shorter than this threshold after trimming. Set according to the minimum expected amplicon length for your marker (e.g., 100 for MiFish/NeoFish 12S amplicons).

-   `truncLen`: Truncates all reads to a fixed length. Useful for markers with predictable amplicon sizes when quality degrades before the end of the read. Leave unset (or set to 0) for variable-length amplicons. Inspect `dada2::plotQualityProfile()` on your Cutadapt output files to decide.

-   `matchIDs`: Ensures that the sequence IDs of R1 and R2 exactly match, discarding pairs that have become unsynchronized during upstream processing.

-   `rm.phix`: Removes reads matching the PhiX bacteriophage genome, a standard Illumina sequencing control that can contaminate libraries.

    ```{r}
    # Calculate number of samples and files being processed
    n_samples <- length(filter_files_per_primer$Unique_File_name)

    n_files <- length(filter_files_per_primer$FWD_R1_filtered) + 
      length(filter_files_per_primer$FWD_R2_filtered)

    # Run filtering in parallel across all read pairs
    {
      message("Filtering ", n_files, " files from ", n_samples, " samples. This may take a few minutes...")
      
      tictoc::tic("Filtering samples")
      
      sample_filtered_out <- 
        dada2::filterAndTrim(fwd = filter_files_per_primer$FWD_R1_cutadapt,
                             filt = filter_files_per_primer$FWD_R1_filtered,
                             rev = filter_files_per_primer$FWD_R2_cutadapt,
                             filt.rev = filter_files_per_primer$FWD_R2_filtered,
                             maxEE = c(2, 2),
                             truncQ = 2,
                             minLen = 100,
                             matchIDs = TRUE,
                             rm.phix = TRUE,
                             # truncLen = c(0, 0),
                             multithread = TRUE,
                             compress = TRUE,
                             verbose = TRUE)
      
      tictoc::toc()
    }
    ```

#### Quality checking post-filtering

After filterAndTrim has been applied, this section re-runs FastQC and MultiQC on the filtered output files to confirm that the quality thresholds were effective. Unlike the post-Cutadapt check — which informed the choice of filtering parameters — this step serves as a validation: it verifies that low-quality bases were successfully removed, that read length distributions conform to the expected amplicon size range, and that no systematic artefacts were introduced by truncation.

The read retention summary produced by `filterAndTrim` (pre_filtered_summary) should be inspected alongside these reports. A well-calibrated filtering step typically achieves an acceptable read retention rate of 80% to 90% (or higher) of the input reads. If the filtered quality profiles still show a high proportion of low-quality bases, or if the retention rate falls significantly below this acceptable range, return to the filterAndTrim parameters to carefully adjust `maxEE`, `truncLen`, or `minLen` for your specific dataset and marker.

-   **Check that trimmed files exist.** Confirms that Cutadapt output files exist before running FastQC/MultiQC.

    ```{r}
    filtered_read_files <- c(filter_files_per_primer$FWD_R1_filtered,
                             filter_files_per_primer$FWD_R2_filtered) %>%
      normalizePath() %>%
      unique()

    filtered_read_files

    if (length(filtered_read_files) > 0) {
      
      message(length(filtered_read_files), " filtered read files found for QC.")
      
      } else {

        stop("No read files found for QC. Check if filterAndTrim ran successfully and output files were written.")
      }
    ```

-   **Run FastQC and MultiQC by project.** Creates per‑project folders and runs quality control in batch using system calls.

    ```{r}
    for (PRJ in all_projects) {
      # Create output directories
      fastqc_path  <- file.path(qual_path, PRJ, "fastqc")
      multiqc_path <- file.path(qual_path, PRJ, "multiqc")
      
      dir.create(fastqc_path, recursive = TRUE, showWarnings = FALSE)
      dir.create(multiqc_path, recursive = TRUE, showWarnings = FALSE)
      
      # Filter reads belonging to this project
      filtered_path <- filter_files_per_primer %>%
        dplyr::filter(Project == PRJ) %>%
        dplyr::select(FWD_R1_filtered, FWD_R2_filtered) %>%
        unlist() %>%
        as.character() %>%
        unique()
      
      project_filtered_reads <- filtered_read_files[stringr::str_detect(filtered_read_files,
                                                                        paste0(filtered_path,
                                                                               collapse = "|"))]
      
      # 1. Run FastQC
      cmd_fastqc <- paste0("echo '", paste(project_filtered_reads, collapse = "\\n"), "' | ",
                           "parallel fastqc {1} -t 20 -o ", 
                           fastqc_path)
      
      system(cmd_fastqc)
        
      # 2. Run MultiQC
      filtered_report_name <- paste0(PRJ, "--post_filtered_quality-MultiQC_report.html")
      
      cmd_multiqc <- paste0("multiqc --interactive ",
                            fastqc_path, " --filename ",
                            file.path(multiqc_path, filtered_report_name))
      
      system(cmd_multiqc)
            
      # 3. Show MultQC report
      browseURL(file.path(multiqc_path, filtered_report_name))
    }
    ```

### Learn Error Rates and Infer ASVs

With primers removed and low-quality bases filtered out, the reads are ready for the core of the DADA2 pipeline: the inference of Amplicon Sequence Variants (ASVs). Unlike traditional clustering methods that group reads into OTUs based on an arbitrary similarity threshold (e.g., 97%), DADA2 uses a machine-learning model to distinguish true biological variants from sequencing errors down to a single nucleotide difference. Because error profiles depend on the specific sequencing chemistry, machine state, and run conditions, the error rates must be learned independently for each sequencing run. Once the error model is computed, identical reads are dereplicated to reduce computational load, and the core `dada()` algorithm is applied to infer the true ASVs present in each sample.

-   **Learn Error Rates.** Iterates through each unique sequencing run to calculate its specific error model and generates diagnostic plots for quality control. This is a computationally intensive step.

    ```{r}
    # Initialize lists to store error models per run
    err_models_F <- list()
    err_models_R <- list()

    # Loop through each sequencing run to learn errors
    for (seq_run in unique(filter_files_per_primer$Run)) {
      
      message(paste0("Learning errors for Run: ", seq_run))
      
      # Get filtered file lists for this run
      seq_files <- filter_files_per_primer %>%
        dplyr::filter(Run == seq_run) %>%
        dplyr::filter(file.exists(FWD_R1_filtered) & file.exists(FWD_R2_filtered))

      seq_fnFs <- seq_files$FWD_R1_filtered
      seq_fnRs <- seq_files$FWD_R2_filtered
      
      # Learn Error Rates
      tictoc::tic("Learning error rates")
      
      message("  Learning forward errors...")
      err_models_F[[seq_run]] <- dada2::learnErrors(fls = seq_fnFs, 
                                                    multithread = TRUE, 
                                                    randomize = TRUE, 
                                                    verbose = TRUE)
                                     
      message("  Learning reverse errors...")                               
      err_models_R[[seq_run]] <- dada2::learnErrors(fls = seq_fnRs, 
                                                    multithread = TRUE, 
                                                    randomize = TRUE, 
                                                    verbose = TRUE)
      
      tictoc::toc()
      
      
      # Error plot R1
      err_plotF <- dada2::plotErrors(err_models_F[[seq_run]], nominalQ = TRUE) + 
        ggplot2::ggtitle(paste("Forward Errors -", seq_run))
        
      # Error plot R2
      err_plotR <- dada2::plotErrors(err_models_R[[seq_run]], nominalQ = TRUE) + 
        ggplot2::ggtitle(paste("Reverse Errors -", seq_run))
      
      # Save error plots for quality control review
      error_plot_path <- file.path(error_path, paste0(seq_run, "_error_rates.pdf"))
      
      ggplot2::ggsave(filename = error_plot_path,
                      plot = gridExtra::arrangeGrob(err_plotF, err_plotR, ncol = 2),
                      width = 12, height = 5)
                      
      message(paste0("  Error plots saved to: ", error_plot_path))
      
      # Open the plot
      utils::browseURL(error_plot_path)
    }
    ```

-   **Dereplicate and Infer ASVs.** Using the pre-calculated error models, identical reads are dereplicated to reduce computational load, and the core `dada()` algorithm is applied to infer the ASVs present in each sample.

    ```{r}
    # Initialize empty lists to store DADA2 output objects per sample
    all_dadaFs <- list()
    all_dadaRs <- list()
    all_derep_forward <- list()
    all_derep_reverse <- list()

    # Loop through each sequencing run for inference
    for (seq_run in unique(filter_files_per_primer$Run)) {
      
      message(paste0("Processing inference for Run: ", seq_run))
      
      # Get filtered file lists for this run
      seq_files <- filter_files_per_primer %>%
        dplyr::filter(Run == seq_run) %>%
        dplyr::filter(file.exists(FWD_R1_filtered) & file.exists(FWD_R2_filtered))

      seq_fnFs <- seq_files$FWD_R1_filtered
      seq_fnRs <- seq_files$FWD_R2_filtered
      names(seq_fnFs) <- seq_files$Unique_File_name_Primer
      names(seq_fnRs) <- seq_files$Unique_File_name_Primer

      # Dereplication
      message("  Dereplicating reads...")
      seq_derep_forward <- dada2::derepFastq(seq_fnFs, verbose = FALSE)
      seq_derep_reverse <- dada2::derepFastq(seq_fnRs, verbose = FALSE)
      
      # Sample Inference (DADA2 Algorithm) applying the specific run's error model
      message("  Inferring ASVs...")
      seq_dadaFs <- dada2::dada(seq_derep_forward, err = err_models_F[[seq_run]],
                                multithread = TRUE, verbose = FALSE)
      seq_dadaRs <- dada2::dada(seq_derep_reverse, err = err_models_R[[seq_run]],
                                multithread = TRUE, verbose = FALSE)
      
      # Append to global lists maintaining list hierarchy
      all_dadaFs <- append(all_dadaFs, seq_dadaFs)
      all_dadaRs <- append(all_dadaRs, seq_dadaRs)
      all_derep_forward <- append(all_derep_forward, seq_derep_forward)
      all_derep_reverse <- append(all_derep_reverse, seq_derep_reverse)
    }

    message("ASV Inference Complete across all runs.")
    ```

### Construct Sequence Table and Remove Chimeras

Following ASV inference, the independently processed forward and reverse reads must be recombined into a single, full-length amplicon sequence. DADA2 strictly requires a perfect overlap between the paired ends; reads that fail to align precisely due to poor quality or insufficient overlap length are discarded. Once the full amplicons are reconstructed, a critical filtration step is applied to remove chimeric sequences. Chimeras are artificial recombinations formed during PCR when an aborted amplicon extension acts as a primer in the next cycle, fusing segments of two distinct biological sequences. Because DADA2 resolves sequences at single-nucleotide precision, these chimeras are readily identified and removed, producing a pristine final sequence table.

-   **Merge paired reads.** Combine the inferred forward and reverse ASVs to reconstruct the full target amplicon. Only reads that perfectly overlap (without mismatches) by at least the specified minimum length are successfully merged.

    ```{r}
    {
      message("Merging paired-end reads...")
      all_mergers <- dada2::mergePairs(dadaF = all_dadaFs,
                                       derepF = all_derep_forward,
                                       dadaR = all_dadaRs,
                                       derepR = all_derep_reverse,
                                       minOverlap = 10,
                                       maxMismatch = 0,
                                       verbose = FALSE)
      
      # Construct Sequence Table (ASV table)
      mergers_seqtab <- dada2::makeSequenceTable(all_mergers)
      
      # Check raw ASV length distribution
      message("Amplicon length distribution before chimera removal:")
      table(nchar(dada2::getSequences(mergers_seqtab)))
    }
    ```

-   **Construct Sequence Table (ASV table) for unmerged reads.** When paired-end reads do not sufficiently overlap to reconstruct the full amplicon, the analysis must proceed with R1 and R2 reads independently, generating separate ASV tables for each read direction.

    ```{r}
    # Construct Sequence Tables for unmerged reads.
    R1_seqtab <- dada2::makeSequenceTable(all_dadaFs)
    R2_seqtab <- dada2::makeSequenceTable(all_dadaRs)
    ```

-   **Remove chimeras.** Identify and remove artifactual sequences formed during PCR amplification when two different biological sequences combine. The `method="consensus"` compares sequences across samples to confidently detect these artifacts.

    ```{r}
    {
      message("Identifying and removing chimeric sequences...")
      
      mergers_seqtab.nochim <- dada2::removeBimeraDenovo(mergers_seqtab,
                                                         method = "consensus",
                                                         multithread = TRUE,
                                                         verbose = TRUE)
      
      # Calculate chimera removal stats
      chimera_prop <- base::sum(mergers_seqtab.nochim) / base::sum(mergers_seqtab)
      message(paste0("Retained ", base::round(chimera_prop * 100, 2), 
                     "% of reads after chimera removal."))
      
      # Inspect final length distribution
      message("Final Amplicon length distribution (No Chimeras):")
      table(nchar(dada2::getSequences(mergers_seqtab.nochim)))
      
      # Save the final pristine sequence table as an R object to prevent data loss
      saveRDS(mergers_seqtab.nochim, file = file.path(nochim_path, "seqtab.nochim.rds"))
      message("Saved merged final sequence table to: ", file.path(nochim_path,
                                                                  "seqtab.nochim.rds"))
    }
    ```

-   **Remove chimeras (Unmerged reads).** Just as with the merged sequences, the independent R1 and R2 sequence tables may contain chimeric amplicons generated during the PCR amplification steps.

    ```{r}
    {
      message("Identifying and removing chimeric sequences from unmerged R1 and R2 reads...")
      
      # Remove chimeras from R1
      R1_seqtab.nochim <- dada2::removeBimeraDenovo(R1_seqtab, 
                                                  method = "consensus", 
                                                  multithread = TRUE, 
                                                  verbose = TRUE)
      
      # Calculate chimera removal stats for R1 reads
      chimera_prop_R1 <- base::sum(R1_seqtab.nochim) / base::sum(R1_seqtab)
      message(paste0("Retained ", base::round(chimera_prop_R1 * 100, 2), 
                     "% of R1 reads after chimera removal."))
      
      # Save the final pristine sequence table as an R object to prevent data loss
      saveRDS(R1_seqtab.nochim, file = file.path(nochim_path, "R1_seqtab.nochim.rds"))
      message("Saved R1 final sequence table to: ", file.path(nochim_path,
                                                              "R1_seqtab.nochim.rds"))

      # Remove chimeras from R2
      R2_seqtab.nochim <- dada2::removeBimeraDenovo(R2_seqtab, method = "consensus", 
                                                  multithread = TRUE, verbose = TRUE)
      
      # Calculate chimera removal stats for R2 reads
      chimera_prop_R2 <- base::sum(R2_seqtab.nochim) / base::sum(R2_seqtab)
      message(paste0("Retained ", base::round(chimera_prop_R2 * 100, 2), 
                     "% of R2 reads after chimera removal."))
      
      # Save the final pristine sequence table as an R object to prevent data loss
      saveRDS(R2_seqtab.nochim, file = file.path(nochim_path, "R2_seqtab.nochim.rds"))
      message("Saved R2 final sequence table to: ", file.path(nochim_path,
                                                              "R2_seqtab.nochim.rds"))
    }
    ```

### Track Read Retention

A fundamental quality control practice in metabarcoding is monitoring the proportion of reads that successfully survive each step of the bioinformatic pipeline. Extreme drops in read counts at specific stages often indicate technical issues. For instance, massive losses during the initial filtering step may point to poor sequencing quality or overly aggressive truncation parameters, whereas massive losses during the merging step usually indicate that the reads are too short to properly overlap (e.g., severe read-through or improper size selection). This section aggregates the sequence retention metrics from raw inputs through the final chimera removal, generating a comprehensive diagnostic table for auditing and reporting.

-   **Compile statistics.** Gather read counts from every processing step (Raw/Cutadapt → Filtered → Denoised → Merged → Non-chimeric) to assess pipeline quality and detect samples where massive data loss ocurred.

    ```{r}
    {
      message("Compiling read retention statistics across all pipeline stages...")
      
      # 1. Helper function to extract counts from DADA objects
      getN <- function(x) { base::sum(dada2::getUniques(x)) }
      
      # 2. Extract RAW read counts
      # We only count R1 because paired-end files have identical line counts
      raw_reads_paths <- sample_idx_tbl %>%
        dplyr::filter(Stage == "FWD_R1") %>%
        dplyr::pull(`Read file`, name = Unique_File_name)
      
      # 3. Extract Cutadapt read counts and build composite key
      # cut_files_per_primer contains Unique_File_name, we use unite() to match the DADA2 outputs
      tbl_Cutadapt <- cut_files_per_primer %>%
        dplyr::filter(Stage == "FWD_R1_cutadapt") %>%
        dplyr::mutate(`Primer detected reads (pairs)` = sapply(`Read file`, function(x) ShortRead::countFastq(x)[,1])) %>%
        tidyr::unite(col = "Unique_File_name_Primer",
                     Unique_File_name, Primer,
                     sep = "--",
                     remove = FALSE) %>%
        dplyr::select(Unique_File_name_Primer, Unique_File_name, Primer, `Primer detected reads (pairs)`)
      
      # 4. Extract DADA2 sequence steps
      tbl_Denoised_R1 <- tibble::enframe(sapply(all_dadaFs, getN),
                                         name = "Unique_File_name_Primer", 
                                         value = "Denoised (R1)")
      
      tbl_Denoised_R2 <- tibble::enframe(sapply(all_dadaRs, getN),
                                         name = "Unique_File_name_Primer", 
                                         value = "Denoised (R2)")
      
      tbl_Merged <- tibble::enframe(rowSums(mergers_seqtab),
                                         name = "Unique_File_name_Primer", 
                                         value = "Merged")
      
      tbl_NonChim <- tibble::enframe(rowSums(mergers_seqtab.nochim),
                                         name = "Unique_File_name_Primer", 
                                         value = "Non-chimeric Merged")
      
      tbl_NonChim_R1 <- tibble::enframe(rowSums(R1_seqtab.nochim), 
                                      name = "Unique_File_name_Primer", 
                                      value = "Non-chimeric R1")
      
      tbl_NonChim_R2 <- tibble::enframe(rowSums(R2_seqtab.nochim), 
                                      name = "Unique_File_name_Primer", 
                                      value = "Non-chimeric R2")
      
      # 5. Combine all tracking metrics into a single cohesive table
      all_track <- list(tbl_Cutadapt,
                        tbl_Denoised_R1,
                        tbl_Denoised_R2,
                        tbl_Merged,
                        tbl_NonChim,
                        tbl_NonChim_R1, 
                        tbl_NonChim_R2) %>%
        purrr::reduce(dplyr::left_join, by = "Unique_File_name_Primer") %>%
        dplyr::left_join(pre_filtered_summary, by = "Unique_File_name") %>%
        # Join with original configuration metadata
        dplyr::left_join(primers_n_samples, by = c("Unique_File_name", "Primer")) %>%
        # Reorder columns logically to match the sequence of events
        dplyr::relocate(Project, Sample, .after = Unique_File_name_Primer) %>%
        dplyr::relocate(`Raw reads (pairs)`, .before = `Primer detected reads (pairs)`)
      
      # Show preview
      all_track %>%
      dplyr::select(Unique_File_name, 
                    `Raw reads (pairs)`, 
                    `Primer detected reads (pairs)`,
                    `Pre-Filtered reads (pairs)`, 
                    `Denoised (R1)`,
                    `Denoised (R2)`,
                    `Non-chimeric Merged`,
                    `Non-chimeric R1`,
                    `Non-chimeric R2`) %>%
      View()
    }
    ```

-   **Save metrics to sheet.** Export the aggregated tracking table to individual spreadsheets for project-specific reporting and external auditing.

    ```{r}
    {
      for (PRJ in all_projects) {
        
        prj_track <- all_track %>%
          dplyr::filter(Project == PRJ)
        
        output_file <- file.path(tables_path, paste0(PRJ, "-reads_and_seqs_counts-",
                                                     Sys.Date(), ".csv"))
        
        readr::write_csv(prj_track, file = output_file)
        }

      message("✓ Saved read tracking metrics to datasheet file.")
    }
    ```

-   **Plot Read Retention.** Transforms the tabular data into a long format to generate faceted bar plots. These visual summaries quickly highlight whether data loss occurred uniformly across stages or if specific samples failed abruptly at a particular step. Plots are saved automatically as PDFs.

    ```{r}
    # Transform tibble to long format for ggplot
    track_tbl <- all_track %>%
      tidyr::gather(key = "Stage", 
                    value = "Read counts",
                    "Raw reads (pairs)",
                    "Primer detected reads (pairs)",
                    "Pre-Filtered reads (pairs)", 
                    "Denoised (R1)",
                    "Denoised (R2)",
                    "Merged",
                    "Non-chimeric Merged") %>%
      dplyr::mutate(Stage = factor(Stage, levels = c("Raw reads (pairs)",
                                                     "Primer detected reads (pairs)",
                                                     "Pre-Filtered reads (pairs)", 
                                                     "Denoised (R1)",
                                                     "Denoised (R2)",
                                                     "Merged",
                                                     "Non-chimeric Merged"))) 
      
    options(scipen = 22)

    # Define a clean Viridis palette explicitly mapping to our 7 stages
    stage_colors <- viridis::turbo(n = 7, direction = -1)

    for (PRJ in all_projects) {

      project_name <- stringr::str_replace_all(PRJ, " ", "_")
      
      # Filter for current project
      prj_data <- track_tbl %>% 
        dplyr::filter(Project == PRJ) %>%
        dplyr::arrange(Unique_File_name)

      N_samples <- length(unique(prj_data$Unique_File_name))

      track_plot <- ggplot2::ggplot(prj_data, aes(y = Stage,
                                                  x = `Read counts`, 
                                                  fill = Stage,
                                                  group = Unique_File_name)) +
        ggplot2::geom_bar(stat = "identity", alpha = 0.85) + 
        ggplot2::scale_fill_manual(values = stage_colors) +
        ggplot2::labs(title = paste0("Project: ", PRJ),
                      subtitle = "Number of sequences per library across data processing stages",
                      x = "Number of sequences",
                      y = "Data processing stage") +
        ggplot2::facet_wrap(Unique_File_name ~ Primer, ncol = 8) +
        ggplot2::theme_bw(base_size = 8) +
        ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, face = "bold"),
                       legend.position = "bottom",
                       strip.text = element_text(size = 7)) +
        # Dynamic coordinate ratio to prevent squished bars
        ggplot2::coord_fixed(ratio = max(prj_data$`Read counts`, na.rm = TRUE) / length(unique(prj_data$Stage)) * 0.75)

      # Save plot
      plot_file <- file.path(figs_path, paste0(project_name, "-samples_track.pdf"))
      
      ggplot2::ggsave(file = plot_file,
                      plot = track_plot,
                      device = "pdf",
                      width = ifelse(N_samples <= 10, 8, 12),
                      height = ifelse(N_samples <= 10, 4, N_samples / 6),
                      units = "cm", 
                      scale = 5, 
                      limitsize = FALSE,
                      dpi = 600)
                      
      message(paste0("  Saved tracking plot for ", PRJ, ": ", basename(plot_file)))
      
      # Open the plot
      utils::browseURL(plot_file)
    }
    ```

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "3_DADA2_worflow-", Sys.Date()))
    ```

## 4. Phyloseq Integration & Data Structuring

Once ASVs are inferred and chimeras removed, the sequence counts must be integrated with the experimental metadata. In this section, we use the Phyloseq package to merge these different data types into a cohesive structure. Subsequently, we 'melt' this object into a tidy, long-format dataframe, calculate relative abundances, and visualize the ASV length distribution to assess data integrity prior to taxonomic assignment.

### Prepare Sample Metadata

Before creating the Phyloseq object, the sample metadata table must be formatted so that its row names exactly match the sample IDs of the DADA2 ASV sequence table. This code extracts the relevant columns from the configuration table, expands rows for samples with multiple primers, and validates the match between the two tables.

-   **Format metadata table.** Prepare the sample metadata by organizing identifiers and ensuring row names exactly match the sample names in the sequence table, a requirement for Phyloseq.

    ```{r}
    # Create sample metadata dataframe for Phylose
    # Attention! Add other metadata columns if needed
    all_samdf <- primers_n_samples %>% 
      dplyr::select("Project",
                    "Sample",
                    # "Researcher",
                    # "Analysis",
                    "Unique_File_name",
                    "Primer",
                    "Lib",
                    "Type",
                    # starts_with("Metadata"),
                    # "obs",
                    # "Extraction control",
                    # "Filtration control",
                    # "PCR control"
                    ) %>% 
      BiocGenerics::unique() %>%
      # Expand rows if multiple primers exist per sample (semi-colon separated)
      tidyr::separate_longer_delim(cols = Primer, delim = ";") %>%
      # Create unique identifier combining file name and primer
      tidyr::unite(col = "Unique_File_name_Primer",
                   Unique_File_name, 
                   Primer,
                   sep = "--",
                   remove = FALSE) %>%
      as.data.frame()

    # Assign row names to match ASV table dimensions
    rownames(all_samdf) <- all_samdf$Unique_File_name_Primer
    ```

-   **Check sample consistency.** Verify if all sequences in the ASV table have matching metadata in the newly created sample dataframe.

    ```{r}
    # Check consistency between sequence table and metadata
    seq_names <- rownames(mergers_seqtab.nochim)
    meta_names <- rownames(all_samdf)

    if (!all(seq_names %in% meta_names)) {
      warning("Caution: Some samples in the sequence table are missing from metadata!")
      print(setdiff(seq_names, meta_names))
    } else {
      message("All samples matched correctly between Sequence Table and Metadata.")
    }
    ```

### Bulid and format the Phyloseq Object

With validated metadata in hand, the ASV abundance table and sample data are combined into a single Phyloseq object. That object is then melted into a long-format tibble, column names are standardized, and relative abundances are calculated to allow fair comparisons across samples with unequal sequencing depths.

-   **Import DADA2 results.** Combine the ASV abundance table and the formatted sample metadata into a unified Phyloseq object.

    ```{r}
    # Create Phyloseq object from DADA2 merged results
    mergers_ps <- phyloseq::phyloseq(phyloseq::otu_table(mergers_seqtab.nochim,
                                                         taxa_are_rows = FALSE),
                                     phyloseq::sample_data(all_samdf))

    #  Unmerged workflow
    # Create Phyloseq object from DADA2 unmerged R1 reads
    R1_ps <- phyloseq::phyloseq(phyloseq::otu_table(R1_seqtab.nochim,
                                                    taxa_are_rows = FALSE),
                                phyloseq::sample_data(all_samdf))

    # Create Phyloseq object from DADA2 unmerged R1 reads
    R2_ps <- phyloseq::phyloseq(phyloseq::otu_table(R2_seqtab.nochim,
                                                    taxa_are_rows = FALSE),
                                phyloseq::sample_data(all_samdf))
    ```

-   **Transform to tidy format.** Convert the Phyloseq object into a long-format dataframe (`psmelt`), filtering out zero-abundance ASVs to optimize memory usage and downstream processing.

    ```{r}
    # Convert Phyloseq object to long tibble
    # For merged data
    mergers_ps_tbl <- phyloseq::psmelt(mergers_ps) %>%
      tibble::as_tibble() %>%
      dplyr::mutate(`Read origin` = "merged") %>%
      dplyr::filter(Abundance > 0) %>% # Filter strictly positive abundance
      dplyr::rename(`ASV (sequence)` = OTU) # Keep ASV sequence as identifier

    # Unmerged workflow
    # R1 reads data
    # R1_ps_tbl <- phyloseq::psmelt(R1_ps) %>%
    #   as_tibble() %>%
    #   mutate(`Read origin` = "R1") %>%
    #   filter(Abundance >=1) %>%
    #   mutate("ASV tip" = OTU) %>%
    #   dplyr::rename("ASV" = "OTU")

    # R2 reads data
    # R2_ps_tbl <- phyloseq::psmelt(R2_ps) %>%
    #   as_tibble() %>%
    #   mutate(`Read origin` = "R2") %>%
    #   filter(Abundance >=1) %>%
    #   mutate("ASV tip" = OTU) %>%
    #   dplyr::rename("ASV" = "OTU")
    ```

-   **Standardize combined table.** Standardize column names and calculate the length of each ASV for plotting.

    ```{r}
    # Combine all processed tables (if multiple sources exist)
    all_ps_tbl <- mergers_ps_tbl %>%
      # dplyr::bind_rows(R1_ps_tbl, R2_ps_tbl) %>% 
      dplyr::select(-Sample) %>%  # Remove duplicate/confusing columns if necessary
      dplyr::rename(Sample = sample_Sample) %>% # Standardize Sample column name
      dplyr::mutate(`ASV Size (bp)` = nchar(`ASV (sequence)`))

    # Show table
    View(all_ps_tbl)
    ```

-   **Compute relative abundances.** Calculate the proportion of each ASV relative to the sample's total library size and the overall dataset, enabling fair comparisons across libraries with unequal sequencing depths.

    ```{r}
    # Calculate relative abundances per sample and dataset
    all_ps_tbl <- all_ps_tbl %>%
      # Calculate Total Abundance of the entire dataset once
      dplyr::mutate(Dataset_Total_Abundance = sum(Abundance)) %>%
      # Group by sample context to calculate sample-specific totals
      dplyr::group_by(Project, 
                      Unique_File_name, 
                      Primer, 
                      `Read origin`) %>%
      dplyr::mutate(Sample_Total_Abundance = sum(Abundance),
                    Relative_Abundance_Dataset = round((Abundance /
                                                        Dataset_Total_Abundance),
                                                        digits = 6),
        Relative_Abundance_Sample  = round((Abundance / 
                                            Sample_Total_Abundance), 
                                            digits = 6)) %>%
      dplyr::ungroup() %>%
      # Clean up temporary total column
      dplyr::select(-Dataset_Total_Abundance) %>%
      dplyr::relocate(Sample_Total_Abundance, 
                      Relative_Abundance_Dataset, 
                      Relative_Abundance_Sample, 
                      .after = Abundance)

    # Show table
    View(all_ps_tbl)
    ```

### Check ASV Length Distribution

Before proceeding to taxonomic assignment, it is important to verify that the inferred ASVs fall within the expected size range for the target amplicon. The plot below displays ASV lengths across all samples, coloured by read origin and scaled by relative abundance, making it straightforward to spot outliers, unexpected size classes, or potential non-target amplicons.

-   **Visualize ASV lengths.** Plot the distribution of ASV sequence lengths prior to BLASTn identification. This diagnostic step helps flag unexpected size distributions or potential non-target amplicons.

    ```{r}
    # Plot ASV length distribution by sample
    asv_length_plot <- all_ps_tbl %>%
      ggplot2::ggplot(ggplot2::aes(y = interaction(`Read origin`, 
                                                   Unique_File_name),
                                   x = `ASV Size (bp)`,
                                   fill = `Read origin`,
                                   col = `Read origin`,
                                   size = Relative_Abundance_Sample,
                                   alpha = 0.25)) +
      ggplot2::geom_jitter(height = 0.3, width = 0.3) +
      # ggplot2::scale_x_continuous(breaks = seq(20, 
      #                                          max(all_ps_tbl$`ASV Size (bp)`, 
      #                                              na.rm = TRUE), 10),
      #                             expand = c(0.02, 0.02)) +
      ggplot2::scale_fill_manual(values = viridis::turbo(n = 12)[c(3, 6, 9, 12)]) +
      ggplot2::scale_colour_manual(values = viridis::turbo(n = 12)[c(3, 6, 9, 12)]) +
      ggplot2::labs(x = "ASV Size (bp)",
                    y = "Sample",
                    title = "ASV Length Distribution by Sample",
                    subtitle = "Size and Read Origin per Sample (all inferred ASVs)") +
      ggplot2::theme_bw(base_size = 8) +
      ggplot2::theme(legend.position = "right",
                     axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                     strip.text.y = ggplot2::element_text(size = 10, angle = 0)) +
      # ggplot2::geom_vline(xintercept = c(10, max(all_ps_tbl$`ASV Size (bp)`, 
      #                                            na.rm = TRUE))) +
      ggplot2::facet_grid(rows = ggplot2::vars(Project, 
                                               Type),
                          cols = ggplot2::vars(Primer),
                          scales = "free_y",
                          space = "free_y")

    # Save plot
    ggplot2::ggsave(filename = file.path(figs_path, 
                                         "ASV_length_distribution-PreBLAST.pdf"),
                    plot = asv_length_plot,
                    device = "pdf",
                    width = 25,
                    height = 30,
                    units = "cm",
                    dpi = 600)

    # Open the plot
      utils::browseURL(file.path(figs_path, "ASV_length_distribution-PreBLAST.pdf"))
    ```

### Save the Environment

-   **Store the objects.** Save all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "4_phyloseq_integration-", Sys.Date(), ".RData"))
    ```

## 5. Taxonomic Assignment

Once the ASV table is validated and integrated into the Phyloseq object, each unique sequence must be assigned a taxonomic identity. This section performs that identification using BLASTn against one or more local nucleotide databases. Because the number of ASVs can be very large, queries are split into batches and executed in parallel using the BLASTr package. A correction step is also applied for custom databases (such as LGC12Sdb) that may lack standard NCBI TaxIDs.

### Prepare and Execute BLASTr

-   **Select unique ASVs.** This step extracts a deduplicated list of sequences, which is then split into manageable chunks for batch processing.

    ```{r}
    # Extract unique ASV sequences for BLAST submission
    asvs_blast_all <- all_ps_tbl %>%
      dplyr::pull(`ASV (sequence)`) %>% 
      unique() %>% 
      stats::na.omit() %>% 
      as.character()
    ```

-   **Run Parallel BLAST.** Execute BLASTn in batches using the `BLASTr` package to handle large datasets efficiently without overloading memory.

    ```{r}
    # Define chunk size for batch processing (e.g., 20,000 sequences per batch)
    chunk_size  <- 50
    asv_chunks  <- split(asvs_blast_all, ceiling(seq_along(asvs_blast_all) / chunk_size))

    # Define local database paths
    db_paths <- "/data/databases/core_nt/core_nt /data/databases/LGC12Sdb/mai25/LGC12Sdb_complete_noGaps-2025-05-21.fasta"

    # Execute BLASTn across all chunks
    blast_results_list <- list()
    {
      tictoc::tic("Total BLASTn taxonomy assignment")
      
      for (i in seq_along(asv_chunks)) {

      message("Processing BLAST chunk ", i, " of ", length(asv_chunks))

      blast_results_list[[i]] <- BLASTr::parallel_blast(
        asv_chunks[[i]],
        db_path = db_paths,
        out_file = file.path(blast_path, paste0("blast_out_res_", i, ".csv")),
        out_RDS = file.path(blast_path, paste0("blast_out_res_", i, ".RDS")),
        total_cores = 60,       # Adjust based on server capacity
        num_threads = 2,
        blast_type = "blastn",
        perc_id = 80,
        perc_qcov_hsp = 80,
        num_alignments = 3
        )

      # Optional: Save intermediate environment after each chunk
      # base::save.image(paste0(env_path, "/", "5_BLASTr-chunk", i, "-", Sys.Date(), ".RData"))
      }
      tictoc::toc()
    }
    ```

-   **Combine BLAST results.** Merge all chunk outputs into a single dataframe, removing rows where no subject was matched.

    ```{r}
    blast_res_full <- dplyr::bind_rows(blast_results_list) %>%
      dplyr::filter(!is.na(`1_subject header`)) %>%
      dplyr::rename(`ASV (Sequence)` = Sequence)

    message(nrow(blast_res_full), " BLAST hits retained across ", 
            length(unique(blast_res_full$`ASV (Sequence)`)), " unique ASVs.")  
    ```

### Correct Database TaxIDs

-   **Fill missing TaxID.** For hits lacking TaxIDs across all databases, extract genus/species names from subject headers and perform parallel NCBI lookups using `myTAI::taxonomy`, updating `1_staxid`, `2_staxid`, and `3_staxid` columns with success rates reported per hit.

    ```{r}
    # Helper function to retrieve TaxID from organism name
    get_taxid <- function(organism_name) {
      result <- tryCatch({
        myTAI::taxonomy(organism = organism_name,
                        db = "ncbi",
                        output = "taxid"
                        )$id[1]
        }, error = function(e) NA_character_)
      
      return(result)
    }

    # List of columns to process: (hit_number, staxid_column, subject_header_column, DB_column)
    hit_columns <- list(c(1, "1_staxid", "1_subject header", "1_DB"),
                        c(2, "2_staxid", "2_subject header", "2_DB"),
                        c(3, "3_staxid", "3_subject header", "3_DB"))

    # Process each hit (1st, 2nd, and 3rd)
    for (hit_info in hit_columns) {
      hit_num <- hit_info[1]
      staxid_col <- hit_info[2]
      subject_header_col <- hit_info[3]
      db_col <- hit_info[4]
      
      message("Processing TaxID lookup for hit ", hit_num, " (", staxid_col, ")")
      
      # Identify rows where TaxID is missing (NA or "N/A")
      missing_taxid_mask <- is.na(blast_res_full[[staxid_col]]) | 
                            blast_res_full[[staxid_col]] == "N/A"
      
      if (!any(missing_taxid_mask)) {
        message("  No missing TaxIDs found for ", staxid_col)
        next
      }
      
      # Filter only rows with missing TaxID
      missing_rows <- blast_res_full[missing_taxid_mask, ]
      
      message("  Found ", sum(missing_taxid_mask), " missing TaxIDs to lookup")
      
      # Extract organism_name from first word of subject header
      organism_names <- stringr::word(missing_rows[[subject_header_col]], 1)
      
      # Lookup TaxIDs in parallel (uses furrr for acceleration)
      taxids <- furrr::future_map_chr(
        organism_names,
        get_taxid,
        .options = furrr::furrr_options(seed = TRUE)
      )
      
      # Update only rows that had missing TaxID
      blast_res_full[[staxid_col]][missing_taxid_mask] <- taxids
      
      # Success report
      success_rate <- mean(!is.na(taxids)) * 100
      message("  Success rate: ", round(success_rate, 1), "% (", 
              sum(!is.na(taxids)), "/", length(taxids), " lookups)")
    }

    # Final summary
    message("\nFinal TaxID coverage:")
    message("Hit 1: ", round(mean(!is.na(blast_res_full$`1_staxid`)) * 100, 1), "%")
    message("Hit 2: ", round(mean(!is.na(blast_res_full$`2_staxid`)) * 100, 1), "%")
    message("Hit 3: ", round(mean(!is.na(blast_res_full$`3_staxid`)) * 100, 1), "%")
    ```

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "5_Taxonomic_assignment-", Sys.Date()))
    ```

## 6. Taxonomy Curation & Assignment

This section refines the raw BLAST results, retrieves complete lineage information from NCBI, and standardizes taxonomic ranks.

### Refine BLAST Hits

-   **Filter low-quality hits.** Remove uninformative descriptions (e.g., "Uncultured") and select the best reliable hit among the top 3 alignments.

    ```{r, echo=TRUE, eval=FALSE}
    # Define patterns for non-informative hits
    bad_res_ids <- paste0(c(
      "Uncultured", "Uncultured organism", "Uncultured prokaryote",
      "Eukaryotic synthetic construct", "16S rRNA amplicon fragment",
      "Invertebrate environmental", "Uncultured Candidatus",
      "Uncultured bacterium", "Uncultured archaeon",
      "Complete Metagenome-Assembled", "PREDICTED: Nomascus"
    ), collapse = "|")

    # Vectorized selection of best hit (1st, 2nd, or 3rd)
    blast_res_processed <- blast_res_full %>%
      dplyr::mutate(
    # Check reliability of each hit
    is_bad_1 = stringr::str_detect(`1_subject header`, bad_res_ids),
    is_bad_2 = stringr::str_detect(`2_subject header`, bad_res_ids),
    is_bad_3 = stringr::str_detect(`3_subject header`, bad_res_ids),

    # Select best valid hit
    Selected_Hit_Origin = dplyr::case_when(
      !is_bad_1 ~ "1",
      is_bad_1 & !is.na(`2_subject header`) & !is_bad_2 ~ "2",
      is_bad_1 & is_bad_2 & !is.na(`3_subject header`) & !is_bad_3 ~ "3",
      TRUE ~ "None"
    ),

    # Assign Final ID and TaxID based on selection
    `blast ID` = dplyr::case_when(
      Selected_Hit_Origin == "1" ~ substr(`1_subject header`, 1, 40),
      Selected_Hit_Origin == "2" ~ substr(`2_subject header`, 1, 40),
      Selected_Hit_Origin == "3" ~ substr(`3_subject header`, 1, 40),
      TRUE ~ "Match_not_reliable"
    ),

    query_taxID = dplyr::case_when(
      Selected_Hit_Origin == "1" ~ `1_staxid`,
      Selected_Hit_Origin == "2" ~ `2_staxid`,
      Selected_Hit_Origin == "3" ~ `3_staxid`,
      TRUE ~ NA_character_
    )
      ) %>%
      # Second pass clean-up for any lingering "Uncultured" that slipped through
      dplyr::mutate(
    `blast ID` = dplyr::if_else(
      stringr::str_detect(`blast ID`, "ncultur|nvironmental"),
      "Match_not_reliable",
      `blast ID`
    )
      )

    # Check results
    # table(blast_res_processed$Selected_Hit_Origin)
    ```

### Clean Species Names

-   **Standardize nomenclature.** Clean up species names by removing artifacts, tags (e.g., "cf.", "aff."), and reducing names to Binomial format (Genus species).

    ```{r, echo=TRUE, eval=FALSE}
    # Clean strings
    blast_res_processed <- blast_res_processed %>%
      dplyr::mutate(
    `blast ID` = `blast ID` %>%
      stringr::str_replace(": ", ":") %>%
      stringr::str_remove("\\n$") %>%
      stringr::str_remove("^(  | |Uncultured |uncultured |Candidatus |MAG:|TPA_asm:|^Cf. |candidate division )") %>%
      stringr::str_remove_all("\\[|\\]") %>%
      stringr::str_replace("cf\\. ", "") %>%
      stringr::str_replace("nr\\. ", "") %>%
      stringr::str_replace(",", "") %>%
      # Extract only the first two words (Genus species)
      stringr::word(1, 2) %>%
      # Handle specific manual corrections
      dplyr::case_match(
        "Human DNA" ~ "Homo sapiens",
        "Human chromosome" ~ "Homo sapiens",
        .default = .
      )
      )

    # Verify cleaned names
    # head(unique(blast_res_processed$`blast ID`))
    ```

### Retrieve Lineage from NCBI

-   **Fetch complete taxonomy.** Query NCBI using the valid TaxIDs to retrieve full lineage information (Kingdom to Species).

    ```{r, echo=TRUE, eval=FALSE}
    # Identify unique TaxIDs to query
    taxids_to_search <- unique(na.omit(blast_res_processed$query_taxID))

    # Execute parallel taxonomy retrieval
    # Note: Ensure BLASTr is loaded. Using 'parallel_get_tax' from the package.
    # devtools::load_all("~/prjcts/BLASTr") # Only if local dev version needed

    taxonomy_tbl <- BLASTr::parallel_get_tax(
      organisms_taxIDs = taxids_to_search,
      retry_times = 5,
      verbose = "full"
    ) %>% 
      BiocGenerics::unique()

    # Check for missing IDs (failed queries)
    missing_taxids <- setdiff(taxids_to_search, taxonomy_tbl$query_taxID)
    if (length(missing_taxids) > 0) {
      message(paste(length(missing_taxids), "TaxIDs could not be retrieved."))
    }
    ```

### Fill Missing Ranks

-   **Impute missing levels.** Fill gaps in the taxonomic tree (e.g., missing Family) by propagating information from higher ranks with a prefix (e.g., "family of Order X").

    ```{r, echo=TRUE, eval=FALSE}
    # Function to fill missing ranks iteratively
    fill_rank <- function(df, target_rank, source_rank, prefix) {
      df[[target_rank]] <- ifelse(
    is.na(df[[target_rank]]) | df[[target_rank]] == "",
    paste0(prefix, df[[source_rank]]),
    df[[target_rank]]
      )
      return(df)
    }

    # Apply filling logic
    taxonomy_tbl_filled <- taxonomy_tbl %>%
      # Manual Superkingdom corrections based on Kingdom/Phylum
      dplyr::mutate(
    `Superkingdom (NCBI)` = dplyr::case_when(
      is.na(`Superkingdom (NCBI)`) & `Kingdom (NCBI)` %in% c("Metazoa", "Fungi", "Viridiplantae") ~ "Eukaryota",
      is.na(`Superkingdom (NCBI)`) & stringr::str_detect(`Phylum (NCBI)`, "bacteria$") ~ "Bacteria",
      TRUE ~ `Superkingdom (NCBI)`
    )
      ) %>%
      # Fill NAs sequentially (top-down)
      fill_rank("Superkingdom (NCBI)", "Kingdom (NCBI)", "superkingdom of ") %>%
      fill_rank("Kingdom (NCBI)", "Superkingdom (NCBI)", "kingdom of ") %>%
      fill_rank("Phylum (NCBI)", "Kingdom (NCBI)", "phylum of ") %>%
      fill_rank("Class (NCBI)", "Phylum (NCBI)", "class of ") %>%
      fill_rank("Order (NCBI)", "Class (NCBI)", "order of ") %>%
      fill_rank("Family (NCBI)", "Order (NCBI)", "family of ") %>%
      fill_rank("Genus (NCBI)", "Family (NCBI)", "genus of ")

    # Merge taxonomy back to BLAST results
    blast_res_tax <- dplyr::left_join(
      blast_res_processed,
      taxonomy_tbl_filled,
      by = "query_taxID"
    )

    # Join back to Main ASV Table
    all_ps_tbl_final <- dplyr::left_join(
      all_ps_tbl,
      blast_res_tax %>% dplyr::select(ASV_tip, `blast ID`, `Superkingdom (NCBI)`:`Species (NCBI)`),
      by = "ASV_tip"
    ) %>%
      dplyr::mutate(`Final ID` = `blast ID`) %>%
      dplyr::filter(Abundance > 0)
    ```

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "6_Taxonomic_curation-", Sys.Date()))
    ```

## 7. Contamination Analysis

This section identifies potential contaminant ASVs by comparing their abundance in samples against their abundance in the negative controls (PCR, extraction, filtration).

### Identify Contaminant ASVs

-   **Create a contaminant reference.** Isolate all ASVs present in control samples to build a reference table of potential contaminants and their maximum relative abundance within those controls.

    ```{r, echo=TRUE, eval=FALSE}
    # Define keywords for control sample types
    control_types <- c(
      "Extraction control", "Extraction Control", "Ext. control", "Ext. Control",
      "Filt. Control", "PCR control", "PCR Control", "Filt. control",
      "Filtration control", "Filtration Control", "Negative control",
      "Negative Control", "Positive control", "Positive Control", "Control"
    )

    # Create a lookup table of ASVs found in any control sample
    all_contam_asvs <- all_ps_tbl_final %>%
      dplyr::filter(Type %in% control_types) %>%
      dplyr::group_by(`ASV (Sequence)`, Unique_File_name) %>%
      # Find the max abundance of each ASV in each specific control well
      dplyrsummarise(`Max. ASV abd. in control` = max(`Relative abundance on sample`, na.rm = TRUE), .groups = "drop") %>%
      # Join back to get the control Type (PCR, Extraction, etc.)
      dplyr::left_join(
    all_ps_tbl_final %>% dplyr::select(Unique_File_name, Type, `Final ID`),
    by = "Unique_File_name"
      ) %>%
      dplyr::distinct()

    # Check the results
    # head(all_contam_asvs)
    # unique(all_contam_asvs$Unique_File_name)
    ```

### Flag Potential Contaminants in Samples

-   **Compare sample vs. control abundance.** For each ASV in a true sample, compare its abundance to its maximum observed abundance in the corresponding negative controls. ASVs that are not significantly more abundant in the sample than in the control are flagged as potential contaminants.

    ```{r, echo=TRUE, eval=FALSE}
    # This vectorized approach replaces the slow row-by-row 'for' loop.
    # It works by reshaping the data to link each sample-ASV pair to its controls.

    # 1. Prepare a long-format map of samples to their specific controls
    sample_to_control_map <- all_ps_tbl_final %>%
      dplyr::filter(!Type %in% control_types) %>%
      dplyr::select(
    Unique_File_name, `ASV (Sequence)`, `Relative abundance on sample`,
    `PCR control`, `Extraction control`, `Filtration control`
      ) %>%
      tidyr::pivot_longer(
    cols = c(`PCR control`, `Extraction control`, `Filtration control`),
    names_to = "Control_Type_Source",
    values_to = "Control_Name_Raw"
      ) %>%
      # Expand rows for semicolon-separated control names
      tidyr::separate_longer_delim(Control_Name_Raw, delim = ";")

    # 2. Create a clean lookup table for control abundances
    control_abundance_lookup <- all_contam_asvs %>%
      dplyr::select(
    Control_Name_Raw = Unique_File_name,
    `ASV (Sequence)`,
    `Max. ASV abd. in control`
      )

    # 3. Join the sample map with the control abundances
    contamination_check_tbl <- sample_to_control_map %>%
      dplyr::left_join(control_abundance_lookup, by = c("Control_Name_Raw", "ASV (Sequence)")) %>%
      # For each sample-ASV, find the highest contamination signal from any of its controls
      dplyr::group_by(Unique_File_name, `ASV (Sequence)`) %>%
      dplyr::summarise(
    max_control_abd = max(`Max. ASV abd. in control`, na.rm = TRUE),
    sample_abd = dplyr::first(`Relative abundance on sample`),
    .groups = "drop"
      ) %>%
      # Calculate fold change (handle cases where control abundance is 0 or NA)
      dplyr::mutate(
    fold_change = gtools::foldchange(sample_abd, max_control_abd),
    `Contamination status` = dplyr::if_else(
      fold_change > 1, # Rule: Must be more abundant in sample than control
      "True Detection",
      "Possible Contamination"
    )
      )

    # 4. Join contamination status back to the main table
    all_ps_tbl_final <- all_ps_tbl_final %>%
      dplyr::left_join(
    contamination_check_tbl %>% dplyr::select(`ASV (Sequence)`, Unique_File_name, `Contamination status`),
    by = c("ASV (Sequence)", "Unique_File_name")
      ) %>%
      # For controls themselves, status is irrelevant; for samples not in the check_tbl, they are True Detections
      dplyr::mutate(
    `Contamination status` = dplyr::coalesce(`Contamination status`, "True Detection")
      )

    # Check results
    # table(all_ps_tbl_final$`Contamination status`)
    ```

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "7_Contamination_analysis-", Sys.Date()))
    ```

## 8. Final Results Visualization

This section generates plots to visualize the distribution of all identified ASVs, their taxonomic assignment, size, and relative abundance across samples.

### Plot ASV Distributions by Project

-   **Generate overview plots.** For each project, create a comprehensive plot showing every ASV, colored by BLAST score and shaped by superkingdom. This allows for a high-level assessment of the taxonomic landscape and data quality.

    ```{r, echo=TRUE, eval=FALSE}
    # Define shapes for taxonomic groups
    SHAPES <- c("Archaea" = 25, "Bacteria" = 24, "Eukaryota" = 21, "Viruses" = 23)

    # Loop through each project to create and save plots
    for (PRJ in all_projects) {

      message(paste0("Plotting project: ", PRJ))

      # Prepare data for plotting (calculate pseudo-score)
      plot_data <- all_ps_tbl_final %>%
    dplyr::filter(Project == PRJ) %>%
    dplyr::mutate(
      `BLASTn pseudo-score` = ((`1_indentity` * (`1_qcovhsp` / 10)) / 10),
      # Nullify score for uncultured hits
      `BLASTn pseudo-score` = dplyr::case_when(
        stringr::str_detect(`1_subject header`, "ncult") ~ NA_real_,
        TRUE ~ `BLASTn pseudo-score`
      )
    ) %>%
    dplyr::arrange(dplyr::desc(`BLASTn pseudo-score`))

      # Build the plot
      asv_dist_plot <- ggplot2::ggplot(
    plot_data,
    aes(
      y = interaction(Sample, Unique_File_name, sep = " - "),
      x = `ASV Size (pb)`,
      fill = `BLASTn pseudo-score`,
      col = `BLASTn pseudo-score`,
      shape = `Superkingdom (NCBI)`,
      size = `Relative abundance on sample`,
      alpha = 0.2,
      text = paste0(
        "Final ID: ", `Final ID`, "<br>",
        "Sample: ", Sample, "<br>",
        "ASV Size (bp): ", `ASV Size (bp)`, "<br>",
        "Pseudo-score: ", round(`BLASTn pseudo-score`, 2)
      )
    )
      ) +
      ggplot2::geom_jitter(height = 0.4, width = 0.25) +
      ggplot2::scale_fill_gradientn(
    name = "BLASTn pseudo-score (%)", na.value = "grey95",
    colours = blast_score_palette,
    limits = c(60, 100), breaks = seq(60, 100, 5)
      ) +
      ggplot2::scale_color_gradientn(guide = "none", na.value = "grey80", colours = c("#330000", "darkred", "red", "yellow", "green", "darkgreen"), limits = c(60, 100)) +
      ggplot2::scale_size_continuous(name = "Relative Abundance (%)") +
      ggplot2::scale_shape_manual(name = "Superkingdom (NCBI)", values = SHAPES, na.value = 22) +
      ggplot2::labs(x = "ASV Size (bp)", y = "Sample", title = paste("EcoMol -", PRJ)) +
      ggplot2::theme_bw(base_size = 14) +
      ggplot2::theme(legend.position = "bottom", legend.key.width = unit(2, 'cm')) +
      ggplot2::facet_grid(rows = vars(`Read origin`, Primer), scales = "free_y", space = "free_y")

      # Save static plot
      n_samples_in_project <- length(unique(plot_data$Unique_File_name))
      ggsave(
    filename = file.path(results_path, paste0(PRJ, "-ASV_length_by_sample--All_ASVs.pdf")),
    plot = asv_dist_plot,
    device = "pdf",
    width = 40,
    height = round((n_samples_in_project * 0.5) + 5),
    units = "cm",
    limitsize = FALSE,
    dpi = 300
      )

      # Save interactive plot
      asv_dist_plotly <- plotly::ggplotly(asv_dist_plot, tooltip = "text")
      htmlwidgets::saveWidget(
    widget = asv_dist_plotly,
    selfcontained = TRUE,
    file = file.path(results_path, paste0(PRJ, "-ASV_length_by_sample--All_ASVs.html"))
      )
    }
    ```

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "8_Final_visualization-", Sys.Date()))
    ```

## 9. Final Results Curation

This section refines the identification confidence levels and categorizes organisms into functional groups (e.g., Metazoa) for downstream ecological analysis.

### Assign Identification Confidence

-   **Calculate confidence scores.** Compute a pseudo-score based on BLAST identity and coverage, and assign taxonomic ranks (Species, Genus, Family, etc.) based on standard percentage thresholds.

    ```{r, echo=TRUE, eval=FALSE}
    # Calculate scores and assign taxonomic resolution
    all_ps_tbl_final <- all_ps_tbl_final %>%
      dplyr::mutate(
    `Curated ID` = `Final ID`,
    # Weighted score: heavily weights identity (10x) over coverage
    `BLASTn pseudo-score` = ((10 * `1_indentity`) + `1_qcovhsp`) / 11,

    # Assign resolution level based on score thresholds
    Identification = dplyr::case_when(
      `BLASTn pseudo-score` >= 98 ~ `Final ID`, 
      `BLASTn pseudo-score` >= 95 ~ `Genus (NCBI)`, 
      `BLASTn pseudo-score` >= 90 ~ `Family (NCBI)`, 
      `BLASTn pseudo-score` >= 80 ~ `Order (NCBI)`, 
      `BLASTn pseudo-score` >= 60 ~ `Class (NCBI)`,
      TRUE ~ "Unidentified"
    ),

    `Identification Max. taxonomy` = dplyr::case_when(
      `BLASTn pseudo-score` >= 98 ~ "Species",
      `BLASTn pseudo-score` >= 95 ~ "Genus",
      `BLASTn pseudo-score` >= 90 ~ "Family",
      `BLASTn pseudo-score` >= 80 ~ "Order",
      `BLASTn pseudo-score` >= 60 ~ "Class",
      TRUE ~ "Unidentified"
    )
      )
    ```

### Mark Metazoan Taxa

-   **Categorize Metazoans.** Flag ASVs belonging to Metazoa or other target groups (Zooplankton, Phytoplankton, Benthos) based on their taxonomy. This facilitates filtering for specific ecological guilds.

    ```{r, echo=TRUE, eval=FALSE}
    # 1. Define Biological Groups Lists (Expert Curation)
    # Benthos
    bentos_phy <- c("Annelida", "Arthropoda", "Mollusca")
    bentos_cls <- c("Bivalvia", "Clitellata", "Gastropoda", "Insecta")

    # Zooplankton
    zplnk_phy <- c("Amoebozoa", "Arthropoda", "Cercozoa", "Ciliophora", "Rotifera")
    zplnk_cls <- c("Branchiopoda", "Eurotatoria", "Filosia", "Hexanauplia", "Lobosa", "Oligohymenophorea")

    # Periphyton
    prftn_phy <- c("Bacillariophyta", "Charophyta", "Chlorophyta", "Cryptophyta", "Cyanobacteria", "Euglenophyta", "Rhodophyta")
    prftn_cls <- c("Bacillariophyceae", "Chlorophyceae", "Chrysophyceae", "Coscinodiscophyceae", "Cryptophyceae", "Cyanophyceae", "Euglenophyceae", "Florideophyceae", "Mediophyceae", "Ulvophyceae", "Zygnematophyceae")

    # Phytoplankton (Comprehensive list)
    ftplnk_phy <- c(
      "Bacillariophyta", "Charophyta", "Chlorophyta", "Cryptophyta", "Cyanobacteria", "Dinophyta", "Euglenophyta", "Ochrophyta",
      "Apicomplexa", "Cercozoa", "Ciliophora", "Discosea", "Endomyxa", "Euglenozoa", "Evosea", "Foraminifera", "Fornicata",
      "Haptophyta", "Hemimastigophora", "Heterolobosea", "Nebulidia", "Nibbleridia", "Oomycota", "Parabasalia", "Perkinsozoa",
      "Preaxostyla", "Rhodophyta", "Tubulinea"
    )
    ftplnk_cls <- c(
      "Bacillariophyceae", "Chlorophyceae", "Chrysophyceae", "Coscinodiscophyceae", "Cryptophyceae", "Cyanophyceae",
      "Dinophyceae", "Euglenophyceae", "Mediophyceae", "Trebouxiophyceae", "Ulvophyceae", "Zygnematophyceae",
      "Actinophryidae", "Aphelidea", "Bigyra", "Bolidophyceae", "Breviatea", "Centroplasthelida", "Choanoflagellata",
      "Developea", "Dictyochophyceae", "Eustigmatophyceae", "Filasterea", "Glaucocystophyceae", "Hyphochytriomycetes",
      "Ichthyosporea", "Olisthodiscophyceae", "Pelagophyceae", "Phaeophyceae", "Phaeothamniophyceae", "Raphidophyceae",
      "Synurophyceae", "Xanthophyceae"
    )

    # Combine all target classes and phyla for quick lookup
    all_target_phy <- unique(c(bentos_phy, zplnk_phy, prftn_phy, ftplnk_phy))
    all_target_cls <- unique(c(bentos_cls, zplnk_cls, prftn_cls, ftplnk_cls))

    # 2. Apply Logic to Mark Metazoa/Targets
    all_ps_tbl_final <- all_ps_tbl_final %>%
      dplyr::mutate(
    `Possible Metazoa` = dplyr::case_when(
      # Exclusions (Hard filters)
      stringr::str_detect(`blast ID`, "Chordate environmental") ~ FALSE,
      stringr::str_detect(`blast ID`, "Match_not_reliable") ~ FALSE,
      stringr::str_detect(`Superkingdom (NCBI)`, "Bacteria|Archaea|Viruses") ~ FALSE, # Note: Viruses added
      is.na(`Kingdom (NCBI)`) ~ FALSE,
      
      # Inclusions (Target Groups)
      stringr::str_detect(`Superkingdom (NCBI)`, "Eukaryota") & (`Phylum (NCBI)` %in% all_target_phy) ~ TRUE,
      stringr::str_detect(`Superkingdom (NCBI)`, "Eukaryota") & (`Class (NCBI)` %in% all_target_cls) ~ TRUE,
      
      # General Metazoa/Viridiplantae catch-all (if not excluded above)
      stringr::str_detect(`Kingdom (NCBI)`, "Metazoa|Viridiplantae") ~ TRUE,
      
      # Default
      TRUE ~ FALSE
    )
      )

    # Check distribution
    # table(all_ps_tbl_final$`Possible Metazoa`, useNA = "always")
    ```

### Final Table Formatting

-   **Reorganize columns.** Select and rename columns to produce a clean, user-friendly output table for reporting.

    ```{r, echo=TRUE, eval=FALSE}
    # Create final clean table
    results_table_clean <- all_ps_tbl_final %>%
      # Rename control columns for clarity
      dplyr::rename(
    `PCR Control` = `PCR control`,
    `Ext. Control` = `Extraction control`,
    `Filt. Control` = `Filtration control`
      ) %>%
      # Select and order essential columns
      dplyr::select(
    Researcher, Project, Primer, Sample, Unique_File_name,
    `Read origin`, `Relative abundance on sample`, Abundance,
    `BLASTn pseudo-score`, Identification, `Curated ID`, `Possible Metazoa`,
    `Superkingdom (NCBI)`, `Phylum (NCBI)`, `Class (NCBI)`, `Order (NCBI)`, 
    `Family (NCBI)`, `Genus (NCBI)`, `Species (NCBI)`,
    `ASV (Sequence)`, `ASV Size (pb)`,
    `Contamination status`,
    starts_with("Metadata")
      )

    # Save final results
    writexl::write_xlsx(
      results_table_clean,
      path = file.path(results_path, paste0("Final_ASV_Table_", Sys.Date(), ".xlsx"))
    )
    ```

### Save the Environment!

-   **Save the R environment.** Stores all objects and variables created in this section so you can reload them later if the pipeline fails or you need to resume from this point.

    ```{r}
    base::save.image(paste0(env_path, "/", "9_Final_results-", Sys.Date()))
    ```
